---
title: "Initial Model Evaluation"
subtitle: "Standard Goodness-of-Fit Metrics Without Bootstrapping"
author: "Siard van den Bosch"
date: 2025-03-06
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 10};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Compare goodness-of-fit metrics across unadjusted and adjusted fits. |
| **Inputs**     | `unadjusted_fits.rds`, `adjusted_data.rds`, `adjusted_fits.rds` |
| **Outputs**    | `fit_metrics.rds`, `nested_rlm_unadjusted.rds`, `nested_rlm_adjusted.rds`, `nested_ols_unadjusted.rds`, `nested_ols_adjusted.rds` |
| **Method**     | Goodness-of-fit summaries; nested Wald and deviance tests |
| **Requires**   | [Step 7 (Unadjusted Models)](/analysis/models/unadjusted/index.qmd), [Step 8 (Influence Diagnostics)](/analysis/models/assumptions/index.qmd), [Step 9 (Adjusted Models)](/analysis/models/adjusted/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- Wald and deviance tests: standard frequentist tests for incremental fit in nested specifications [@burnham2004]
- MM-estimation: high breakdown point combined with asymptotic efficiency [@yohai1987]
- Model 0 retained in RLM comparisons: provides a robust-scale baseline for nested tests.
- Model 0 excluded from OLS comparisons: variance comparisons against an intercept-only model are uninformative.
:::

::: {style="margin-block-end: 1rem;"}
:::

## Setup

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear environment
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions (consistent with steps 7-9)
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "format_p_value", "as_kable", "named_list", "structure_models",
  "add_style", "format_coef", "pval_str", "footnote", "sci_to_tex",
  "import_pipes", "dim_ci_rows", "set_globals", "row_grouping",
  "suppress_output", "beautify_names", "md_to_html", "numeric_df"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()

# global recompute switch
# TRUE re-runs the heavy steps; FALSE loads cached assets when they exist
recompute_heavy <- FALSE
```

```{r}
#| label: load-model-data
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# define input file paths
adjusted_data_path <- here::here("assets", "data", "saved", "adjusted_data.rds")
unadjusted_fits_path <- here::here("assets", "models", "unadjusted_fits.rds")
adjusted_fits_path <- here::here("assets", "models", "adjusted_fits.rds")

# verify input files exist
required_files <- c(adjusted_data_path, unadjusted_fits_path, adjusted_fits_path)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop("Missing required input files: ", paste(basename(missing_files), collapse = ", "))
}

# define output asset paths (gated by recompute_heavy below)
fit_metrics_path <- here::here("assets", "models", "fit_metrics.rds")
nested_rlm_unadj_path <- here::here("assets", "models", "nested_rlm_unadjusted.rds")
nested_rlm_adj_path <- here::here("assets", "models", "nested_rlm_adjusted.rds")
nested_ols_unadj_path <- here::here("assets", "models", "nested_ols_unadjusted.rds")
nested_ols_adj_path <- here::here("assets", "models", "nested_ols_adjusted.rds")

# load source data from step 8 (sensitivity dataset with influential obs removed)
source_data <- readRDS(adjusted_data_path)

# load fits and structure into nested list (unadjusted/adjusted × ols/robust)
unadjusted_fits <- readRDS(unadjusted_fits_path)
adjusted_fits <- readRDS(adjusted_fits_path)
model_fits <- structure_models(unadjusted_fits, adjusted_fits)

# extract model ids from stored fits
model_ids <- names(model_fits$adjusted$robust) %>%
  stringr::str_extract("[0-9]+$") %>%
  as.integer() %>%
  sort()
```

```{r}
#| label: verify-model-structure
#| results: asis
#| code-fold: true
#| column: body-outset-right

# verification block to confirm step 7/9 alignment
n_models <- length(model_ids)
model_range <- paste0("Models ", min(model_ids), "–", max(model_ids))
n_obs <- nrow(source_data)

cat(glue(
  "<div class='callout-note'>
<strong>Step 7/9 Model Verification</strong><br><br>
<strong>Models loaded:</strong> {n_models} models ({model_range})<br>
<strong>Unadjusted fits:</strong> From Step 7 (<code>unadjusted_fits.rds</code>)<br>
<strong>Adjusted fits:</strong> From Step 9 (<code>adjusted_fits.rds</code>)<br>
<strong>Analysis dataset:</strong> {n_obs} observations (sensitivity sample from Step 8)<br>
<strong>Model types:</strong> Composite (Models 0–6) and Disaggregated (Models 7–11)
</div>"
))
```

## Model-Hypothesis Mapping

Model numbers, descriptions, and hypothesis-test mapping; see Step 7 for formula details.

**Composite SES Models (Domain-Specific Parental Averages):**

| Model | Description                 | Tests                     | Hypotheses  |
| ----- | --------------------------- | ------------------------- | ----------- |
| 0     | Null (intercept only)       | Baseline                  | —           |
| 1     | Adult covariates + own SES  | Adjustment set            | —           |
| 2     | + Parental SES (composite)  | Parental SES main effects | H1, RQ1     |
| 3     | + Partner SES (composite)   | Partner SES main effects  | H2, RQ2     |
| 4     | + Gender × SES interactions | Gender moderation         | H3, H4, RQ3 |
| 5     | + Parental × Partner SES    | SES synergy               | H5, RQ4     |
| 6     | + Gender × synergy          | Gender-moderated synergy  | H6, RQ4     |

**Disaggregated SES Models (Separate Mother/Father/Partner Effects):**

| Model | Description                          | Tests                         | Hypotheses  |
| ----- | ------------------------------------ | ----------------------------- | ----------- |
| 7     | Covariates + Mother SES + Father SES | Differential parental effects | H1, RQ1     |
| 8     | + Partner SES (disaggregated)        | Partner SES main effects      | H2, RQ2     |
| 9     | + Gender × Mother/Father/Partner SES | Gender moderation by parent   | H3, H4, RQ3 |
| 10    | + Mother × Partner, Father × Partner | Parent-specific synergy       | H5, RQ4     |
| 11    | + Gender × synergy terms             | Gender-moderated synergy      | H6, RQ4     |

## Model Fit Statistics

```{r}
#| label: define-get-metrics
#| message: false
#| warning: false
#| code-fold: true
#| column: body-outset-right

# extract goodness-of-fit metrics from model object
get_metrics <- function(model_id, adjusted, robust, fits = model_fits) {
  lev_type <- ifelse(adjusted, "adjusted", "unadjusted")
  est_method <- ifelse(robust, "robust", "ols")
  model_key <- paste0("model_", model_id)

  model_fit <- fits[[lev_type]][[est_method]][[model_key]]$fit

  if (!inherits(model_fit, c("lmrob", "lm"))) {
    stop("Unsupported model class: ", class(model_fit)[1])
  }

  fit_sum <- summary(model_fit)
  n_obs <- length(fit_sum$residuals)
  n_params <- length(coef(model_fit)) + 1

  # compute information criteria (built-in for lm, approximated for lmrob)
  if (inherits(model_fit, "lm") && !inherits(model_fit, "lmrob")) {
    log_lik <- logLik(model_fit)
    aic <- AIC(model_fit)
    bic <- BIC(model_fit)
    rse_lbl <- "Residual_Std_Error"
  } else {
    # approximate log-likelihood for robust models assuming normal errors
    log_lik <- -n_obs / 2 * (log(2 * pi) + 2 * log(fit_sum$sigma) + 1)
    aic <- -2 * log_lik + 2 * n_params
    bic <- -2 * log_lik + log(n_obs) * n_params
    rse_lbl <- "Robust_RSE"
  }

  metrics <- data.frame(
    R_squared = fit_sum$r.squared,
    Adj_R_squared = fit_sum$adj.r.squared,
    RSE = fit_sum$sigma,
    AIC = aic,
    BIC = bic
  )
  colnames(metrics)[3] <- rse_lbl

  metrics
}
```

```{r}
#| label: compute-all-metrics
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# load cached metrics when available, otherwise compute the full grid
if (!recompute_heavy && file.exists(fit_metrics_path)) {
  metrics_df <- readRDS(fit_metrics_path)
} else {
  # create parameter grid for all model × leverage × method combinations
  params <- tidyr::expand_grid(
    model = model_ids,
    adjusted = c(TRUE, FALSE),
    robust = c(TRUE, FALSE)
  )

  # compute metrics for each combination
  metrics_df <- purrr::pmap_dfr(params, function(model, adjusted, robust) {
    met <- get_metrics(model, adjusted, robust)
    data.frame(
      model = model,
      leverage = ifelse(adjusted, "Adjusted", "Unadjusted"),
      method = ifelse(robust, "RLM", "OLS"),
      R_squared = met$R_squared,
      Adj_R_squared = met$Adj_R_squared,
      RSE = met[, 3],
      AIC = met$AIC,
      BIC = met$BIC
    )
  }) %>%
    dplyr::group_by(model, leverage) %>%
    dplyr::mutate(
      leverage = dplyr::if_else(dplyr::row_number() == 1,
        stringr::str_to_title(leverage), NA_character_
      )
    ) %>%
    dplyr::ungroup()

  # suppress R² for intercept-only model (model 0) as it is undefined
  r2_cols <- grepl("R_squared", names(metrics_df))
  metrics_df[metrics_df$model == 0, r2_cols] <- NA

  saveRDS(metrics_df, fit_metrics_path)
}

# create row labels for table grouping
model_rows <- paste("Model", metrics_df$model)
```

```{r}
#| label: model-fits-table
#| tbl-cap: "Summary of Model Fit Statistics for All Linear Regression Models (RLM and OLS)"
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# table column names and footnote
col_names_fits <- c(
  "Leverage", "Method", "$R^2$", "$R^2 (adj.)$",
  "Std. Error (RSE / Robust RSE)", "AIC", "BIC"
)

footnote_text_fits <- glue(
  "'Leverage' indicates whether influential observations were removed (Adjusted vs. Unadjusted).",
  "'Method' specifies estimation approach (OLS or RLM).",
  "$R^2$ denotes coefficient of determination;",
  "$R^2 (adj.)$ is its adjusted variant. 'Std. Error'",
  "reports residual standard error (OLS) or",
  "robust residual scale estimate (RLM).",
  "AIC and BIC are Akaike and Bayesian information criteria. ",
  "Higher $R^2$ indicates better fit;",
  "lower AIC/BIC suggests better model performance. ",
  "RLM AIC/BIC are normal-likelihood approximations and are not directly comparable to OLS values."
)

metrics_df %>%
  dplyr::select(-model) %>%
  as_kable(sci_threshold = FALSE, omit_leading_zero = TRUE, digits = 3) %>%
  kableExtra::pack_rows(index = row_grouping(model_rows), escape = FALSE) %>%
  footnote() %>%
  add_style()
```

## Nested Model Comparisons

```{r}
#| label: define-compare-models
#| message: false
#| warning: false
#| code-fold: true
#| column: body-outset-right

# compare two nested models using Wald and deviance tests
compare_models <- function(mod1, mod2) {
  extract_p_value <- function(x) {
    pr_cols <- grep("^Pr", names(x), value = TRUE)
    if (length(pr_cols) == 0) {
      return(NA_real_)
    }
    tail(x[[pr_cols[1]]], 1)
  }

  tryCatch(
    {
      if (inherits(mod1, "lmrob")) {
        wald <- anova(mod1, mod2, test = "Wald")
        dev <- anova(mod1, mod2, test = "Deviance")

        data.frame(
          df1 = mod1$degree.freedom,
          df2 = mod2$degree.freedom,
          wald_stat = tail(wald$Test.Stat, 1),
          wald_p = extract_p_value(wald),
          deviance_stat = tail(dev$Test.Stat, 1),
          deviance_p = extract_p_value(dev),
          df = tail(wald$Df, 1)
        )
      } else {
        wald <- lmtest::waldtest(mod1, mod2, test = "Chisq")
        aov <- anova(mod1, mod2)

        data.frame(
          df1 = mod1$df.residual,
          df2 = mod2$df.residual,
          wald_stat = wald$Chisq[2],
          wald_p = extract_p_value(wald),
          deviance_stat = aov$F[2],
          deviance_p = extract_p_value(aov),
          df = aov$Df[2]
        )
      }
    },
    error = function(e) {
      warning("Model comparison failed: ", e$message)
      data.frame(
        df1 = NA, df2 = NA, wald_stat = NA, wald_p = NA,
        deviance_stat = NA, deviance_p = NA, df = NA
      )
    }
  )
}
```

```{r}
#| label: define-compare-nested-models
#| message: false
#| warning: false
#| code-fold: true
#| column: body-outset-right

# generate nested model comparison table
compare_nested_models <- function(fits, leverage, method,
                                  col_names = NULL, footnote_text = NULL) {
  models <- purrr::map(fits[[leverage]][[method]], ~ .x$fit)

  mn <- names(models)
  model_nums <- as.integer(sub(".*_", "", mn))

  ord <- order(model_nums)
  models <- models[ord]
  mn <- mn[ord]
  model_nums <- model_nums[ord]

  is_nested <- function(mod1, mod2) {
    terms1 <- attr(terms(mod1), "term.labels")
    terms2 <- attr(terms(mod2), "term.labels")
    all(terms1 %in% terms2) && !setequal(terms1, terms2)
  }

  num_combs <- combn(model_nums, 2, simplify = FALSE)
  name_combs <- combn(mn, 2, simplify = FALSE)

  nested_idx <- purrr::map_lgl(name_combs, function(nv) {
    is_nested(models[[nv[1]]], models[[nv[2]]])
  })

  num_combs <- num_combs[nested_idx]
  name_combs <- name_combs[nested_idx]

  model_pairs <- Map(
    function(nums, nv) setNames(nums, nv),
    num_combs, name_combs
  )

  comps <- model_pairs %>%
    purrr::map_df(~ {
      mod1 <- models[[names(.x)[1]]]
      mod2 <- models[[names(.x)[2]]]
      data.frame(
        model_a = .x[1],
        model_b = .x[2],
        model_pair = sprintf("%d vs. %d", .x[1], .x[2]),
        suppress_output(compare_models(mod1, mod2))
      )
    })

  labs <- glue("Model {comps$model_a}")

  comps %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      wald_stat = format_coef(wald_stat, wald_p),
      deviance_stat = format_coef(deviance_stat, deviance_p)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(-model_a, -model_b, -wald_p, -deviance_p) %>%
    as_kable() %>%
    kableExtra::pack_rows(index = row_grouping(labs), escape = FALSE) %>%
    footnote() %>%
    add_style()
}
```

### Robust Linear Models (RLM)

```{r}
#| label: nested-rlm-table-attributes
#| message: false
#| warning: false
#| code-fold: true
#| column: body-outset-right

# column names and footnote for RLM nested comparisons
col_names <- c(
  "$\\text{}$",
  "$DF_{\\text{pseudo1}}$",
  "$DF_{\\text{pseudo2}}$",
  "Wald $\\chi^2$",
  "Deviance",
  "DF"
)

footnote_text <- glue(
  "{pval_str(use_dot_for_0.1 = TRUE, html = FALSE)}. ",
  "Pseudo-degrees of freedom ($DF_{{\\text{{pseudo}}}}$)",
  "represent adjusted degrees of freedom in robust models. ",
  "$DF_{{\\text{{pseudo1}}}}$ corresponds to the simpler model;",
  "$DF_{{\\text{{pseudo2}}}}$ to the more complex model. ",
  "Wald and Deviance statistics assess improvements in nested models."
)
```

```{r}
#| label: nested-models-rlm-unadjusted
#| tbl-cap: "Wald and Deviance Tests for Nested Robust Linear Regression Models (Models 0–11, with influential observations)"
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

if (!recompute_heavy && file.exists(nested_rlm_unadj_path)) {
  nested_rlm_unadjusted <- readRDS(nested_rlm_unadj_path)
} else {
  nested_rlm_unadjusted <- compare_nested_models(model_fits, "unadjusted", "robust")
  saveRDS(nested_rlm_unadjusted, nested_rlm_unadj_path)
}

nested_rlm_unadjusted
```

```{r}
#| label: nested-models-rlm-adjusted
#| tbl-cap: "Wald and Deviance Tests for Nested Robust Linear Regression Models (Models 0–11, adjusted for influential observations)"
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

if (!recompute_heavy && file.exists(nested_rlm_adj_path)) {
  nested_rlm_adjusted <- readRDS(nested_rlm_adj_path)
} else {
  nested_rlm_adjusted <- compare_nested_models(model_fits, "adjusted", "robust")
  saveRDS(nested_rlm_adjusted, nested_rlm_adj_path)
}

nested_rlm_adjusted
```

### OLS Models

```{r}
#| label: nested-ols-table-attributes
#| message: false
#| warning: false
#| code-fold: true
#| column: body-outset-right

# column names and footnote for OLS nested comparisons
col_names <- c(
  "$\\text{}$",
  "$DF_{\\text{resid1}}$",
  "$DF_{\\text{resid2}}$",
  "Wald $\\chi^2$",
  "F",
  "DF"
)

footnote_text <- glue(
  "{pval_str(use_dot_for_0.1 = TRUE, html = FALSE)}. ",
  "$DF_{{\\text{{resid}}}}$ are residual degrees of freedom. ",
  "Wald $\\chi^2$ from `lmtest::waldtest()`. F statistic from `anova()`. ",
  "DF reports difference in degrees of freedom between models."
)
```

```{r}
#| label: nested-models-ols-unadjusted
#| tbl-cap: "Wald and Explained Variance Tests for Nested OLS Linear Regression Models (Models 1–11, with influential observations)"
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# exclude model 0 (intercept-only) from OLS comparisons
if (!recompute_heavy && file.exists(nested_ols_unadj_path)) {
  nested_ols_unadjusted <- readRDS(nested_ols_unadj_path)
} else {
  nested_ols_unadjusted <- purrr::map_depth(model_fits, 2, ~ .x[-1]) %>%
    compare_nested_models("unadjusted", "ols")
  saveRDS(nested_ols_unadjusted, nested_ols_unadj_path)
}

nested_ols_unadjusted
```

```{r}
#| label: nested-models-ols-adjusted
#| tbl-cap: "Wald and Explained Variance Tests for Nested OLS Linear Regression Models (Models 1–11, adjusted for influential observations)"
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# exclude model 0 (intercept-only) from OLS comparisons
if (!recompute_heavy && file.exists(nested_ols_adj_path)) {
  nested_ols_adjusted <- readRDS(nested_ols_adj_path)
} else {
  nested_ols_adjusted <- purrr::map_depth(model_fits, 2, ~ .x[-1]) %>%
    compare_nested_models("adjusted", "ols")
  saveRDS(nested_ols_adjusted, nested_ols_adj_path)
}

nested_ols_adjusted
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File   |                    | Content                                                                                                  |
| --------------------------- | -------------------------------------------------------------------------------------------------------- |
| `fit_metrics.rds`           | Goodness-of-fit grid (R², adjusted R², RSE, AIC, BIC) across all model × leverage × method combinations. |
| `nested_rlm_unadjusted.rds` | Rendered Wald/deviance nested-comparison table, robust models, with influential observations.            |
| `nested_rlm_adjusted.rds`   | Rendered Wald/deviance nested-comparison table, robust models, influential observations removed.         |
| `nested_ols_unadjusted.rds` | Rendered Wald/F nested-comparison table, OLS models, with influential observations.                      |
| `nested_ols_adjusted.rds`   | Rendered Wald/F nested-comparison table, OLS models, influential observations removed.                   |

**Next:** [Step 11 (Interaction Effects)](/analysis/models/goodness-of-fit/interaction-effects/index.qmd)
