---
title: "Influence Diagnostics & Sensitivity Sample"
subtitle: "Diagnosing Unadjusted Fits and Constructing the Trimmed Dataset"
author: "Siard van den Bosch"
date: 2025-01-22
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 8};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Diagnose unadjusted fits and build a sensitivity-filtered dataset |
| **Inputs**     | `unadjusted_data.rds`, `unadjusted_fits.rds` |
| **Outputs**    | `adjusted_data.rds`, `unadjusted_influence_results.rds` |
| **Method**     | MM-estimation residual diagnostics; OLS companion tests; multi-criterion influence flagging |
| **Requires**   | [Step 7 (Unadjusted Models)](/analysis/models/unadjusted/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- MM-estimation as primary estimator: bounded influence and high efficiency without strict distributional assumptions [@yohai1987]
- Descriptive residual summaries over normality tests: large-sample normality tests reject trivial departures [@maronna2006]
- OLS companions for classical tests: enables transparent reporting of Breusch-Pagan, RESET, and Durbin-Watson on identical row sets [@koenker1981note]
- Combined-criteria influence flagging over automatic deletion: MM-estimators already downweight outliers via bounded influence [@hampel1986robust]
- Cross-model aggregation of flags: targets cases repeatedly problematic across nested specifications [@cook1982residuals]
- Sensitivity dataset as robustness check: full and filtered fits are compared, not corrected away [@rousseeuw1987]
:::

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear global environment, garbage collect, and detach packages
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions (consistent with step 7)
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "as_kable", "set_globals", "pval_str", "add_style", "sci_to_tex",
  "md_to_html", "describe", "format_p_value", "std_res", "footnote",
  "beautify_names", "min_bootstraps", "numeric_df", "import_pipes"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()
```

```{r}
#| label: load-step7-inputs
#| code-fold: show
#| column: body-outset-right

# define input paths from step 7
unadjusted_data_path <- here::here("assets", "data", "saved", "unadjusted_data.rds")
unadjusted_fits_path <- here::here("assets", "models", "unadjusted_fits.rds")

# validate input files exist
if (!file.exists(unadjusted_data_path)) {
  stop("Missing input from Step 7: ", unadjusted_data_path)
}
if (!file.exists(unadjusted_fits_path)) {
  stop("Missing input from Step 7: ", unadjusted_fits_path)
}

# load unadjusted dataset
unadjusted_data <- readRDS(unadjusted_data_path)
if (!is.data.frame(unadjusted_data)) {
  stop("'unadjusted_data' must be a data.frame.")
}

# retrieve robust regression models from step 7
# step 7's robust_analysis() returns structured_models objects:
# - $robust_model_results$model_object for lmrob fits
# - $ols_model_results$model_object for OLS fits
fits_raw <- readRDS(unadjusted_fits_path)
if (!is.list(fits_raw) || length(fits_raw) < 2) {
  stop("Expected a list of model fits in ", unadjusted_fits_path)
}

# extract lmrob fits, dropping null model (fit_0)
lmrob_fits <- fits_raw %>%
  purrr::map(~ .x$robust_model_results$model_object) %>%
  tail(-1)

if (length(lmrob_fits) == 0) {
  stop("No models found after dropping the null model.")
}
if (is.null(names(lmrob_fits)) || any(names(lmrob_fits) == "")) {
  names(lmrob_fits) <- paste0("fit_", seq_along(lmrob_fits))
}

# extract OLS companion fits (pre-computed in step 7)
ols_fits <- fits_raw %>%
  purrr::map(~ .x$ols_model_results$model_object) %>%
  tail(-1)
names(ols_fits) <- names(lmrob_fits)

# record model count for downstream use
n_models <- length(lmrob_fits)
cat(glue("Loaded {n_models} substantive models (Models 1-{n_models}) from Step 7.\n\n"))
```

```{r}
#| label: verify-step7-alignment
#| results: asis
#| code-fold: true
#| column: body-outset-right

# display verification summary
verify_msg <- glue(
  "<div class='callout-note'>
<strong>Step 7 Alignment Verification</strong><br><br>
<strong>Data source:</strong> <code>unadjusted_data.rds</code> ({nrow(unadjusted_data)} observations, {ncol(unadjusted_data)} variables)<br>
<strong>Model fits:</strong> <code>unadjusted_fits.rds</code> ({n_models} substantive models)<br>
<strong>Model names:</strong> {paste(names(lmrob_fits), collapse = ', ')}<br>
<strong>OLS companions:</strong> Pre-computed in Step 7 (not rebuilt)
</div>"
)
cat(verify_msg)
```

## Diagnostic Tests and Assumption Checks

### Assessing Residual Shape

```{r}
#| label: define-skew-kurt-function
#| code-fold: true
#| column: body-outset-right

# compute sample skewness and kurtosis without extra packages
skew_kurt <- function(x) {
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 4) {
    return(c(skewness = NA_real_, kurtosis = NA_real_))
  }

  m <- mean(x)
  s <- stats::sd(x)
  if (!is.finite(s) || s <= .Machine$double.eps) {
    return(c(skewness = NA_real_, kurtosis = NA_real_))
  }

  z <- (x - m) / s
  c(
    skewness = mean(z^3),
    kurtosis = mean(z^4)
  )
}
```

```{r}
#| label: define-residual-shape-function
#| code-fold: true
#| column: body-outset-right

# summarize standardized robust residuals for a single model
residual_shape <- function(fit) {
  r <- stats::residuals(fit) / fit$scale
  r <- r[is.finite(r)]

  if (length(r) == 0) {
    return(data.frame(
      n = 0L,
      mean = NA_real_,
      sd = NA_real_,
      median = NA_real_,
      MAD = NA_real_,
      skewness = NA_real_,
      kurtosis = NA_real_,
      p_gt2 = NA_real_,
      p_gt3 = NA_real_
    ))
  }

  sk <- skew_kurt(r)

  data.frame(
    n = length(r),
    mean = mean(r),
    sd = stats::sd(r),
    median = stats::median(r),
    MAD = stats::mad(r),
    skewness = unname(sk["skewness"]),
    kurtosis = unname(sk["kurtosis"]),
    p_gt2 = mean(abs(r) > 2),
    p_gt3 = mean(abs(r) > 3)
  )
}
```

```{r}
#| label: residual-shape-table-setup
#| code-fold: true
#| column: body-outset-right

# table metadata for residual shape summary
col_names <- c(
  "Model", "$n$", "mean", "sd", "median", "MAD",
  "skewness", "kurtosis", "$P(|r|>2)$", "$P(|r|>3)$"
)

footnote_text <- glue(
  "Residual summaries are computed on standardized robust residuals $r = e/\\text{{scale}}$. ",
  "Skewness and kurtosis are sample moment summaries (kurtosis is not excess kurtosis). ",
  "$P(|r|>2)$ and $P(|r|>3)$ quantify tail heaviness relative to a standard normal reference."
)
```

```{r}
#| label: residual-shape-table
#| tbl-cap: "Descriptive Summary of Standardized Robust Residuals"
#| message: false
#| code-fold: show
#| column: body-outset-right

# compute residual shape summary across all models
purrr::imap_dfr(lmrob_fits, ~ {
  out <- residual_shape(.x)
  out$Model <- gsub("fit_", "", .y)
  out
}) %>%
  dplyr::select(Model, dplyr::everything()) %>%
  as_kable(digits = 3) %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: define-combined-qq-function
#| code-fold: true
#| column: body-outset-right

# create combined Q-Q plots using standardized robust residuals
combined_qq_plot <- function(models, labels, jitter = TRUE) {
  shapes <- (seq_along(models) - 1) %% 25

  plot_data <- purrr::map2_df(
    models, labels,
    ~ data.frame(std_resid = std_res(.x), Model = .y)
  )

  # preserve the supplied label order in the legend (default factor levels
  # sort alphabetically: 1, 10, 11, 2, ...)
  plot_data$Model <- factor(plot_data$Model, levels = unique(unlist(labels)))

  ggplot2::ggplot(plot_data, ggplot2::aes(
    sample = std_resid,
    colour = Model,
    shape = Model
  )) +
    ggplot2::stat_qq(
      size = 2,
      alpha = 0.8,
      position = if (jitter) ggplot2::position_jitter(width = 0.1) else ggplot2::position_identity()
    ) +
    ggplot2::geom_abline(
      intercept = 0,
      slope = 1,
      linetype = "dashed",
      linewidth = 0.5,
      color = "black"
    ) +
    viridis::scale_color_viridis(
      discrete = TRUE,
      option = "D",
      begin = 0.1,
      end = 0.9,
      name = "Model"
    ) +
    ggplot2::scale_shape_manual(values = shapes, name = "Model") +
    ggplot2::labs(
      x = "Theoretical Quantiles",
      y = "Standardized Robust Residuals"
    ) +
    ggplot2::guides(
      colour = ggplot2::guide_legend(nrow = 1, byrow = TRUE),
      shape = ggplot2::guide_legend(nrow = 1, byrow = TRUE)
    )
}
```

```{r}
#| label: normal-qq-plots
#| fig-cap: "Q-Q Plots of Standardized Robust Residuals"
#| code-fold: true
#| column: body-outset-right

combined_qq_plot(lmrob_fits, as.list(as.character(seq_along(lmrob_fits))))
```

### Evaluating Heteroskedasticity

```{r}
#| label: bp-test-table-setup
#| code-fold: true
#| column: body-outset-right

# table metadata for Breusch-Pagan tests
col_names <- c("Model", "Statistic", "df", "$p$-value")

footnote_text <- glue(
  "Studentized Breusch-Pagan (Koenker) tests are computed on OLS companion models ",
  "fit to the same rows as each robust model. The null is constant variance. ",
  "These tests are used as diagnostics rather than gatekeepers for model validity."
)
```

```{r}
#| label: bp-test-for-heteroscedasticity
#| tbl-cap: "Studentized Breusch-Pagan Tests on OLS Companion Models"
#| message: false
#| code-fold: show
#| column: body-outset-right

# compute Breusch-Pagan tests across all OLS companion models
bp_results <- purrr::imap_dfr(ols_fits, ~ {
  test_res <- lmtest::bptest(.x, studentize = TRUE)

  data.frame(
    Model = gsub("fit_", "", .y),
    Statistic = unname(test_res$statistic),
    df = unname(test_res$parameter),
    p_value = format_p_value(
      test_res$p.value,
      add_stars = TRUE,
      use_concat = TRUE,
      omit_p_label = TRUE,
      use_dot_for_0.1 = TRUE
    )
  )
})

bp_results %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: define-scale-location-function
#| code-fold: true
#| column: body-outset-right

# create scale-location plot for a single robust model
scale_location_plot <- function(fit, name = NULL) {
  point_color <- viridis::viridis(4)[2]
  line_color <- viridis::viridis(4)[4]

  if (!"scale" %in% names(fit)) {
    stop("Model does not contain 'scale' component.")
  }

  std_resid <- stats::residuals(fit) / fit$scale

  model_data <- tibble::tibble(
    .fitted = stats::fitted(fit),
    sqrt_abs_std_resid = sqrt(abs(std_resid))
  )

  ggplot2::ggplot(model_data, ggplot2::aes(.fitted, sqrt_abs_std_resid)) +
    ggplot2::geom_point(alpha = 0.6, color = point_color) +
    ggplot2::geom_smooth(
      method = "loess",
      se = TRUE,
      formula = y ~ x,
      color = line_color,
      method.args = list(
        span = 1,
        degree = 1,
        family = "symmetric",
        control = stats::loess.control(surface = "direct")
      )
    ) +
    ggplot2::labs(
      title = name,
      x = "Fitted values",
      y = expression(sqrt("| Std. Residuals |"))
    )
}
```

```{r}
#| label: scale-location-plot-generation
#| fig-cap: "Scale-Location Plots (Robust Models)"
#| fig-subcap: true
#| layout-ncol: 2
#| message: false
#| code-fold: true
#| column: body-outset-right

purrr::iwalk(lmrob_fits, function(fit, nm) {
  plot(scale_location_plot(fit, name = paste0("Model ", gsub("fit_", "", nm))))
})
```

### Testing Mean-Structure Adequacy

```{r}
#| label: define-response-predictor-function
#| code-fold: true
#| column: body-outset-right

# marginal model plot: outcome vs linear predictor (fitted values);
# under a correct mean structure the data loess tracks the y = x line
response_predictor_plot <- function(fit, name = NULL) {
  point_color <- viridis::viridis(4)[2]
  line_color <- viridis::viridis(4)[4]

  mf <- stats::model.frame(fit)
  if (ncol(mf) < 2) return(NULL)

  d <- tibble::tibble(
    fitted = stats::fitted(fit),
    y = mf[[1]]
  )

  ggplot2::ggplot(d, ggplot2::aes(fitted, y)) +
    ggplot2::geom_point(alpha = 0.4, color = point_color) +
    ggplot2::geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
    ggplot2::geom_smooth(
      method = "loess", se = TRUE, formula = y ~ x, color = line_color,
      method.args = list(span = 1, degree = 1, family = "symmetric",
        control = stats::loess.control(surface = "direct"))
    ) +
    ggplot2::labs(
      title = name,
      x = "Linear predictor (fitted)",
      y = names(mf)[1]
    )
}
```

```{r}
#| label: response-predictor-plots
#| fig-cap: "Outcome vs Linear Predictor (Marginal Model Plots)"
#| fig-subcap: true
#| layout-ncol: 2
#| message: false
#| code-fold: true
#| column: body-outset-right

purrr::iwalk(lmrob_fits, function(fit, nm) {
  p <- response_predictor_plot(fit, name = paste0("Model ", gsub("fit_", "", nm)))
  if (!is.null(p)) plot(p)
})
```

```{r}
#| label: define-residuals-fitted-function
#| code-fold: true
#| column: body-outset-right

# create residuals vs fitted plot for a single robust model
residuals_fitted_plot <- function(fit, name = NULL) {
  point_color <- viridis::viridis(4)[2]
  line_color <- viridis::viridis(4)[4]

  std_resid <- stats::residuals(fit) / fit$scale

  model_data <- tibble::tibble(
    .fitted = stats::fitted(fit),
    std_resid = std_resid
  )

  ggplot2::ggplot(model_data, ggplot2::aes(.fitted, std_resid)) +
    ggplot2::geom_point(alpha = 0.6, color = point_color) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    ggplot2::geom_smooth(
      method = "loess",
      se = TRUE,
      formula = y ~ x,
      color = line_color,
      method.args = list(
        span = 1,
        degree = 1,
        family = "symmetric",
        control = stats::loess.control(surface = "direct")
      )
    ) +
    ggplot2::labs(
      title = name,
      x = "Fitted values",
      y = "Standardized Residuals"
    )
}
```

```{r}
#| label: residuals-fitted-plots
#| fig-cap: "Residuals vs Fitted (Robust Models)"
#| fig-subcap: true
#| layout-ncol: 2
#| message: false
#| code-fold: true
#| column: body-outset-right

purrr::iwalk(lmrob_fits, function(fit, nm) {
  plot(residuals_fitted_plot(fit, name = paste0("Model ", gsub("fit_", "", nm))))
})
```

```{r}
#| label: reset-test-table-setup
#| code-fold: true
#| column: body-outset-right

# table metadata for RESET tests
col_names <- c("Model", "Statistic", "df1", "df2", "$p$-value")

footnote_text <- glue(
  "RESET tests are computed on OLS companion models. ",
  "The null hypothesis is correct mean specification. ",
  "These tests are used as diagnostics rather than gatekeepers."
)
```

```{r}
#| label: reset-test-for-specification
#| tbl-cap: "Ramsey RESET Tests on OLS Companion Models"
#| message: false
#| code-fold: show
#| column: body-outset-right

# compute RESET tests across all OLS companion models
reset_results <- purrr::imap_dfr(ols_fits, ~ {
  test_res <- lmtest::resettest(.x, power = 2:3, type = "fitted")

  data.frame(
    Model = gsub("fit_", "", .y),
    Statistic = unname(test_res$statistic),
    df1 = unname(test_res$parameter[1]),
    df2 = unname(test_res$parameter[2]),
    p_value = format_p_value(
      test_res$p.value,
      add_stars = TRUE,
      use_concat = TRUE,
      omit_p_label = TRUE,
      use_dot_for_0.1 = TRUE
    )
  )
})

reset_results %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Checking Autocorrelation

```{r}
#| label: dw-test-table-setup
#| code-fold: true
#| column: body-outset-right

# table metadata for Durbin-Watson tests
col_names <- c("Model", "Statistic", "$p$-value")

footnote_text <- glue(
  "Durbin-Watson tests are computed on OLS companion models. ",
  "The null hypothesis is no first-order autocorrelation."
)
```

```{r}
#| label: dw-test-for-autocorrelation
#| tbl-cap: "Durbin-Watson Tests on OLS Companion Models"
#| message: false
#| code-fold: show
#| column: body-outset-right

# compute Durbin-Watson tests across all OLS companion models
dw_results <- purrr::imap_dfr(ols_fits, ~ {
  test_res <- lmtest::dwtest(.x)

  data.frame(
    Model = gsub("fit_", "", .y),
    Statistic = unname(test_res$statistic),
    p_value = format_p_value(
      test_res$p.value,
      add_stars = TRUE,
      use_concat = TRUE,
      omit_p_label = TRUE,
      use_dot_for_0.1 = TRUE
    )
  )
})

dw_results %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Assessing Multicollinearity

```{r}
#| label: define-vif-functions
#| code-fold: true
#| column: body-outset-right

# compute VIF from model matrix
calc_vif <- function(mm) {
  int <- match("(Intercept)", colnames(mm), nomatch = 0L)
  if (int > 0L) {
    mm <- mm[, -int, drop = FALSE]
  }

  if (ncol(mm) < 2L) {
    return(setNames(numeric(0), character(0)))
  }

  ok <- rowSums(!is.finite(mm)) == 0L
  mm <- mm[ok, , drop = FALSE]

  if (nrow(mm) < 3L) {
    out <- rep(NA_real_, ncol(mm))
    names(out) <- colnames(mm)
    return(out)
  }

  vifs <- vapply(seq_len(ncol(mm)), function(j) {
    y <- mm[, j]
    x <- mm[, -j, drop = FALSE]

    sst <- sum((y - mean(y))^2)
    if (!is.finite(sst) || sst <= .Machine$double.eps) {
      return(NA_real_)
    }

    aux <- stats::lm.fit(x = cbind(1, x), y = y)
    rss <- sum(aux$residuals^2)
    r2 <- 1 - rss / sst
    r2 <- max(min(r2, 1), 0)

    if (r2 >= 1 - sqrt(.Machine$double.eps)) {
      return(Inf)
    }

    1 / (1 - r2)
  }, numeric(1))

  names(vifs) <- colnames(mm)
  vifs
}

# extract maximum VIF from a fitted model
get_max_vif <- function(fit) {
  mm <- tryCatch(stats::model.matrix(fit), error = function(e) NULL)
  if (is.null(mm)) {
    return(NA_real_)
  }

  vifs <- calc_vif(mm)
  if (length(vifs) == 0 || all(is.na(vifs))) {
    return(NA_real_)
  }

  max(vifs, na.rm = TRUE)
}
```

```{r}
#| label: compute-max-vifs
#| code-fold: true
#| column: body-outset-right

# compute max VIF for each model (with progress bar)
max_vifs <- pbapply::pbsapply(lmrob_fits, get_max_vif)
```

```{r}
#| label: vif-table-setup
#| code-fold: true
#| column: body-outset-right

# table metadata for VIF summary
col_names <- c("Model", "Max VIF$^a$")

footnote_text <- glue(
  "The VIF quantifies how strongly each regressor is linearly explained by the others. ",
  "$^a$The maximum VIF across regressors is reported for each model."
)
```

```{r}
#| label: vif-table
#| tbl-cap: "Variance Inflation Factors (VIFs)"
#| code-fold: show
#| column: body-outset-right

tibble::tibble(
  Model = seq_along(lmrob_fits),
  MaxVIF = as.numeric(max_vifs)
) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Identifying Influential Observations

```{r}
#| label: define-influence-helper-functions
#| code-fold: true
#| column: body-outset-right

# extract participant IDs from fitted model
get_pid <- function(fit) {
  mf <- tryCatch(stats::model.frame(fit), error = function(e) NULL)
  pid <- if (!is.null(mf)) rownames(mf) else NULL

  if (is.null(pid) || anyNA(pid) || any(pid == "")) {
    pid <- as.character(seq_along(stats::residuals(fit)))
  }

  pid
}

# create long-format influence diagnostics for a single model
influence_long_df <- function(fit, model_name) {
  pid <- get_pid(fit)
  n <- length(pid)

  r <- stats::residuals(fit) / fit$scale
  h <- tryCatch(stats::hatvalues(fit), error = function(e) rep(NA_real_, n))
  w <- fit$rweights
  if (is.null(w)) {
    w <- rep(NA_real_, n)
  }

  p <- length(stats::coef(fit))
  lev_thr <- 2 * p / n

  score <- abs(r) * (h / pmax(1 - h, .Machine$double.eps))

  tibble::tibble(
    pid = as.character(pid),
    Model = model_name,
    std_resid = as.numeric(r),
    leverage = as.numeric(h),
    rweight = as.numeric(w),
    lev_thr = lev_thr,
    score = as.numeric(score)
  )
}

# map flagged pids back to data rows
match_ids_to_rows <- function(data, pids) {
  pids <- unique(as.character(pids))
  if (length(pids) == 0) {
    return(integer(0))
  }

  candidate_cols <- c("pid", "id", "obs_id", "row_id", "case_id", "participant_id", "idnummer")
  data_cols_lower <- tolower(names(data))

  for (nm in candidate_cols) {
    j <- match(nm, data_cols_lower, nomatch = 0L)
    if (j > 0L) {
      col <- names(data)[j]
      hits <- which(as.character(data[[col]]) %in% pids)
      if (length(hits) > 0) {
        return(hits)
      }
    }
  }

  if (!is.null(rownames(data))) {
    hits <- which(rownames(data) %in% pids)
    if (length(hits) > 0) {
      return(hits)
    }
  }

  idx <- suppressWarnings(as.integer(pids))
  if (!anyNA(idx) && all(idx >= 1) && all(idx <= nrow(data))) {
    return(idx)
  }

  integer(0)
}
```

- **in**: `lmrob_fits` (per-model robust diagnostics)
- **out**: `influence_results$influential_obs`, `influence_results$influential_ids`
- **rule**: flag observation in a model if robust weight is below the global 5th percentile and $|r|>2.5$, or if leverage exceeds $2p/n$ and $|r|>2.5$; mark as influential if flagged in at least 25% of models or if its maximum score reaches the 99th percentile

```{r}
#| label: compute-influence-summary
#| code-fold: show
#| column: body-outset-right

# define flagging thresholds
alpha_weight <- 0.05
resid_cut <- 2.5

# compute influence diagnostics across all models
infl_long <- purrr::imap_dfr(lmrob_fits, ~ {
  influence_long_df(.x, model_name = gsub("fit_", "", .y))
})

# global robust-weight threshold (bottom 5% across all models)
w_thr <- stats::quantile(infl_long$rweight, probs = alpha_weight, na.rm = TRUE)

# apply flagging criteria
infl_long <- infl_long %>%
  dplyr::mutate(
    low_weight = is.finite(rweight) & rweight < w_thr,
    high_leverage = is.finite(leverage) & leverage > lev_thr, # per-model 2p/n column
    large_resid = is.finite(std_resid) & abs(std_resid) > resid_cut,
    flag_model = (low_weight & large_resid) | (high_leverage & large_resid)
  )

k <- length(lmrob_fits)

# aggregate influence flags across models
infl_summary <- infl_long %>%
  dplyr::group_by(pid) %>%
  dplyr::summarise(
    n_flag = sum(flag_model, na.rm = TRUE),
    max_abs_resid = max(abs(std_resid), na.rm = TRUE),
    max_leverage = max(leverage, na.rm = TRUE),
    min_rweight = min(rweight, na.rm = TRUE),
    max_score = max(score, na.rm = TRUE),
    .groups = "drop"
  )

# identify influential observations using combined criteria
score_thr <- stats::quantile(infl_summary$max_score, probs = 0.99, na.rm = TRUE)
flag_min_models <- ceiling(k * 0.25)

influential_obs <- infl_summary %>%
  dplyr::mutate(
    influential = (n_flag >= flag_min_models) | (max_score >= score_thr)
  ) %>%
  dplyr::arrange(dplyr::desc(influential), dplyr::desc(n_flag), dplyr::desc(max_score))

influential_ids <- influential_obs %>%
  dplyr::filter(influential) %>%
  dplyr::pull(pid) %>%
  unique()

# store results
influence_results <- list(
  influential_obs = influential_obs,
  influential_ids = influential_ids,
  thresholds = list(
    w_thr = w_thr,
    resid_cut = resid_cut,
    lev_rule = "2p/n",
    score_thr = score_thr,
    min_models = flag_min_models
  )
)
```

```{r}
#| label: display-influence-message
#| results: asis
#| code-fold: true
#| column: body-outset-right

# compute summary statistics
num_influential <- length(influence_results$influential_ids)
data_size <- nrow(unadjusted_data)
pct_influential <- if (data_size > 0) round(100 * num_influential / data_size, 2) else NA_real_

htmltools::HTML(glue(
  "<p>Out of {data_size} observations, {num_influential} ({pct_influential}%) were flagged as potentially influential.<br>",
  "Global robust-weight threshold: {round(influence_results$thresholds$w_thr, 3)} (alpha = {alpha_weight}). ",
  "Residual cut: |r| > {resid_cut}. ",
  "Leverage rule: {influence_results$thresholds$lev_rule}. ",
  "Across-model flag: at least {influence_results$thresholds$min_models} of {k} models or max score above the 99th percentile.",
  "</p>"
))
```

```{r}
#| label: influence-table-setup
#| code-fold: true
#| column: body-outset-right

# table metadata for influential observations
col_names <- c(
  "pid", "Influential", "n_flag", "max |r|", "max leverage", "min rweight", "max score"
)

footnote_text <- glue(
  "Rows are aggregated across models. 'n_flag' is the number of models where the observation ",
  "meets the within-model rule (low weight and large residual, or high leverage and large residual). ",
  "'max score' is max(|r| × h/(1-h)) across models and is used as an additional screen."
)
```

```{r}
#| label: influence-table
#| tbl-cap: "Potentially Influential Observations (Across Models)"
#| message: false
#| code-fold: show
#| column: page-right

influence_results$influential_obs %>%
  dplyr::mutate(
    Influential = ifelse(influential, 1, 0)
  ) %>%
  dplyr::select(pid, Influential, n_flag, max_abs_resid, max_leverage, min_rweight, max_score) %>%
  as_kable(digits = 3) %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: define-influence-plot-function
#| code-fold: true
#| column: body-outset-right

# create influence diagnostic panel for a single model
plot_influence_panel <- function(df, flagged_ids, name = NULL) {
  df <- df %>%
    dplyr::mutate(
      flagged = pid %in% flagged_ids,
      label = ifelse(flagged, pid, NA_character_),
      influence = 2 / pmax(1 - leverage, .Machine$double.eps)
    )

  breaks <- unique(stats::quantile(df$influence, c(0, 0.33, 0.67, 1), na.rm = TRUE))
  breaks <- breaks[is.finite(breaks)]
  n_intervals <- length(breaks) - 1

  if (n_intervals <= 0) {
    df$influence_cat <- factor("Low", levels = "Low")
  } else {
    labels <- c("Low", "Medium", "High")[seq_len(n_intervals)]
    df$influence_cat <- cut(
      df$influence,
      breaks = breaks,
      labels = labels,
      include.lowest = TRUE
    )
  }

  p <- ggplot2::ggplot(df, ggplot2::aes(leverage, std_resid, color = influence_cat)) +
    ggplot2::geom_point(alpha = 0.6) +
    ggplot2::geom_smooth(
      method = "loess",
      se = TRUE,
      alpha = 0.3,
      formula = y ~ x,
      method.args = list(
        span = 1,
        degree = 1,
        family = "symmetric",
        control = stats::loess.control(surface = "direct")
      )
    ) +
    ggplot2::labs(
      title = name,
      x = "Leverage",
      y = "Standardized Residuals",
      color = "Influence"
    ) +
    ggplot2::scale_color_viridis_d(option = "D", direction = -1) +
    ggplot2::guides(color = ggplot2::guide_legend(title = "Influence"))

  if (requireNamespace("ggrepel", quietly = TRUE) && any(df$flagged, na.rm = TRUE)) {
    p <- p + ggrepel::geom_text_repel(
      ggplot2::aes(label = label),
      size = 2,
      max.overlaps = 20,
      na.rm = TRUE
    )
  }

  p
}
```

```{r}
#| label: generate-influence-plots
#| fig-cap: "Leverage vs Standardized Robust Residuals (Influence Categories)"
#| message: false
#| code-fold: show
#| layout-ncol: 2
#| layout-type: grid
#| column: body-outset-right

flagged_ids <- influence_results$influential_ids

# split() orders panels alphabetically (1, 10, 11, 2, ...); restore numeric order
infl_split <- split(infl_long, infl_long$Model)
infl_split <- infl_split[order(as.integer(names(infl_split)))]
invisible(purrr::iwalk(infl_split, ~ print(
  plot_influence_panel(.x, flagged_ids, name = paste0("Model ", .y))
)))
```

## Create Sensitivity Dataset

- *in*: `unadjusted_data`, `influence_results$influential_ids`
- *out*: `adjusted_data.rds`, `unadjusted_influence_results.rds`
- *rule*: drop rows whose pid matches any influential id; pass through unchanged if none

```{r}
#| label: save-adjusted-data
#| code-fold: show
#| column: body-outset-right

# identify rows to exclude based on influential observation flags
rows_to_drop <- match_ids_to_rows(unadjusted_data, influence_results$influential_ids)

# create sensitivity-filtered dataset
if (length(rows_to_drop) == 0) {
  adjusted_data <- unadjusted_data
} else {
  adjusted_data <- unadjusted_data[-rows_to_drop, , drop = FALSE]
}

# define output directories
output_dir <- here::here("assets", "data", "saved")
models_dir <- here::here("assets", "models")

# save outputs
saveRDS(adjusted_data, file = file.path(output_dir, "adjusted_data.rds"))
saveRDS(influence_results, file = file.path(models_dir, "unadjusted_influence_results.rds"))
```

```{r}
#| label: export-summary
#| results: asis
#| code-fold: true
#| column: body-outset-right

export_msg <- glue(
  "<div class='callout-note'>
<strong>Exported Files</strong><br><br>
<strong>Sensitivity dataset:</strong> <code>adjusted_data.rds</code> ({nrow(adjusted_data)} observations after filtering)<br>
<strong>Influence diagnostics:</strong> <code>unadjusted_influence_results.rds</code><br>
<strong>Observations flagged:</strong> {num_influential} ({pct_influential}%)<br>
<strong>Observations retained:</strong> {nrow(adjusted_data)} ({round(100 * nrow(adjusted_data) / data_size, 1)}%)
</div>"
)
cat(export_msg)
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                                              | Content                                             |
|---------------------------------------------------|-----------------------------------------------------|
| `assets/data/saved/adjusted_data.rds`             | sensitivity-filtered dataset (flagged rows removed) |
| `assets/models/unadjusted_influence_results.rds`  | per-observation flags, thresholds, influential IDs  |

**Next:** [Step 9 (Adjusted Models)](/analysis/models/adjusted/index.qmd)
