---
title: "Adjusted Regression Models: Robust Estimates"
subtitle: "Modeling After Removing Influential Observations"
author: "Siard van den Bosch"
date: 2025-02-18
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 9};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Refit the Step 7 nested model set on the outlier-trimmed sensitivity dataset |
| **Inputs**     | `adjusted_data.rds`, `models.rds` |
| **Outputs**    | `adjusted_fit_{0..11}.rds`, `adjusted_fits.rds` |
| **Method**     | MM-estimation; gender-stratified parallel bootstrap |
| **Requires**   | [Step 7 (Unadjusted Models)](/analysis/models/unadjusted/index.qmd), [Step 8 (Influence Diagnostics)](/analysis/models/assumptions/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body

# clear global environment, garbage collect, and detach packages
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions (consistent with steps 7-8)
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "numeric_df", "fit_robust_model", "standardize", "named_list",
  "set_globals", "import_pipes", "fit_models", "prepare_model_matrices",
  "sci_to_tex", "create_lmrob_control", "process_model", "save_function"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()

# global recompute switch
# TRUE re-runs the heavy steps; FALSE loads cached assets when they exist
recompute_heavy <- FALSE

# define analysis type
type <- "adjusted"
options(width = 250)
```

```{r}
#| label: library-imports
#| message: false
#| code-fold: true
#| column: body

library(compiler)
library(parallel)
library(robustbase)
library(glue)
```

## Analysis Parameters

```{r}
#| label: analysis-parameters
#| code-fold: show
#| column: body

# bootstrap configuration
bootstrap_params <- list(
  n_boots   = 5000L,
  pct       = 0.95,
  alpha     = 0.05,
  delta     = 0.01,
  seed      = 625L,
  max_cores = 8L
)

# robust regression configuration
robust_params <- list(
  psi            = "bisquare",
  std_data       = TRUE,
  coef_threshold = 1e4
)
```

## Bootstrap Functions

```{r}
#| label: fn-min-bootstrap-iterations
#| code-fold: true
#| column: body

min_bootstrap_iterations <- function(data, statistic = mean, alpha = 0.05, delta = 0.01) {
  if (!inherits(data, c("data.frame", "matrix", "numeric", "integer"))) {
    stop("'data' must be data.frame, matrix, or numeric vector.")
  }
  statistic <- match.fun(statistic)
  bootstrap_data <- if (inherits(data, c("data.frame", "matrix"))) data[, 1] else data

  n <- length(bootstrap_data)
  prelim_boots <- 100L
  prelim_stats <- replicate(
    prelim_boots,
    statistic(sample(bootstrap_data, n, replace = TRUE))
  )

  var_boot <- stats::var(prelim_stats)
  za2 <- stats::qnorm(1 - alpha / 2)^2
  n_boot <- ceiling(za2 * var_boot / (delta^2))

  max(n_boot, 100L)
}
```

```{r}
#| label: fn-bootstrap-pvalues
#| code-fold: true
#| column: body

bootstrap_pvalues <- function(bootstrap_estimates, observed_coef) {
  if (is.null(bootstrap_estimates) || ncol(bootstrap_estimates) == 0) {
    return(rep(NA_real_, length(observed_coef)))
  }
  purrr::map_dbl(seq_along(observed_coef), function(j) {
    valid <- bootstrap_estimates[!is.na(bootstrap_estimates[, j]), j]
    n_valid <- length(valid)
    p_low <- (sum(valid <= 0) + 1) / (n_valid + 1)
    p_high <- (sum(valid >= 0) + 1) / (n_valid + 1)
    min(1, 2 * min(p_low, p_high))
  })
}
```

```{r}
#| label: fn-bootstrap-linear-model
#| code-fold: true
#| column: body

bootstrap_linear_model <- function(indices, dm, resp, ctrl = NULL, type = "robust",
                                   coef_threshold = 1e4) {
  tryCatch(
    {
      if (type == "robust") {
        fit <- robustbase::lmrob.fit(
          x = dm[indices, , drop = FALSE],
          y = resp[indices],
          control = ctrl
        )
        if (is.null(fit$coefficients) || any(abs(fit$coefficients) > coef_threshold)) {
          alt_ctrl <- robustbase::lmrob.control(method = "S", efficiency = 0.85, max.it = 100)
          fit <- robustbase::lmrob.fit(dm[indices, , drop = FALSE], resp[indices], alt_ctrl)
        }
      } else {
        fit <- stats::lm.fit(dm[indices, , drop = FALSE], resp[indices])
      }

      if (is.null(fit$coefficients)) {
        return(rep(NA_real_, ncol(dm)))
      }
      if (any(abs(fit$coefficients) > coef_threshold)) {
        return(rep(NA_real_, ncol(dm)))
      }
      fit$coefficients
    },
    error = function(e) rep(NA_real_, ncol(dm))
  )
}
```

```{r}
#| label: fn-evaluate-boot-stats
#| code-fold: true
#| column: body

evaluate_boot_stats <- function(boot_matrix, min_boots, pct = 0.95) {
  if (!is.matrix(boot_matrix)) stop("boot_matrix must be a matrix")
  if (pct <= 0 || pct >= 1) stop("pct must be between 0 and 1")

  num_samples <- nrow(boot_matrix)
  valid_samples <- sum(stats::complete.cases(boot_matrix))

  if (num_samples < min_boots) {
    warning(glue::glue("Only {num_samples} bootstrap samples, below minimum required ({min_boots})."))
  }
  if (valid_samples < min_boots) {
    warning(glue::glue("Only {valid_samples} valid samples, below minimum required ({min_boots})."))
  }

  probs <- c((1 - pct) / 2, 1 - (1 - pct) / 2)
  conf_int <- t(apply(boot_matrix, 2, stats::quantile, probs = probs, na.rm = TRUE))

  list(
    conf_int = conf_int,
    bootstrap_estimates = boot_matrix,
    valid_samples = valid_samples
  )
}
```

```{r}
#| label: fn-weighted-bootstrap
#| code-fold: true
#| column: body

weighted_bootstrap <- function(sample_size, response, strata,
                               outlier_percentile = 0.9, min_stratum_size = 5) {
  # plain stratified resampling: draw within strata proportional to stratum size
  # response, outlier_percentile, and min_stratum_size are retained for
  # signature compatibility and intentionally unused; conditioning the resample
  # on the response distribution understates bootstrap variability
  if (!is.factor(strata)) strata <- as.factor(strata)

  strata_levels <- levels(strata)
  strata_counts <- table(strata)
  strata_prop <- strata_counts / length(strata)

  strata_sizes <- floor(strata_prop * sample_size)
  shortfall <- sample_size - sum(strata_sizes)
  if (shortfall > 0) {
    additional <- sample(strata_levels, shortfall, prob = strata_prop)
    for (lev in additional) strata_sizes[lev] <- strata_sizes[lev] + 1
  }

  samples <- purrr::map(strata_levels, function(level) {
    idx <- which(strata == level)
    n_level <- strata_sizes[level]
    if (n_level == 0 || length(idx) == 0) {
      return(integer(0))
    }
    sample(idx, size = n_level, replace = TRUE)
  })

  do.call(c, samples)
}
```

```{r}
#| label: fn-parallel-bootstrap
#| code-fold: true
#| column: body

parallel_bootstrap <- function(data, n_boots, robust_fit, ols_fit, resp, dm, ctrl,
                               min_boots, pct = 0.95, max_cores = 8L, seed = 625L) {
  if (!("woman" %in% names(data))) stop("Stratification variable 'woman' not found")
  stopifnot(length(resp) == nrow(data), nrow(dm) == nrow(data))

  strata <- as.factor(data$woman)
  n_cores <- min(parallel::detectCores(), max_cores)
  cl <- parallel::makeCluster(n_cores)
  on.exit(parallel::stopCluster(cl), add = TRUE)
  parallel::clusterSetRNGStream(cl, iseed = seed)

  parallel::clusterExport(
    cl,
    varlist = c(
      "data", "resp", "dm", "robust_fit", "ols_fit", "ctrl",
      "bootstrap_linear_model", "strata", "weighted_bootstrap"
    ),
    envir = environment()
  )
  parallel::clusterEvalQ(cl, library(robustbase))

  chunk_size <- min(500L, n_boots)
  n_chunks <- ceiling(n_boots / chunk_size)

  robust_results <- vector("list", n_boots)
  ols_results <- vector("list", n_boots)
  result_idx <- 0L

  for (chunk in seq_len(n_chunks)) {
    chunk_start <- (chunk - 1L) * chunk_size + 1L
    chunk_end <- min(chunk * chunk_size, n_boots)
    chunk_size_actual <- chunk_end - chunk_start + 1L

    message(glue::glue("Processing bootstrap chunk {chunk}/{n_chunks}..."))

    boot_indices_chunk <- parallel::parLapply(
      cl, seq_len(chunk_size_actual),
      function(...) weighted_bootstrap(nrow(data), resp, strata)
    )

    if (!is.null(robust_fit)) {
      robust_chunk <- parallel::parLapply(cl, boot_indices_chunk, function(idxs) {
        bootstrap_linear_model(idxs, dm, resp, ctrl = ctrl, type = "robust")
      })
      for (i in seq_along(robust_chunk)) {
        robust_results[[result_idx + i]] <- robust_chunk[[i]]
      }
    }

    ols_chunk <- parallel::parLapply(cl, boot_indices_chunk, function(idxs) {
      bootstrap_linear_model(idxs, dm, resp, ctrl = NULL, type = "ols")
    })
    for (i in seq_along(ols_chunk)) {
      ols_results[[result_idx + i]] <- ols_chunk[[i]]
    }

    result_idx <- result_idx + chunk_size_actual
  }

  robust_out <- if (!is.null(robust_fit)) {
    robust_matrix <- do.call(rbind, robust_results[seq_len(n_boots)])
    evaluate_boot_stats(robust_matrix, min_boots, pct)
  } else {
    list(conf_int = NULL, bootstrap_estimates = NULL, valid_samples = 0L)
  }

  ols_matrix <- do.call(rbind, ols_results[seq_len(n_boots)])
  ols_out <- evaluate_boot_stats(ols_matrix, min_boots, pct)

  list(
    robust_out = robust_out,
    ols_out = ols_out,
    n_boots = n_boots,
    completed_boots = list(
      robust = if (is.null(robust_fit)) 0L else nrow(do.call(rbind, robust_results[seq_len(n_boots)])),
      ols = nrow(ols_matrix)
    )
  )
}
```

```{r}
#| label: fn-process-model-bootstrap
#| code-fold: true
#| column: body

# bootstrap-enabled process_model (overrides sourced version)
process_model_bootstrap <- function(fit, type, n_boots, min_boots, data, resp, dm, ctrl, pct, seed = 625L) {
  if (is.null(fit)) {
    return(NULL)
  }

  model_output <- list(
    coefficients = coef(fit),
    standard_errors = sqrt(diag(stats::vcov(fit))),
    confidence_intervals = NULL,
    p_values = NULL,
    bootstrap_estimates = NULL,
    bootstrap_quality_score = 0,
    valid_bootstrap_count = 0L,
    bootstrap_convergence_rate = 0,
    diagnostics = list()
  )

  if (n_boots > 0) {
    message(glue::glue("Running {n_boots} bootstrap iterations for {type} model..."))

    robust_fit_for_boot <- if (type == "robust") fit else NULL
    ols_fit_for_boot <- if (type == "ols") fit else NULL

    boot_results <- tryCatch(
      {
        parallel_bootstrap(
          data = data, n_boots = n_boots,
          robust_fit = robust_fit_for_boot, ols_fit = ols_fit_for_boot,
          resp = resp, dm = dm, ctrl = ctrl, min_boots = min_boots, pct = pct, seed = seed
        )
      },
      error = function(e) {
        warning(glue::glue("Bootstrap failed: {e$message}"))
        NULL
      }
    )

    if (!is.null(boot_results)) {
      boot_out <- if (type == "robust") boot_results$robust_out else boot_results$ols_out

      if (!is.null(boot_out) && !is.null(boot_out$bootstrap_estimates)) {
        model_output$bootstrap_estimates <- boot_out$bootstrap_estimates
        model_output$confidence_intervals <- boot_out$conf_int
        model_output$valid_bootstrap_count <- boot_out$valid_samples
        model_output$bootstrap_convergence_rate <- boot_out$valid_samples / n_boots
        model_output$p_values <- bootstrap_pvalues(
          boot_out$bootstrap_estimates, model_output$coefficients
        )
        model_output$bootstrap_quality_score <- boot_out$valid_samples / n_boots
      }
    }
  }

  model_output
}
```

```{r}
#| label: fn-robust-analysis-bootstrap
#| code-fold: show
#| column: body

# main analysis function with bootstrap support
robust_analysis_bootstrap <- function(data, form, n_boots = NULL, pct = 0.95,
                                      alpha = 0.05, delta = 0.01, ctrl = NULL,
                                      seed = 625, psi = "bisquare", std_data = TRUE) {
  set.seed(seed, kind = "default")

  if (std_data) data <- standardize(data)

  mm_list <- prepare_model_matrices(form, data)
  dm <- mm_list$design_matrix
  resp <- mm_list$response

  min_boots <- if (!is.null(n_boots) && n_boots > 0) {
    message(glue::glue("Bootstrap enabled with {n_boots} iterations."))
    n_boots
  } else {
    n_boots <- 0
    message("Bootstrapping disabled. Fitting robust and OLS models only.")
    NA_integer_
  }

  ctrl <- create_lmrob_control(ctrl, psi)
  fits <- fit_models(data, form, ctrl)

  robust_results <- process_model_bootstrap(
    fits$robust_fit, "robust", n_boots, min_boots, data, resp, dm, ctrl, pct,
    seed = seed
  )
  ols_results <- process_model_bootstrap(
    fits$ols_fit, "ols", n_boots, min_boots, data, resp, dm, ctrl, pct,
    seed = seed
  )

  analysis_results <- list(
    robust_model_results = list(
      model_object = fits$robust_fit,
      analysis_output = robust_results
    ),
    ols_model_results = list(
      model_object = fits$ols_fit,
      analysis_output = ols_results
    ),
    analysis_metadata = list(
      total_observations = nrow(data),
      number_of_predictors = ncol(dm),
      bootstrap_iterations = n_boots,
      minimum_required_bootstraps = min_boots,
      analysis_timestamp = Sys.time(),
      robustbase_version = utils::packageVersion("robustbase")
    ),
    detailed_outputs = list(robust = robust_results, ols = ols_results)
  )

  class(analysis_results) <- "structured_models"
  analysis_results
}
```

```{r}
#| label: byte-compilation
#| code-fold: true
#| column: body

# byte compile performance-critical functions
min_bootstrap_iterations <- compiler::cmpfun(min_bootstrap_iterations)
bootstrap_pvalues <- compiler::cmpfun(bootstrap_pvalues)
bootstrap_linear_model <- compiler::cmpfun(bootstrap_linear_model)
evaluate_boot_stats <- compiler::cmpfun(evaluate_boot_stats)
weighted_bootstrap <- compiler::cmpfun(weighted_bootstrap)
parallel_bootstrap <- compiler::cmpfun(parallel_bootstrap)
process_model_bootstrap <- compiler::cmpfun(process_model_bootstrap)
robust_analysis_bootstrap <- compiler::cmpfun(robust_analysis_bootstrap)
```

## Load Data and Model Specifications

```{r}
#| label: load-inputs
#| code-fold: show
#| column: body

# paths to input files
data_path <- here::here("assets", "data", "saved", "adjusted_data.rds")
models_path <- here::here("assets", "models", "models.rds")
models_dir <- here::here("assets", "models")

# consolidated output path (gated by recompute_heavy below)
consolidated_path <- file.path(models_dir, glue::glue("{type}_fits.rds"))

# validate inputs exist
if (!file.exists(data_path)) stop("Missing input from Step 8: ", data_path)
if (!file.exists(models_path)) stop("Missing model specifications from Step 7: ", models_path)

# load inputs
adjusted_data <- readRDS(data_path)
model_specs <- readRDS(models_path)

# ensure models directory exists
if (!dir.exists(models_dir)) dir.create(models_dir, recursive = TRUE)
```

```{r}
#| label: verify-step8-data
#| results: asis
#| code-fold: true

n_obs <- nrow(adjusted_data)
n_vars <- ncol(adjusted_data)
n_models <- length(model_specs)

verify_msg <- glue::glue(
  "<div class='callout-note'>
<strong>Step 8 Data Verification</strong><br><br>
<strong>Input:</strong> <code>adjusted_data.rds</code> ({n_obs} observations, {n_vars} variables)<br>
<strong>Source:</strong> Sensitivity dataset from Step 8 (influential observations removed)<br>
<strong>Models:</strong> {n_models} model specifications loaded from Step 7<br>
<strong>Bootstrap iterations:</strong> {bootstrap_params$n_boots}
</div>"
)
cat(verify_msg)
```

## Model-Hypothesis Mapping

**Baseline and Composite SES Models (Models 0–6):**

| Model | Description                 | Tests                     | Hypotheses  |
| ----- | --------------------------- | ------------------------- | ----------- |
| 0     | Null (intercept only)       | Baseline                  | –           |
| 1     | Adult covariates + own SES  | Adjustment set            | –           |
| 2     | + Parental SES (composite)  | Parental SES main effects | H1, RQ1     |
| 3     | + Partner SES               | Partner SES main effects  | H2, RQ2     |
| 4     | + Gender × SES interactions | Gender moderation         | H3, H4, RQ3 |
| 5     | + Parental × Partner SES    | SES synergy               | H5, RQ4     |
| 6     | + Gender × synergy          | Gender-moderated synergy  | H6, RQ4     |

**Disaggregated SES Models (Models 7–11):**

| Model | Description                          | Tests                         | Hypotheses  |
| ----- | ------------------------------------ | ----------------------------- | ----------- |
| 7     | Covariates + Mother SES + Father SES | Differential parental effects | H1, RQ1     |
| 8     | + Partner SES (disaggregated)        | Partner SES main effects      | H2, RQ2     |
| 9     | + Gender × Mother/Father/Partner SES | Gender moderation by parent   | H3, H4, RQ3 |
| 10    | + Mother × Partner, Father × Partner | Parent-specific synergy       | H5, RQ4     |
| 11    | + Gender × synergy terms             | Gender-moderated synergy      | H6, RQ4     |

## Fit All Models

```{r}
#| label: fit-all-models
#| message: false
#| code-fold: show
#| column: body

# reuse cached fits when available, otherwise refit the full nested set
if (!recompute_heavy && file.exists(consolidated_path)) {
  message("cached fits found; skipping model fitting (set recompute_heavy <- TRUE to refit)")
} else {
  # fit all models using purrr::imap for named iteration
  all_fits <- purrr::imap(model_specs, function(model_formula, model_name) {
    message(glue::glue("\n{strrep('=', 60)}\nFitting {model_name}...\n"))

    fit_result <- robust_analysis_bootstrap(
      data      = adjusted_data,
      form      = model_formula,
      n_boots   = bootstrap_params$n_boots,
      pct       = bootstrap_params$pct,
      alpha     = bootstrap_params$alpha,
      delta     = bootstrap_params$delta,
      ctrl      = NULL,
      seed      = bootstrap_params$seed,
      psi       = robust_params$psi,
      std_data  = robust_params$std_data
    )

    # save individual fit
    fit_filename <- glue::glue("{type}_fit_{gsub('model_', '', model_name)}.rds")
    saveRDS(fit_result, file.path(models_dir, fit_filename))

    fit_result
  })

  # assign to individual objects for compatibility with downstream steps
  # vectorised, no loop: rename model_* -> fit_* and bind into the chunk environment
  list2env(setNames(all_fits, sub("^model", "fit", names(all_fits))), environment())
}
```

## Save Consolidated Model Fits

```{r}
#| label: save-all-fits
#| message: false
#| code-fold: show
#| column: body

# rebuild and persist the consolidated list only when (re)fitting
if (recompute_heavy || !file.exists(consolidated_path)) {
  consolidated_fits <- named_list(
    fit_0, fit_1, fit_2, fit_3, fit_4, fit_5, fit_6,
    fit_7, fit_8, fit_9, fit_10, fit_11
  )

  saveRDS(consolidated_fits, consolidated_path)
}
```

```{r}
#| label: export-summary
#| results: asis
#| code-fold: true

# display a project-relative path so rendered pages never expose the
# local filesystem layout and stay portable across machines
models_dir_rel <- sub(paste0(here::here(), "/"), "", models_dir, fixed = TRUE)

export_msg <- glue::glue(
  "<div class='callout-note'>
<strong>Exported Files</strong><br><br>
<strong>Model specifications:</strong> Loaded from Step 7 ({length(model_specs)} models)<br>
<strong>Analysis dataset:</strong> <code>adjusted_data.rds</code> ({nrow(adjusted_data)} observations)<br>
<strong>Individual fits:</strong> <code>adjusted_fit_0.rds</code> through <code>adjusted_fit_11.rds</code><br>
<strong>Consolidated fits:</strong> <code>adjusted_fits.rds</code><br><br>
<strong>Composite models (2–6):</strong> Use education_parental, isei08_parental, education_partner, isei08_partner<br>
<strong>Disaggregated models (7–11):</strong> Use education/isei08 × mother/father/partner<br><br>
All files saved to: <code>{models_dir_rel}</code>
</div>"
)
cat(export_msg)
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                                     | Content                                             |
| ---------------------------------------- | --------------------------------------------------- |
| `assets/models/adjusted_fit_{0..11}.rds` | Per-model robust fits with bootstrap output         |
| `assets/models/adjusted_fits.rds`        | Named list consolidating all 12 adjusted model fits |

**Next:** [Step 10 (Initial Model Evaluation)](/analysis/models/goodness-of-fit/model-fit/index.qmd)
