---
title: "Missing Data Imputation"
subtitle: "Random Forest Approach via `missForest`"
author: "Siard van den Bosch"
date: 2024-11-20
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 5};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Impute missing values via random-forest imputation on partnered cases |
| **Inputs**     | `data_without_outliers.sav` |
| **Outputs**    | `imputed_data_combined.sav`, `imputed_results_missforest.rds` |
| **Method**     | `missForest` random-forest imputation; variablewise OOB error  |
| **Requires**   | [Step 4 (Outliers Detection)](/analysis/data/outliers/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- `missForest`: handles mixed numeric and factor variables without distributional assumptions [@stekhovenbuhlmann2012]
- Partition by partnership: partner variables have structural NA for non-partnered cases and must not be imputed [@littlerubin2019]
- `variablewise = TRUE`: yields per-variable OOB error for targeted diagnostics; imputed values are unaffected
- Serial execution (`parallelize = "no"`): preserves variablewise OOB reporting; `parallelize = "variables"` returns zero-valued OOB in current `missForest`
- OOB error: bootstrap aggregation in random forests provides internal validation [@breiman2001]
- NRMSE and PFC: standard fidelity metrics for mixed-type random-forest imputation [@tangishwaran2017]
:::

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear global env and force full garbage collection
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))

# detach non-base packages to avoid namespace interference
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions for data handling and analysis
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "set_globals", "summarize_na", "as_kable", "describe", "copy_labels",
  "add_style", "md_to_html", "import_pipes", "footnote", "format_p_value",
  "beautify_names", "row_grouping", "numeric_df"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()

# load dataset excluding outliers from step 4
data_without_outliers <- haven::read_sav(
  here::here("assets", "data", "saved", "data_without_outliers.sav")
)
```

## Verify Dataset from Step 4

```{r}
#| label: verify-step4-data
#| results: asis
#| code-fold: show
#| column: body-outset-right

# verify expected row count and key variables from step 4
n_obs <- nrow(data_without_outliers)
required_vars <- c("idnummer", "partner", "education", "isei08", "dhd_index")
vars_present <- required_vars[required_vars %in% names(data_without_outliers)]
vars_missing <- setdiff(required_vars, vars_present)

if (length(vars_missing) > 0) {
  stop(
    "Missing required variables from Step 4: ",
    paste(vars_missing, collapse = ", ")
  )
}

# count partnered vs non-partnered for context
n_partnered <- sum(data_without_outliers$partner == 1, na.rm = TRUE)
n_non_partnered <- sum(data_without_outliers$partner == 0, na.rm = TRUE)

htmltools::HTML(glue(
  "<div class='callout-note'>
  <strong>Step 4 Data Verification</strong><br><br>
  <strong>Observations:</strong> {n_obs}<br>
  <strong>Partnered respondents:</strong> {n_partnered} (receive imputation)<br>
  <strong>Non-partnered respondents:</strong> {n_non_partnered} ",
  "(held out, no partner data to impute)<br>
  <strong>Required variables present:</strong> ",
  "<code>{paste(vars_present, collapse = '</code>, <code>')}</code>
  </div>"
))
```

## Configure Dataset for Imputation

- *in*: `data_without_outliers`
- *out*: `df_impute` (partnered subsample, prepared), `df_non_partnered` (held out), `passthrough_frame` (intact passthrough values)
- *rule*: filter `partner == 1` for imputation; convert columns with `<= 5` unique values to factors

```{r}
#| label: imputation-setup
#| code-fold: show
#| column: body-outset-right

# rename variables for clarity and consistency with pipeline conventions
df_with_missings <- data_without_outliers %>%
  dplyr::rename(
    education_self = education,
    isei08_self = isei08
  )

# nominal classification codes, egp class codes, character labour-force
# status, and derived composites are excluded from imputation: codes and
# classes are not interval-scaled, lf_status is character, and composites
# are recomputed from imputed components in step 6
passthrough_vars <- grep(
  "^isco08r|^brc14|^egp11|^lf_status|_parental$",
  names(df_with_missings),
  value = TRUE
)

# capture the intact passthrough values before any numeric coercion:
# numeric_df converts character columns with as.numeric, which would turn
# the lf_status columns into all-NA
passthrough_frame <- df_with_missings %>%
  dplyr::select(idnummer, dplyr::all_of(passthrough_vars))

# extract non-partnered data (no partner variables to impute); passthrough
# columns are dropped so both branches align at the final bind
df_non_partnered <- df_with_missings %>%
  numeric_df() %>%
  dplyr::filter(partner == 0) %>%
  dplyr::select(-dplyr::all_of(passthrough_vars))

# extract partnered data for imputation
df_partnered <- df_with_missings %>%
  numeric_df() %>%
  dplyr::filter(partner == 1)

# prepare data for missForest: remove identifiers, convert low-cardinality to factors
df_impute <- df_partnered %>%
  dplyr::select(
    -idnummer, -partner,
    -dplyr::all_of(passthrough_vars)
  ) %>%
  haven::zap_labels() %>%
  droplevels() %>%
  as.data.frame() %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::where(~ dplyr::n_distinct(.x, na.rm = TRUE) <= 5L),
      as.factor
    )
  )

# validate column types for missForest compatibility
bad_cols <- names(df_impute)[!(vapply(df_impute, is.numeric, logical(1)) |
  vapply(df_impute, is.factor, logical(1)))]
if (length(bad_cols)) {
  stop(
    "missForest requires numeric or factor columns: ",
    paste(bad_cols, collapse = ", ")
  )
}

# check for completely missing variables (missForest cannot impute these)
all_na <- names(df_impute)[colSums(!is.na(df_impute)) == 0]
if (length(all_na)) {
  stop("Variables with all values missing: ", paste(all_na, collapse = ", "))
}
```

## Define Imputation Parameters

```{r}
#| label: imputation-parameters
#| code-fold: show
#| column: body-outset-right

# imputation hyperparameters
imputation_params <- list(
  seed = 625L,
  ntree = 200L,
  mtry = floor(sqrt(ncol(df_impute))),
  maxiter = 10L,
  variablewise = TRUE,
  replace = TRUE
)
```

## Run missForest Imputation

```{r}
#| label: imputation-missforest
#| code-fold: show
#| column: body-outset-right

# set seed for reproducibility
set.seed(imputation_params$seed)

# choose a safe number of workers (reserve one core for system)
n_workers <- max(1L, parallel::detectCores(logical = FALSE) - 1L)

# determine parallelization strategy based on problem size
missing_rate <- vapply(df_impute, function(col) mean(is.na(col)), numeric(1))
n_miss_vars <- sum(missing_rate > 0)

# IMPORTANT: parallelize = "variables" together with variablewise = TRUE
# returns a length-correct but all-zero OOBerror vector in current missForest
# (the imputation itself completes correctly but OOB diagnostics are useless).
# Serial execution preserves per-variable OOB reporting. Runtime impact is
# modest for this dataset (911 rows, ~31 variables, 200 trees): tens of
# seconds to about a minute on a modern laptop.
parallelize <- "no"

# detect missForest capabilities (legacy randomForest vs ranger backend)
mf_formals <- names(formals(missForest::missForest))
has_backend <- "backend" %in% mf_formals
has_num_threads <- "num.threads" %in% mf_formals

# do not pass a backend argument; let missForest use its default.
# Earlier attempts to explicitly set backend = "ranger" or to suppress it in
# combination with parallelize = "variables" did not restore OOB reporting;
# the serial parallelize = "no" path above is what fixes it.
backend <- NULL

# set up foreach backend only when missForest needs it
use_foreach <- parallelize == "variables" ||
  (parallelize == "forests" && !has_num_threads)

if (use_foreach) {
  clust <- parallel::makeCluster(n_workers)
  on.exit(parallel::stopCluster(clust), add = TRUE)
  doParallel::registerDoParallel(clust)
  doRNG::registerDoRNG(seed = imputation_params$seed)
}

# assemble missForest arguments
mf_args <- list(
  xmis = df_impute,
  ntree = imputation_params$ntree,
  mtry = imputation_params$mtry,
  maxiter = imputation_params$maxiter,
  variablewise = imputation_params$variablewise,
  verbose = TRUE,
  parallelize = parallelize,
  replace = imputation_params$replace
)

if (has_num_threads && parallelize == "forests") mf_args$num.threads <- n_workers

# execute imputation
imp_mf <- do.call(missForest::missForest, mf_args)

# store results with metadata for documentation
imputed_results <- list(
  imp = imp_mf,
  seed = imputation_params$seed,
  n_workers = n_workers,
  parallelize = parallelize,
  backend = backend,
  ntree = imputation_params$ntree,
  mtry = imputation_params$mtry,
  maxiter = imputation_params$maxiter
)

saveRDS(
  imputed_results,
  here::here("assets", "data", "saved", "imputed_results_missforest.rds")
)
```

## Calculate Imputation Performance Metrics

```{r}
#| label: define-metrics-table
#| code-fold: true
#| column: body-outset-right

# table metadata for OOB metrics display
footnote_text <- glue(
  "OOB = out-of-bag. Numeric variables report OOB MSE when variablewise = TRUE; ",
  "factor variables report OOB PFC (proportion falsely classified). ",
  "For numeric variables, MSE is rescaled to approximate NRMSE: ",
  "sqrt(MSE) / SD(observed). ",
  "Lower values indicate better OOB performance. ",
  "Settings: ntree = {imputed_results$ntree}, mtry = {imputed_results$mtry}, ",
  "maxiter = {imputed_results$maxiter}, parallelize = {imputed_results$parallelize}."
)

col_names <- c(
  "Variable", "Error Metric", "OOB Imputation Error", "Missing Rate",
  "Rating"
)
```

```{r}
#| label: imputation-metrics
#| code-fold: show
#| column: body-outset-right

# extract imputation results for analysis
imp_mf <- imputed_results$imp

# calculate missing rates for all variables
missing_info <- data.frame(
  var = names(df_impute),
  missing_rate = vapply(df_impute, function(col) mean(is.na(col)), numeric(1)),
  stringsAsFactors = FALSE
)

# compute standard deviations for numeric variables (observed values only)
num_vars <- names(df_impute)[vapply(df_impute, is.numeric, logical(1))]
lookup_sd <- data.frame(
  var = num_vars,
  sd_obs = vapply(df_impute[num_vars], stats::sd, numeric(1), na.rm = TRUE),
  stringsAsFactors = FALSE
)

# assign variable names to OOB errors
raw_oob <- imp_mf$OOBerror
names(raw_oob) <- colnames(df_impute)

# create base dataframe for variable-wise metrics
mf_metrics <- data.frame(
  var = colnames(df_impute),
  erm = ifelse(vapply(df_impute, is.numeric, logical(1)), "NRMSE (approx)", "PFC"),
  oob = as.numeric(raw_oob),
  stringsAsFactors = FALSE
) %>%
  dplyr::left_join(lookup_sd, by = "var") %>%
  dplyr::left_join(missing_info, by = "var") %>%
  dplyr::mutate(
    # for numeric variables, convert OOB MSE to approximate NRMSE
    oob = dplyr::case_when(
      erm == "NRMSE (approx)" & is.finite(sd_obs) & sd_obs > 0 ~ sqrt(oob) / sd_obs,
      erm == "NRMSE (approx)" ~ NA_real_,
      TRUE ~ oob
    ),
    # add a compact rating to prioritize review
    rating = dplyr::case_when(
      missing_rate == 0 ~ "Complete",
      oob < 0.5 ~ "Excellent",
      oob < 0.7 ~ "Good",
      oob < 0.9 ~ "Acceptable",
      TRUE ~ "Review"
    )
  ) %>%
  dplyr::select(var, erm, oob, missing_rate, rating)

# compute a global, missingness-weighted summary from variable-wise metrics
global_df <- mf_metrics %>%
  dplyr::filter(missing_rate > 0) %>%
  dplyr::group_by(erm) %>%
  dplyr::summarise(
    oob = stats::weighted.mean(oob, w = missing_rate, na.rm = TRUE),
    missing_rate = mean(missing_rate),
    rating = "Global (weighted)",
    .groups = "drop"
  ) %>%
  dplyr::mutate(var = "Global") %>%
  dplyr::select(var, erm, oob, missing_rate, rating)

# prepare table for display (exclude complete variables)
mf_metrics_tbl <- dplyr::bind_rows(mf_metrics, global_df) %>%
  dplyr::filter(missing_rate > 0 | var == "Global") %>%
  dplyr::mutate(var = beautify_names(var)) %>%
  dplyr::arrange(dplyr::desc(var == "Global"), erm, var)

# display the metrics table
mf_metrics_tbl %>%
  as_kable(cols = col_names, digits = 3) %>%
  footnote() %>%
  add_style()
```

## Visualize Imputation Diagnostics

```{r}
#| label: imputation-diagnostic-plot-function
#| code-fold: show
#| column: body-outset-right

# function to create density comparison plots for imputation diagnostics
plot_imputation_diagnostic <- function(orig_data, imputed_data, variable) {
  x_orig <- orig_data[[variable]]
  x_imp <- imputed_data[[variable]]

  miss_idx <- is.na(x_orig)
  n_miss <- sum(miss_idx)
  n_tot <- length(x_orig)

  # nothing to diagnose if no missingness
  if (n_miss == 0) {
    return(invisible(NULL))
  }

  # caption text showing missing rate
  caption_txt <- sprintf("Missing: %d (%.1f%%)", n_miss, 100 * n_miss / n_tot)

  # factor/character path: bar chart of proportions
  if (is.factor(x_orig) || is.character(x_orig)) {
    x_orig_f <- as.factor(x_orig)
    x_imp_f <- as.factor(x_imp)

    # observed distribution (excluding NAs)
    obs_tab <- prop.table(table(x_orig_f[!is.na(x_orig_f)]))

    # imputed-at-missing distribution
    imp_tab <- prop.table(table(x_imp_f[miss_idx]))

    levs <- union(names(obs_tab), names(imp_tab))
    df_plot <- data.frame(
      level = levs,
      observed = as.numeric(obs_tab[levs]),
      imputed_missing = as.numeric(imp_tab[levs])
    )
    df_plot[is.na(df_plot)] <- 0

    df_long <- df_plot %>%
      tidyr::pivot_longer(
        cols = c("observed", "imputed_missing"),
        names_to = "source",
        values_to = "prop"
      ) %>%
      dplyr::mutate(
        source = factor(
          source,
          levels = c("observed", "imputed_missing"),
          labels = c("Observed (non-missing)", "Imputed (missing only)")
        )
      )

    print(
      ggplot2::ggplot(df_long, ggplot2::aes(x = level, y = prop, fill = source)) +
        ggplot2::geom_col(position = "dodge") +
        ggplot2::labs(
          title = gsub("<.*?>", "", beautify_names(variable)),
          x = gsub("<.*?>", "", beautify_names(variable)),
          y = "Proportion",
          fill = NULL,
          caption = caption_txt
        ) +
        ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
    )

    return(invisible(NULL))
  }

  # numeric path: density comparison
  x_orig_num <- as.numeric(x_orig)
  x_obs <- x_orig_num[!miss_idx]
  x_imp_miss <- as.numeric(x_imp)[miss_idx]

  # keep only finite values
  x_obs <- x_obs[is.finite(x_obs)]
  x_imp_miss <- x_imp_miss[is.finite(x_imp_miss)]

  # if too few points for density, fall back to point plot
  if (length(x_imp_miss) < 2L || length(x_obs) < 2L) {
    df_fallback <- data.frame(
      value = c(x_obs, x_imp_miss),
      source = c(
        rep("Observed (non-missing)", length(x_obs)),
        rep("Imputed (missing only)", length(x_imp_miss))
      )
    )

    print(
      ggplot2::ggplot(df_fallback, ggplot2::aes(x = value, y = source)) +
        ggplot2::geom_point(alpha = 0.6) +
        ggplot2::labs(
          title = gsub("<.*?>", "", beautify_names(variable)),
          x = gsub("<.*?>", "", beautify_names(variable)),
          y = NULL,
          caption = paste0(caption_txt, " | density skipped (too few points)")
        )
    )

    return(invisible(NULL))
  }

  # compute kernel density estimates
  dens_obs <- stats::density(x_obs)
  dens_imp <- stats::density(x_imp_miss)

  df_dens <- data.frame(
    x = dens_obs$x,
    observed = dens_obs$y,
    imputed_missing = stats::approx(
      x = dens_imp$x, y = dens_imp$y,
      xout = dens_obs$x, rule = 2
    )$y
  )

  df_long <- df_dens %>%
    tidyr::pivot_longer(
      cols = c("observed", "imputed_missing"),
      names_to = "source",
      values_to = "density"
    ) %>%
    dplyr::mutate(
      source = factor(
        source,
        levels = c("observed", "imputed_missing"),
        labels = c("Observed (non-missing)", "Imputed (missing only)")
      )
    )

  print(
    ggplot2::ggplot(df_long, ggplot2::aes(x = x, y = density, color = source)) +
      ggplot2::geom_line(linewidth = 1.2) +
      ggplot2::labs(
        title = gsub("<.*?>", "", beautify_names(variable)),
        x = gsub("<.*?>", "", beautify_names(variable)),
        y = "Density",
        color = NULL,
        caption = caption_txt
      )
  )

  invisible(NULL)
}
```

```{r}
#| label: imputation-diagnostics-prep
#| code-fold: true
#| column: body-outset-right

# identify variables with missing values for the diagnostic figures
vars_with_missing <- missing_info$var[missing_info$missing_rate > 0]
```

```{r}
#| label: imputation-diagnostics-figs
#| fig-cap: "Observed versus Imputed Distributions (imputed values shown only at originally missing cells)"
#| fig-subcap: true
#| layout-ncol: 2
#| layout-type: grid
#| message: false
#| code-fold: show
#| column: body-outset-right

# generate diagnostic plots for each variable with missing values
vars_with_missing %>%
  purrr::walk(~ plot_imputation_diagnostic(
    orig_data = df_impute,
    imputed_data = imp_mf$ximp,
    variable = .x
  ))
```

## Compare Descriptive Statistics

```{r}
#| label: comparison-imputed-data
#| message: false
#| code-fold: show
#| column: body-outset-right

# compute descriptive statistics for observed and imputed numeric data
summ_obs <- df_impute %>%
  numeric_df() %>%
  describe(print_only = FALSE) %>%
  dplyr::rename(variable = var) %>%
  dplyr::select(-n)

summ_imp <- imp_mf$ximp %>%
  numeric_df() %>%
  describe(print_only = FALSE) %>%
  dplyr::rename(variable = var) %>%
  dplyr::select(-n)

comparison <- versus::compare(
  summ_obs, summ_imp,
  by = variable
) %>%
  versus::value_diffs_stacked(column = -dplyr::any_of("p")) %>%
  dplyr::mutate(diff = val_b - val_a, .after = "val_b") %>%
  dplyr::arrange(variable, column)

# format column names for display via literal lookup (regex replacement
# consumes backslashes in stringr/ICU replacement strings)
metric_labels <- c(
  "mean" = "$\\bar{x}$", "sd" = "$\\sigma$",
  "skew" = "$Sk$", "se" = "$SE$", "kurtosis" = "$K$",
  "norm" = "Normality"
)
comparison$column <- md_to_html(
  dplyr::coalesce(unname(metric_labels[comparison$column]), comparison$column)
)
```

```{r}
#| label: comparison-imputed-data-attr
#| code-fold: true
#| column: body-outset-right

# table metadata for comparison display
footnote_text <- glue(
  "$N$ = {nrow(df_impute)}; Descriptive statistics compared between observed ",
  "values (missing ignored) and the fully imputed dataset."
)
col_names <- c("Metric", "Observed", "Imputed", "Difference")
```

```{r}
#| label: tbl-comparison-imputed-data-tbl
#| tbl-cap: "Comparison of Observed and Fully Imputed Data (Numeric Variables)"
#| message: false
#| code-fold: show
#| column: body-outset-right

# prepare row grouping for table display
pack_row_indices <- row_grouping(comparison$variable)
names(pack_row_indices) %<>% beautify_names()

# display the comparison table with grouped rows
comparison %>%
  dplyr::select(-variable) %>%
  dplyr::rename(
    Metric = column,
    Observed = val_a,
    Imputed = val_b,
    Difference = diff
  ) %>%
  as_kable() %>%
  kableExtra::pack_rows(
    index = pack_row_indices,
    escape = FALSE, label_row_css = "text-align: left;"
  ) %>%
  footnote() %>%
  add_style()
```

## Integrate Final Dataset

- *in*: `imp_mf$ximp` (imputed partnered cases), `df_non_partnered` (held out), `passthrough_frame` (intact passthrough values)
- *out*: `imputed_data_combined.sav`
- *rule*: rejoin imputed partnered with held-out non-partnered; join the passthrough columns once from the pre-coercion frame; preserve structural NAs on partner variables for non-partnered cases

```{r}
#| label: imputed-data-combined
#| message: false
#| code-fold: show
#| column: body-outset-right

# verify complete cases remain unchanged post-imputation
complete_idx <- which(complete.cases(df_impute))
stopifnot(identical(df_impute[complete_idx, ], imp_mf$ximp[complete_idx, ]))

# add identifier and partner status back to imputed data
imp_mf$ximp$idnummer <- df_partnered$idnummer
imp_mf$ximp$partner <- 1

# combine imputed partnered data with non-partnered data, then join the
# passthrough columns once from the pre-coercion frame so class codes and
# character labour-force status arrive intact for both branches; copy_labels
# stays before the bind because it also restores the factor columns built
# for missForest to their labelled numeric types
imputed_data_combined <- copy_labels(
  source = df_with_missings,
  target = imp_mf$ximp
) %>%
  dplyr::bind_rows(df_non_partnered) %>%
  dplyr::left_join(passthrough_frame, by = "idnummer") %>%
  dplyr::select(idnummer, dplyr::everything())
```

```{r}
#| label: structural-score-reset
#| code-fold: show
#| column: body-outset-right

# step 5 sensitivity arm: reset imputed occupation scores where the
# occupation is structurally absent (not in labour force, no occupation
# reported); only occupation_unknown is item missingness. FALSE preserves
# the current estimand for the like-for-like comparison against the
# pre-reconciliation results.
reset_structural_scores <- FALSE
if (reset_structural_scores) {
  reset_map <- list(
    c(lf = "lf_status",         vars = "isei08_self",    s = "siops08"),
    c(lf = "lf_status_father",  vars = "isei08_father",  s = "siops08_father"),
    c(lf = "lf_status_mother",  vars = "isei08_mother",  s = "siops08_mother"),
    c(lf = "lf_status_partner", vars = "isei08_partner", s = "siops08_partner")
  )
  for (m in reset_map) {
    structural <- imputed_data_combined[[m["lf"]]] %in%
      c("not_in_labour_force", "no_occupation_reported")
    imputed_data_combined[[m["vars"]]][structural] <- NA
    imputed_data_combined[[m["s"]]][structural] <- NA
  }
}

# save the final combined dataset
haven::write_sav(
  imputed_data_combined,
  here::here("assets", "data", "saved", "imputed_data_combined.sav")
)
```

```{r}
#| label: imputation-summary
#| results: asis
#| code-fold: true
#| column: body-outset-right

# generate summary of imputation process
htmltools::HTML(glue(
  "<div class='callout-note'>
  <strong>Imputation Complete</strong><br><br>
  <strong>Partnered respondents imputed:</strong> {nrow(df_partnered)}<br>
  <strong>Non-partnered respondents (held out):</strong> {nrow(df_non_partnered)}<br>
  <strong>Combined dataset:</strong> {nrow(imputed_data_combined)} observations<br>
  <strong>Variables with imputed values:</strong> {length(vars_with_missing)}<br><br>
  <strong>Output:</strong> <code>imputed_data_combined.sav</code> for Step 6
  </div>"
))
```

## Descriptive Statistics

```{r}
#| label: descriptive-stats-col-names
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "$\\ $", "$n$", "$\\bar{x}$", "$\\text{Min.}$", "$\\text{Max.}$",
  "$SD$", "$\\text{Skewness}$", "$\\text{Kurtosis}$", "$SE$",
  "$\\text{Normality}^{\\dagger}$"
)

footnote_text <- glue(
  "Shapiro-Wilk Test$^{{\\dagger}}$;",
  "values are normality test statistics (Shapiro-Wilk $W$ for $n \\leq 5000$, Lilliefors $D$ otherwise); ",
  "smaller $W$ (larger $D$) suggests stronger departure from normality;",
  "'&#x2014;' indicates variables with $\\leq 4$ unique values",
  "(normality test not applicable)"
)
```

```{r}
#| label: tbl-descriptive-stats-imputed-data-combined
#| tbl-cap: "Descriptive Statistics for Imputed Dataset (Combined Partnered and Non-Partnered Subgroups)"
#| code-fold: show
#| column: page-right

imputed_data_combined %>%
  dplyr::select(-idnummer) %>%
  setNames(beautify_names(names(.))) %>%
  describe(
    normality = TRUE,
    print_only = FALSE
  ) %>%
  dplyr::rowwise() %>%
  dplyr::ungroup() %>%
  dplyr::select(-p) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                                               | Content                                                            |
| -------------------------------------------------- | ------------------------------------------------------------------ |
| `assets/data/saved/imputed_data_combined.sav`      | Imputed partnered cases rejoined with held-out non-partnered       |
| `assets/data/saved/imputed_results_missforest.rds` | `missForest` output with run metadata (seed, ntree, mtry, maxiter) |

**Next:** [Step 6 (Post-Imputation Finalization)](/analysis/data/post-imputation/index.qmd)
