---
title: "Unadjusted Regression Models"
subtitle: "Baseline MM-Estimation Fits Before Outlier Adjustment"
author: "Siard van den Bosch"
date: 2024-12-15
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 7};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Produce baseline MM-estimation fits for Models 0–11 on the full sample |
| **Inputs**     | `analysis_ready.sav` |
| **Outputs**    | `models.rds`, `unadjusted_data.rds`, `unadjusted_fit_{0..11}.rds`, `unadjusted_fits.rds` |
| **Method**     | MM-estimation robust regression (bisquare ψ); bootstrapping disabled |
| **Requires**   | [Step 6 (Post-Imputation Finalization)](/analysis/data/post-imputation/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- Energy adjustment: `dhd_kcal` enters as a covariate so that parental and partner SEP coefficients reflect dietary composition rather than caloric volume, consistent with the DHD15 scoring framework [@looman2017dutch]
- Lifestyle covariate: `smoking` is adjusted for as a co-occurring health behaviour; this follows standard practice in SES–diet research and is not attributed to a single source
- Own SES at baseline: entering `education_self` and `isei08_self` in Model 1 yields parental and partner SEP coefficients interpreted net of own attainment, matching the wording of H1 and H2
- Parallel composite and disaggregated families: Models 2–6 test mean-of-parents effects; Models 7–11 reproduce the same nesting with separate mother and father terms to expose parent-specific contributions
:::

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear global environment, garbage collect, and detach packages
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "robust_analysis", "standardize", "create_lmrob_control", "save_function",
  "named_list", "numeric_df", "set_globals", "import_pipes", "process_model",
  "fit_robust_model", "sci_to_tex", "prepare_model_matrices", "fit_models"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()

# global recompute switch
# TRUE re-runs the heavy steps; FALSE loads cached assets when they exist
recompute_heavy <- FALSE

# define analysis type
type <- "unadjusted"
options(width = 250)

# load analysis-ready dataset from step 6
analysis_data <- haven::read_sav(
  here::here("assets", "data", "saved", "analysis_ready.sav")
)

models_dir <- here::here("assets", "models")
if (!dir.exists(models_dir)) dir.create(models_dir, recursive = TRUE)

# consolidated output path; cached fits are reused when this file exists
consolidated_path <- file.path(models_dir, paste0(type, "_fits.rds"))

# refit when forced or when no cached consolidated fits exist
refit_models <- recompute_heavy || !file.exists(consolidated_path)

# when refitting, purge only this step's outputs (stale fits and specs);
# step 9 fits and step 10 caches stay intact
if (refit_models) {
  unlink(Sys.glob(file.path(models_dir, "unadjusted_*")))
  unlink(file.path(models_dir, "models.rds"))
}
```

## Verify Dataset from Step 6

Confirm expected variables are present before model fitting.

```{r}
#| label: verify-step6-data
#| results: asis
#| code-fold: show
#| column: body-outset-right

# define expected variables from step 6
ses_composites <- c("education_parental", "isei08_parental")

# partner SES (structural NA for non-partnered)
partner_ses <- c("education_partner", "isei08_partner")

# individual parent variables for disaggregated models
parental_disagg <- c(
  "education_father", "education_mother",
  "isei08_father", "isei08_mother"
)

# core regression variables
core_vars <- c(
  "dhd_index", "woman", "age", "partner",
  "education_self", "isei08_self"
)

# optional variables (may or may not be present)
optional_vars <- c("dhd_kcal", "smoking", "income_hh")

# check presence
ses_present <- ses_composites[ses_composites %in% names(analysis_data)]
ses_missing <- setdiff(ses_composites, ses_present)
partner_present <- partner_ses[partner_ses %in% names(analysis_data)]
partner_missing <- setdiff(partner_ses, partner_present)
parental_present <- parental_disagg[parental_disagg %in% names(analysis_data)]
parental_missing <- setdiff(parental_disagg, parental_present)
core_present <- core_vars[core_vars %in% names(analysis_data)]
core_missing <- setdiff(core_vars, core_present)
optional_present <- optional_vars[optional_vars %in% names(analysis_data)]

# sample sizes
n_total <- nrow(analysis_data)
n_partnered <- sum(analysis_data$partner == 1, na.rm = TRUE)
n_non_partnered <- sum(analysis_data$partner == 0, na.rm = TRUE)

# display verification results
verify_msg <- glue(
  "<div class='callout-note'>
<strong>Dataset Verification (Step 6)</strong><br><br>
<strong>Input:</strong> {n_total} observations<br>
<strong>Partnered:</strong> {n_partnered} ({round(100 * n_partnered / n_total, 1)}%)<br>
<strong>Non-partnered:</strong> {n_non_partnered} ({round(100 * n_non_partnered / n_total, 1)}%)<br><br>
<strong>Core variables:</strong> {if (length(core_missing) == 0) 'All present &#10003;' else paste('Missing:', paste(core_missing, collapse = ', '))}<br>
<strong>Parental composites:</strong> {if (length(ses_missing) == 0) 'All present &#10003;' else paste('Missing:', paste(ses_missing, collapse = ', '))}<br>
<strong>Parental disaggregated:</strong> {if (length(parental_missing) == 0) 'All present &#10003;' else paste('Missing:', paste(parental_missing, collapse = ', '))}<br>
<strong>Partner SES:</strong> {if (length(partner_missing) == 0) 'All present &#10003;' else paste('Missing:', paste(partner_missing, collapse = ', '))}<br>
<strong>Optional variables:</strong> {if (length(optional_present) > 0) paste(optional_present, collapse = ', ') else 'None'}
</div>"
)
cat(verify_msg)

# stop if critical variables are missing
if (length(core_missing) > 0) {
  stop("Missing core variables: ", paste(core_missing, collapse = ", "))
}
if (length(ses_missing) > 0) {
  warning("Missing parental composites: ", paste(ses_missing, collapse = ", "))
}
```

## Model Specifications

### Model 0: Null Model

```{r}
#| label: specify-model-0
#| message: false
#| code-fold: show
#| column: body-outset-right

model_0 <- dhd_index ~ 1
```

### Model 1: Adult Covariates and Own SES (Adjustment Set)

```{r}
#| label: specify-model-1
#| message: false
#| code-fold: show
#| column: body-outset-right

# base model: demographics + lifestyle + own SES
base_formula <- dhd_index ~ woman + age + partner +
  education_self + isei08_self

# nutritional-epidemiology covariate: total energy intake
if ("dhd_kcal" %in% names(analysis_data)) {
  base_formula <- update(base_formula, . ~ . + dhd_kcal)
}
# lifestyle covariate: smoking status
if ("smoking" %in% names(analysis_data)) {
  base_formula <- update(base_formula, . ~ . + smoking)
}
# household economics (optional, conditional on availability)
if ("income_hh" %in% names(analysis_data)) {
  base_formula <- update(base_formula, . ~ . + income_hh)
}

model_1 <- base_formula
```

## Composite SES Models (Models 2–6)

These models use parental composites that average mother and father within education and occupation domains.

### Model 2: Add Parental SES Main Effects (H1, RQ1)

```{r}
#| label: specify-model-2
#| message: false
#| code-fold: show
#| column: body-outset-right

model_2 <- update(model_1, . ~ . +
  education_parental +
  isei08_parental)
```

### Model 3: Add Partner SES Main Effects (H2, RQ2)

```{r}
#| label: specify-model-3
#| message: false
#| code-fold: show
#| column: body-outset-right

model_3 <- update(model_2, . ~ . +
  education_partner:partner +
  isei08_partner:partner)
```

### Model 4: Gender Moderation of Parental and Partner SES (H3, H4, RQ3)

```{r}
#| label: specify-model-4
#| message: false
#| code-fold: show
#| column: body-outset-right

model_4 <- update(model_3, . ~ . +
  woman:education_parental +
  woman:isei08_parental +
  woman:education_partner:partner +
  woman:isei08_partner:partner)
```

### Model 5: Parental × Partner SES Synergy (H5, RQ4)

```{r}
#| label: specify-model-5
#| message: false
#| code-fold: show
#| column: body-outset-right

model_5 <- update(model_4, . ~ . +
  education_parental:education_partner:partner +
  isei08_parental:isei08_partner:partner)
```

### Model 6: Gender Moderation of Synergy (H6, RQ4)

```{r}
#| label: specify-model-6
#| message: false
#| code-fold: show
#| column: body-outset-right

model_6 <- update(model_5, . ~ . +
  woman:education_parental:education_partner:partner +
  woman:isei08_parental:isei08_partner:partner)
```

## Disaggregated SES Models (Models 7–11)

These models use separate variables for mother, father, and partner education and occupation.

### Model 7: Mother and Father SES Main Effects (H1, RQ1)

```{r}
#| label: specify-model-7
#| message: false
#| code-fold: show
#| column: body-outset-right

model_7 <- update(model_1, . ~ . +
  education_mother +
  education_father +
  isei08_mother +
  isei08_father)
```

### Model 8: Add Partner SES Main Effects (H2, RQ2)

```{r}
#| label: specify-model-8
#| message: false
#| code-fold: show
#| column: body-outset-right

model_8 <- update(model_7, . ~ . +
  education_partner:partner +
  isei08_partner:partner)
```

### Model 9: Gender × Mother/Father/Partner SES Interactions (H3, H4, RQ3)

```{r}
#| label: specify-model-9
#| message: false
#| code-fold: show
#| column: body-outset-right

model_9 <- update(model_8, . ~ . +
  woman:education_mother +
  woman:education_father +
  woman:isei08_mother +
  woman:isei08_father +
  woman:education_partner:partner +
  woman:isei08_partner:partner)
```

### Model 10: Mother × Partner and Father × Partner Synergy (H5, RQ4)

```{r}
#| label: specify-model-10
#| message: false
#| code-fold: show
#| column: body-outset-right

model_10 <- update(model_9, . ~ . +
  education_mother:education_partner:partner +
  education_father:education_partner:partner +
  isei08_mother:isei08_partner:partner +
  isei08_father:isei08_partner:partner)
```

### Model 11: Gender × Parent-Specific Synergy (H6, RQ4)

```{r}
#| label: specify-model-11
#| message: false
#| code-fold: show
#| column: body-outset-right

model_11 <- update(model_10, . ~ . +
  woman:education_mother:education_partner:partner +
  woman:education_father:education_partner:partner +
  woman:isei08_mother:isei08_partner:partner +
  woman:isei08_father:isei08_partner:partner)
```

## Prepare Data for Model Fitting

```{r}
#| label: prepare-data
#| message: false
#| code-fold: show
#| column: body-outset-right

# collect model objects in numeric order
model_names <- ls(pattern = "^model_[0-9]+$")
model_names <- model_names[order(as.integer(sub("model_", "", model_names)))]
models <- lapply(model_names, function(nm) get(nm, envir = .GlobalEnv))
names(models) <- model_names

# save model specifications
saveRDS(models, file = file.path(models_dir, "models.rds"))

# prepare analysis data
unadjusted_data <- numeric_df(analysis_data) %>%
  dplyr::mutate(
    # ensure partner is numeric binary
    partner = as.numeric(partner == 1),

    # condition partner SES on having a partner (set to 0 for non-partnered)
    education_partner = dplyr::if_else(partner == 0, 0, education_partner),
    isei08_partner = dplyr::if_else(partner == 0, 0, isei08_partner)
  )

# identify variables used in models
vars_in_models <- unique(unlist(lapply(models, function(m) {
  as.character(attr(stats::terms(m), "variables"))[-1]
})))

# define analysis variable set
analysis_vars <- c(
  "dhd_index",
  # parental composites
  "education_parental", "isei08_parental",
  # disaggregated parental
  "education_father", "education_mother",
  "isei08_father", "isei08_mother",
  # partner
  "education_partner", "isei08_partner",
  # own SES
  "education_self", "isei08_self",
  # demographics
  "partner", "woman", "age"
)

# add optional covariates if present
if ("dhd_kcal" %in% names(unadjusted_data)) {
  analysis_vars <- c(analysis_vars, "dhd_kcal")
}
if ("smoking" %in% names(unadjusted_data)) {
  analysis_vars <- c(analysis_vars, "smoking")
}
if ("income_hh" %in% names(unadjusted_data)) {
  analysis_vars <- c(analysis_vars, "income_hh")
}

vars_keep <- unique(c(vars_in_models, analysis_vars))
vars_keep <- intersect(vars_keep, names(unadjusted_data))

unadjusted_data <- unadjusted_data[vars_keep]

# save preprocessed dataset
saveRDS(unadjusted_data, file = here::here("assets", "data", "saved", "unadjusted_data.rds"))

# disable bootstrapping for baseline fits
n_bootstraps <- 0
```

```{r}
#| label: data-summary
#| results: asis
#| code-fold: true
#| column: body-outset-right

data_msg <- glue(
  "<div class='callout-note'>
<strong>Analysis Dataset Prepared</strong><br><br>
<strong>Observations:</strong> {nrow(unadjusted_data)}<br>
<strong>Variables retained:</strong> {ncol(unadjusted_data)}<br>
<strong>Models specified:</strong> {length(models)} (Models 0-{length(models) - 1})<br>
<strong>Bootstrap iterations:</strong> {n_bootstraps} (disabled for baseline)<br><br>
<strong>Parental composites:</strong> <code>education_parental</code>, <code>isei08_parental</code><br>
<strong>Disaggregated parental:</strong> <code>education_mother</code>, <code>education_father</code>, <code>isei08_mother</code>, <code>isei08_father</code><br>
<strong>Partner SES:</strong> <code>education_partner</code>, <code>isei08_partner</code>
</div>"
)
cat(data_msg)
```

## Fit Regression Models

### Fit Model 0: Null Model (Baseline)

```{r}
#| label: fit-model-0
#| eval: !expr refit_models
#| code-fold: show
#| column: body-outset-right

fit_0 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_0,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_0, file = file.path(models_dir, glue("{type}_fit_0.rds")))
```

### Fit Model 1: Adult Covariates and Own SES (Baseline)

```{r}
#| label: fit-model-1-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_1 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_1,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_1, file = file.path(models_dir, glue("{type}_fit_1.rds")))
```

### Fit Model 2: Parental SES Main Effects, Composite (H1, RQ1)

```{r}
#| label: fit-model-2-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_2 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_2,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_2, file = file.path(models_dir, glue("{type}_fit_2.rds")))
```

### Fit Model 3: Partner SES Main Effects, Composite (H2, RQ2)

```{r}
#| label: fit-model-3-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_3 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_3,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_3, file = file.path(models_dir, glue("{type}_fit_3.rds")))
```

### Fit Model 4: Gender Moderation of SES, Composite (H3, H4, RQ3)

```{r}
#| label: fit-model-4-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_4 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_4,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_4, file = file.path(models_dir, glue("{type}_fit_4.rds")))
```

### Fit Model 5: Parental × Partner Synergy, Composite (H5, RQ4)

```{r}
#| label: fit-model-5-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_5 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_5,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_5, file = file.path(models_dir, glue("{type}_fit_5.rds")))
```

### Fit Model 6: Gender-Moderated Synergy, Composite (H6, RQ4)

```{r}
#| label: fit-model-6-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_6 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_6,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_6, file = file.path(models_dir, glue("{type}_fit_6.rds")))
```

### Fit Model 7: Mother/Father SES Main Effects, Disaggregated (H1, RQ1)

```{r}
#| label: fit-model-7-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_7 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_7,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_7, file = file.path(models_dir, glue("{type}_fit_7.rds")))
```

### Fit Model 8: Partner SES Main Effects, Disaggregated (H2, RQ2)

```{r}
#| label: fit-model-8-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_8 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_8,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_8, file = file.path(models_dir, glue("{type}_fit_8.rds")))
```

### Fit Model 9: Gender × Mother/Father/Partner SES, Disaggregated (H3, H4, RQ3)

```{r}
#| label: fit-model-9-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_9 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_9,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_9, file = file.path(models_dir, glue("{type}_fit_9.rds")))
```

### Fit Model 10: Parent × Partner Synergy, Disaggregated (H5, RQ4)

```{r}
#| label: fit-model-10-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_10 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_10,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_10, file = file.path(models_dir, glue("{type}_fit_10.rds")))
```

### Fit Model 11: Gender × Parent-Specific Synergy, Disaggregated (H6, RQ4)

```{r}
#| label: fit-model-11-hidden
#| eval: !expr refit_models
#| message: false
#| code-fold: show
#| column: page-right

fit_11 <- robust_analysis(
  data = get(glue("{type}_data")),
  form = model_11,
  n_boots = n_bootstraps,
  pct = 0.95,
  alpha = 0.05,
  delta = 0.01,
  ctrl = NULL,
  seed = 625,
  psi = "bisquare",
  std_data = TRUE
)

saveRDS(fit_11, file = file.path(models_dir, glue("{type}_fit_11.rds")))
```

## Save All Model Fits

```{r}
#| label: save-all-fits
#| message: false
#| code-fold: show
#| column: body-outset-right

# rebuild and persist the consolidated list only when (re)fitting
if (refit_models) {
  all_fits <- named_list(
    fit_0, fit_1, fit_2, fit_3, fit_4, fit_5, fit_6,
    fit_7, fit_8, fit_9, fit_10, fit_11
  )

  saveRDS(all_fits, consolidated_path)
} else {
  message("cached fits found; skipping model fitting (set recompute_heavy <- TRUE to refit)")
}
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                                       | Content                                               |
| ------------------------------------------ | ----------------------------------------------------- |
| `assets/models/models.rds`                 | Formula specifications for Models 0–11                |
| `assets/data/saved/unadjusted_data.rds`    | Preprocessed analysis dataset for the unadjusted fits |
| `assets/models/unadjusted_fit_{0..11}.rds` | One MM-estimation fit object per model specification  |
| `assets/models/unadjusted_fits.rds`        | Consolidated named list of all 12 fitted models       |

**Next:** [Step 8 (Influence Diagnostics)](/analysis/models/assumptions/index.qmd)
