---
title: "Detecting & Removing Outliers"
subtitle: "Univariate and Multivariate Approaches for Data Refinement"
author: "Siard van den Bosch"
date: 2024-12-17
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 4}
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Remove univariate and multivariate outliers from the analysis dataset |
| **Inputs**     | `data_for_imputation.sav` |
| **Outputs**    | `data_without_outliers.sav`, `outlier_cases_removed.csv` |
| **Method**     | IQR/MAD univariate detection; LOF multivariate detection |
| **Requires**   | [Step 3 (Data Cleaning)](/analysis/data/cleaning/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- IQR for `dhd_kcal`: classical robust rule for energy intake distributions [@tukey1977]
- MAD for `dhd_index`: robust to non-normality, 50% breakdown point, threshold 2.5 approximates 99% normal coverage [@leys2013]
- LOF for multivariate detection: density-based, no distributional assumption [@breunig2000]
- Tukey adjusted-boxplot cutoff: skew-corrected via medcouple, default alpha = 0.05 controls Type-I error [@hubert2008]
- Partner variables excluded from complete-case filter: structural missingness for non-partnered respondents [@vanbuuren2018]
- SES composites excluded from outlier detection: extreme socioeconomic values represent legitimate population variation, not measurement error [@aguinis2013]
- Reconciled score and status columns (`siops08*`, `egp11*`, `lf_status*`) excluded from detection: preserves the complete-case pool and LOF feature set of the pre-reconciliation analysis
:::

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear global env and force full garbage collection
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))

# detach non-base packages to avoid namespace interference
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions for data handling and visualization
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "remove_invariant", "md_to_html", "set_globals", "add_style",
  "format_numeric", "as_kable", "format_p_value", "import_pipes",
  "footnote", "beautify_names", "numeric_df", "describe"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()

# load dataset prepared for imputation (from step 3)
main <- haven::read_sav(
  here::here("assets", "data", "saved", "data_for_imputation.sav"),
  user_na = FALSE
)
```

```{r}
#| label: define-constants
#| code-fold: true
#| column: body-outset-right

# partner-specific variables with structural missingness (NA by design for non-partnered)
partner_vars <- c("education_partner", "isei08_partner")

# parental SES composites from step 3 (dimension_referent convention)
ses_composites <- c("education_parental", "isei08_parental")

# target variables for univariate outlier detection
univariate_targets <- list(
  dhd_kcal = "iqr",
  dhd_index = "mad"
)
```

## Verify Dataset from Step 3

```{r}
#| label: verify-step3-variables
#| results: asis
#| code-fold: true
#| column: body-outset-right

# identify which structural and SES variables are present
structural_vars_present <- partner_vars[
  partner_vars %in% names(main)
]
ses_present <- ses_composites[ses_composites %in% names(main)]

# count non-partnered respondents
n_non_partnered <- sum(main$partner == 0, na.rm = TRUE)

# generate informative message about variable handling
if (length(ses_present) > 0 || length(structural_vars_present) > 0) {
  ses_msg <- glue(
    "<div class='callout-note'>
    <strong>SES Variables from Step 3</strong><br><br>
    <strong>Parental composites:</strong> ",
    if (length(ses_present) > 0) {
      glue("<code>{paste(ses_present, collapse = '</code>, <code>')}</code>")
    } else {
      "None detected"
    },
    "<br>
    <strong>Partner indicators:</strong> ",
    if (length(structural_vars_present) > 0) {
      glue("<code>{paste(structural_vars_present, collapse = '</code>, <code>')}</code>")
    } else {
      "None detected"
    },
    "<br><br>
    <strong>Structural missingness:</strong> Partner-specific variables ",
    "(<code>education_partner</code>, <code>isei08_partner</code>) contain NA values ",
    "by design for non-partnered respondents (n = {n_non_partnered}). ",
    "These cases represent a legitimate subpopulation and are <em>not</em> excluded ",
    "from outlier detection procedures. Partner variables are excluded from ",
    "complete-case filtering and multivariate LOF detection to prevent inappropriate ",
    "sample reduction.
    </div>"
  )
  cat(ses_msg)
}
```

## Clean Dataset for Outlier Detection

```{r}
#| label: data-preparation
#| code-fold: show
#| column: body-outset-right

# remove invariant columns and replace non-finite values with NA
clean_data <- main %>%
  # reconciled score, class, and status columns stay out of outlier
  # detection entirely: siops08 and egp11 would shrink the complete-case
  # pool through their missingness and duplicate the occupational signal
  # already carried by isei08; lf_status is character
  dplyr::select(-dplyr::matches("^siops08|^egp11|^lf_status")) %>%
  remove_invariant() %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.numeric),
      ~ ifelse(is.finite(.x), .x, NA)
    )
  ) %>%
  # select numeric columns with non-zero variance
  dplyr::select(dplyr::where(~ is.numeric(.x) && sd(.x, na.rm = TRUE) != 0))

# identify partner-specific variables present in cleaned data
partner_vars_in_data <- intersect(partner_vars, names(clean_data))

# retain complete cases EXCLUDING partner-specific variables
# this preserves non-partnered respondents with legitimate NA values
clean_data <- if (length(partner_vars_in_data) > 0) {
  clean_data %>%
    dplyr::filter(complete.cases(
      dplyr::select(., -dplyr::any_of(partner_vars_in_data))
    ))
} else {
  clean_data %>%
    dplyr::filter(complete.cases(.))
}

# record sample sizes
n_original <- nrow(main)
n_clean <- nrow(clean_data)
n_excluded_cleaning <- n_original - n_clean
```

```{r}
#| label: cleaning-summary
#| results: asis
#| code-fold: true
#| column: body-outset-right

cat(glue(
  "<p><strong>Data Cleaning Summary:</strong> ",
  "Of {n_original} observations in the input dataset, ",
  "{n_clean} ({round(100 * n_clean / n_original, 1)}%) were retained for ",
  "outlier detection after removing cases with missing values on substantive ",
  "variables. Partner-specific variables were excluded from this filter to ",
  "preserve non-partnered respondents.</p>"
))
```

## Detect Univariate Outliers

```{r}
#| label: univariate-outlier-function
#| code-fold: show
#| column: body-outset-right

# detect univariate outliers using IQR, z-score, or MAD method
univariate_outliers <- function(data, var, meth = "iqr", thresh = NULL) {
  stopifnot(meth %in% c("iqr", "zscore", "mad"))

  # set method-specific default threshold
  if (is.null(thresh)) {
    thresh <- if (meth == "iqr") 1.5 else 2.5
  }

  # extract finite, non-NA values
  x <- data[[var]][!is.na(data[[var]]) & is.finite(data[[var]])]

  # calculate thresholds based on method
  lims <- switch(meth,
    "iqr" = {
      q <- quantile(x, c(0.25, 0.75), names = FALSE)
      iqr <- diff(q)
      q + c(-thresh, thresh) * iqr
    },
    "zscore" = mean(x) + c(-1, 1) * thresh * sd(x),
    "mad" = {
      med <- median(x)
      med + c(-1, 1) * thresh * mad(x)
    }
  )

  # identify outlier indices
  outs_idx <- which(x < lims[1] | x > lims[2])

  # compute mean and standard error
  stats_val <- c(mean = mean(x), se = sd(x) / sqrt(length(x)))

  # calculate 95% confidence interval
  ci <- stats_val["mean"] + c(-1.96, 1.96) * stats_val["se"]

  # return NULL if non-finite values present
  if (!all(is.finite(c(stats_val, ci)))) {
    return(NULL)
  }

  list(
    outliers = x[outs_idx],
    ids = data$idnummer[!is.na(data[[var]]) & is.finite(data[[var]])][outs_idx],
    method = meth,
    lower_threshold = lims[1],
    upper_threshold = lims[2],
    mean = stats_val["mean"],
    ci_lower = ci[1],
    ci_upper = ci[2]
  )
}
```

```{r}
#| label: detect-outliers-function
#| code-fold: show
#| column: body-outset-right

# detect outliers across multiple variables with specified methods
detect_outliers <- function(data, vars, meths) {
  data$outlier <- NA_character_

  # apply univariate detection to each variable
  res <- Map(function(v, m) univariate_outliers(data, v, m), vars, meths)

  # flag outliers in data
  for (i in seq_along(vars)) {
    if (!is.null(res[[i]])) {
      ids <- res[[i]]$ids
      label <- glue("Univariate ({beautify_names(vars[i])})")
      data$outlier[data$idnummer %in% ids] <- label
    }
  }

  # extract summary metrics
  metrics <- c(
    "method", "mean", "ci_lower", "ci_upper",
    "upper_threshold", "lower_threshold"
  )

  sum_data <- sapply(metrics, function(m) {
    v <- sapply(res, function(r) if (is.null(r)) NA else r[[m]])
    if (m == "method") toupper(as.character(v)) else round(as.numeric(v), 2)
  })

  # count detections
  detection_counts <- sapply(res, function(r) {
    if (is.null(r)) 0 else length(r$outliers)
  })

  # count unique cases (some may be flagged on multiple variables)
  all_outlier_ids <- unique(unlist(lapply(res, function(r) {
    if (!is.null(r)) r$ids else NULL
  })))
  unique_cases <- length(all_outlier_ids)

  list(
    summary = list(
      variable = c(vars, "Total detections", "Unique cases"),
      count = c(detection_counts, sum(detection_counts), unique_cases),
      tbl = rbind(sum_data, NA, NA)
    ),
    results = res,
    updated_data = data
  )
}
```

### Boxplot Visualization

```{r}
#| label: create-boxplot-function
#| code-fold: show
#| column: body-outset-right

# create boxplot with outliers highlighted and confidence intervals
create_boxplot <- function(data, variable, ids, mean_val, ci_lower, ci_upper) {
  # define colors
  ci_color <- viridis::plasma(20)[4]
  outlier_color <- viridis::plasma(20)[12]
  data_color <- viridis::viridis(10)[4]

  mean_data <- data.frame(
    x = factor(1), y = mean_val,
    ymin = ci_lower, ymax = ci_upper
  )

  # set jitter position with seed for reproducibility
  pos_jitter <- if ("seed" %in% names(formals(ggplot2::position_jitter))) {
    ggplot2::position_jitter(width = 0.3, height = 0.25, seed = 42)
  } else {
    ggplot2::position_jitter(width = 0.3, height = 0.25)
  }

  ggplot2::ggplot(data, ggplot2::aes(x = factor(1), y = .data[[variable]])) +
    ggplot2::geom_boxplot(fill = "lightgrey", alpha = 0.5, outlier.shape = NA) +
    ggplot2::geom_point(
      data = dplyr::filter(data, idnummer %in% ids),
      ggplot2::aes(x = factor(1), y = .data[[variable]], color = "Outliers"),
      size = 1
    ) +
    ggplot2::geom_jitter(
      data = dplyr::filter(data, !idnummer %in% ids),
      alpha = 0.25,
      color = data_color,
      position = pos_jitter,
      show.legend = FALSE
    ) +
    ggplot2::geom_errorbar(
      data = mean_data,
      ggplot2::aes(x = x, ymin = ymin, ymax = ymax, color = "Mean (95% CI)"),
      width = 0.12, linewidth = 1.2,
      inherit.aes = FALSE
    ) +
    ggplot2::geom_point(
      data = mean_data,
      ggplot2::aes(x = x, y = y, color = "Mean (95% CI)"),
      size = 2, shape = 18,
      inherit.aes = FALSE
    ) +
    ggplot2::labs(title = NULL, x = NULL, y = beautify_names(variable), color = NULL) +
    ggplot2::scale_color_manual(
      values = c("Outliers" = outlier_color, "Mean (95% CI)" = ci_color)
    ) +
    garamond_theme
}
```

## Detect Multivariate Outliers

```{r}
#| label: multivariate-outliers-function
#| code-fold: show
#| column: body-outset-right

# detect multivariate outliers using Local Outlier Factor (LOF)
# k is fixed at 20 to avoid the default combined-k behavior of bigutilsr::LOF,
# which aggregates across seq_k = c(4, 10, 30) and complicates interpretation
multivariate_outliers <- function(df, excl = NULL,
                                  id_var = "idnummer", k = 20L) {
  stopifnot(id_var %in% names(df))

  # add row numbers for tracking
  df_with_id <- df %>%
    dplyr::mutate(orig_row = dplyr::row_number())

  # prepare data: remove invariant columns, replace non-finite with NA
  df_clean <- df_with_id %>%
    remove_invariant() %>%
    dplyr::mutate(dplyr::across(
      dplyr::where(is.numeric),
      ~ ifelse(is.finite(.x), .x, NA)
    )) %>%
    dplyr::select(dplyr::where(~ is.numeric(.x) && sd(.x, na.rm = TRUE) != 0)) %>%
    dplyr::filter(complete.cases(.))

  df_clean <- numeric_df(df_clean)

  # keep id mapping separate from LOF feature set
  id_map <- df_clean %>%
    dplyr::select(dplyr::all_of(c(id_var, "orig_row")))

  # exclude identifiers and specified variables from LOF features
  excl_all <- unique(c(excl, id_var, "orig_row"))
  lof_df <- df_clean %>%
    dplyr::select(-dplyr::any_of(excl_all))

  if (ncol(lof_df) == 0) {
    stop("No numeric variables left for LOF after exclusions.")
  }

  # compute log-LOF with single, explicit k (Breunig et al., 2000).
  # with seq_k of length 1, `combine` is irrelevant because only one score
  # exists per observation; we keep the default `combine = max` for
  # compatibility with bigutilsr's API.
  llof <- bigutilsr::LOF(
    U = as.matrix(lof_df),
    seq_k = k,
    log = TRUE
  )

  # Tukey's adjusted boxplot rule (skew-corrected medcouple correction) is
  # the idiomatic bigutilsr cutoff for LOF output (see `?tukey_mc_up`).
  # Default alpha = 0.05 controls Type-I error after multiple-testing
  # correction; this is the canonical usage in the bigutilsr manual and is
  # retained here rather than tightening to the legacy thresh = 0.001,
  # which corresponds to a percentile rule, not a Type-I error.
  lof_cutoff <- bigutilsr::tukey_mc_up(llof)
  lof_outs_idxs <- which(llof > lof_cutoff)

  # map back to original ids
  out_rows <- id_map$orig_row[lof_outs_idxs]
  out_ids <- id_map[[id_var]][lof_outs_idxs]

  # filter out outliers by id
  df_without_outs <- df %>%
    dplyr::filter(!.data[[id_var]] %in% out_ids)

  list(
    df = df_without_outs,
    llof = llof,
    cutoff = lof_cutoff,
    k = k,
    id = out_ids,
    row = out_rows,
    n_features = ncol(lof_df),
    features_used = names(lof_df)
  )
}
```

## Remove Detected Outliers

### Univariate Outlier Removal

```{r}
#| label: univariate-analysis
#| code-fold: show
#| column: body-outset-right

# detect univariate outliers
univariate_analysis <- clean_data %>%
  detect_outliers(
    vars = names(univariate_targets),
    meths = unlist(univariate_targets)
  )

# extract components
uni_summary <- univariate_analysis$summary
uni_results <- univariate_analysis$results
uni_dropped <- univariate_analysis$updated_data %>%
  dplyr::filter(!is.na(outlier))
uni_kept <- univariate_analysis$updated_data %>%
  dplyr::filter(is.na(outlier))
```

### Multivariate Outlier Removal

```{r}
#| label: multi-outliers
#| results: asis
#| code-fold: show
#| column: body-outset-right

# identify columns to exclude from LOF based on variance/uniqueness
cols_low_variance <- names(uni_kept)[sapply(uni_kept, function(x) {
  is.numeric(x) && (length(unique(x)) <= 2 ||
    length(unique(round(x))) > 0.95 * nrow(uni_kept))
})]

# partner-specific variables to exclude (structural missingness)
# uses canonical dimension_referent naming convention
partner_vars_exclude <- c("education_partner", "isei08_partner", "partner")

# combine all exclusions
cols_to_exclude <- unique(c(
  cols_low_variance,
  partner_vars_exclude[partner_vars_exclude %in% names(uni_kept)]
))

# document excluded columns
if (length(cols_to_exclude) > 0) {
  excluded_low_var <- cols_low_variance[cols_low_variance %in% names(uni_kept)]
  excluded_partner <- partner_vars_exclude[partner_vars_exclude %in% names(uni_kept)]

  exclusion_parts <- c()
  if (length(excluded_low_var) > 0) {
    exclusion_parts <- c(exclusion_parts, glue(
      "low variance/near-constant/near-unique: {toString(dQuote(excluded_low_var, q = FALSE))}"
    ))
  }
  if (length(excluded_partner) > 0) {
    exclusion_parts <- c(exclusion_parts, glue(
      "structural missingness (partner-specific): {toString(dQuote(excluded_partner, q = FALSE))}"
    ))
  }

  excluded_msg <- glue(
    "<p><strong>Note.</strong> Columns excluded from multivariate outlier detection:<br>",
    "{paste(paste0('&bull; ', exclusion_parts), collapse = '<br>')}</p>"
  )
} else {
  excluded_msg <- "<p><strong>Note.</strong> No columns were excluded from multivariate outlier detection.</p>"
}
cat(excluded_msg)

# detect and remove multivariate outliers
outliers_removed <- multivariate_outliers(
  df = uni_kept,
  excl = cols_to_exclude
)

lof_scores <- outliers_removed$llof
lof_cutoff <- outliers_removed$cutoff
ids_removed <- outliers_removed$id
n_lof_features <- outliers_removed$n_features

# collect multivariate outliers from original dataset
multi_dropped <- main %>%
  dplyr::filter(idnummer %in% ids_removed) %>%
  dplyr::mutate(outlier = "Multivariate")

# combine univariate and multivariate outliers
cases_with_outliers <- dplyr::bind_rows(uni_dropped, multi_dropped) %>%
  dplyr::distinct(idnummer, .keep_all = TRUE)

# calculate removal counts
rows_to_remove_uni <- nrow(uni_dropped)
rows_to_remove_multi <- nrow(multi_dropped)
expected_removed <- rows_to_remove_uni + rows_to_remove_multi
actual_removed <- nrow(cases_with_outliers)

# check for overlap
overlap_count <- expected_removed - actual_removed
if (overlap_count > 0) {
  cat(glue(
    "<p><strong>Note.</strong> {overlap_count} case(s) were flagged by both ",
    "univariate and multivariate methods and are counted once in the total.</p>"
  ))
}

cat(glue(
  "<p><strong>LOF detection (k = {outliers_removed$k}):</strong> ",
  "{length(ids_removed)} multivariate outlier(s) identified via Tukey's ",
  "adjusted boxplot rule on log-LOF scores (bigutilsr::tukey_mc_up, ",
  "default alpha = 0.05 after multiple-testing correction; ",
  "log-LOF cutoff = {round(lof_cutoff, 3)}; max observed log-LOF = ",
  "{round(max(lof_scores, na.rm = TRUE), 3)}).</p>"
))
```

```{r}
#| label: lof-features-summary
#| results: asis
#| code-fold: true
#| column: body-outset-right

cat(glue(
  "<p><strong>LOF Detection Summary:</strong> Multivariate outlier detection ",
  "was performed using {n_lof_features} numeric features after exclusions.</p>"
))
```

## Summarize Outlier Detection Results

### Univariate Outlier Summary

```{r}
#| label: univariate-definitions-attrs
#| code-fold: true
#| column: body-outset-right

# table metadata for univariate outlier summary
univariate_tbl_meta <- list(
  col_names = c(
    "$\\ $", "$n$", "$\\text{Method}$", "$\\bar{x}$",
    "$CI_{\\text{lower}}$", "$CI_{\\text{upper}}$",
    "$T^{+}$", "$T^{-}$"
  ),
  footnote_text = glue(
    "Outliers were detected using the Interquartile Range (IQR) ",
    "method for DHD kcal, which identifies extreme deviations based ",
    "on quartiles, and the Median Absolute Deviation (MAD) method ",
    "for DHD index, emphasizing robust measures of spread. ",
    "Threshold values $T^+$ (upper) and $T^-$ (lower) are ",
    "calculated from these methods to classify outliers. ",
    "'Total detections' sums all flagged values; 'Unique cases' ",
    "counts distinct respondents (some may be flagged on multiple variables)."
  )
)

# assign to parent frame for use by as_kable and footnote
col_names <- univariate_tbl_meta$col_names
footnote_text <- univariate_tbl_meta$footnote_text
```

```{r}
#| label: tbl-univariate-outliers
#| tbl-cap: "Number of Univariate Outliers per Variable"
#| code-fold: show
#| column: body-outset-right

data.frame(
  var = uni_summary$variable,
  count = uni_summary$count
) %>%
  cbind(uni_summary$tbl) %>%
  dplyr::mutate(var = beautify_names(var)) %>%
  format_numeric() %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Univariate Outliers Visualization

```{r}
#| label: fig-univariate-outliers-kcal
#| fig-cap: "DHD KCAL Outliers (IQR)"
#| code-fold: show
#| column: page-right

if (!is.null(uni_results[["dhd_kcal"]])) {
  print(create_boxplot(
    data = clean_data,
    variable = "dhd_kcal",
    ids = uni_results[["dhd_kcal"]]$ids,
    mean_val = uni_results[["dhd_kcal"]]$mean,
    ci_lower = uni_results[["dhd_kcal"]]$ci_lower,
    ci_upper = uni_results[["dhd_kcal"]]$ci_upper
  ))
}
```

```{r}
#| label: fig-univariate-outliers-index
#| fig-cap: "DHD Index Outliers (MAD)"
#| code-fold: show
#| column: page-right

if (!is.null(uni_results[["dhd_index"]])) {
  print(create_boxplot(
    data = clean_data,
    variable = "dhd_index",
    ids = uni_results[["dhd_index"]]$ids,
    mean_val = uni_results[["dhd_index"]]$mean,
    ci_lower = uni_results[["dhd_index"]]$ci_lower,
    ci_upper = uni_results[["dhd_index"]]$ci_upper
  ))
}
```

### Combined Outlier Cases

```{r}
#| label: combined-definitions-attrs
#| code-fold: true
#| column: body-outset-right

footnote_text <- glue(
  "$N_{{\\text{{total}}}} = {actual_removed}$; ",
  "$N_{{\\text{{univariate}}}} = {rows_to_remove_uni}$; ",
  "$N_{{\\text{{multivariate}}}} = {rows_to_remove_multi}$.<br>",
  "Cases were excluded using two independent outlier detection approaches:<br>",
  "1) Univariate outliers: Identified via boxplot analysis with variable-specific ",
  "thresholds (DHD-kcal: $Q_1 - 1.5 \\times IQR$ and $Q_3 + 1.5 \\times IQR$; ",
  "DHD-index: median $\\pm 2.5 \\times MAD$).<br>",
  "2) Multivariate outliers: Detected using Local Outlier Factor (LOF) method ",
  "(Tukey's adjusted boxplot cutoff on log-LOF scores) on {n_lof_features} numeric features ",
  "with non-zero variance and complete cases. ",
  "Partner-specific variables were excluded due to structural missingness. ",
  "Reconciled score and status columns were excluded from detection to keep ",
  "the pool and feature set comparable with the pre-reconciliation analysis."
)
```

```{r}
#| label: tbl-cases-with-outliers
#| tbl-cap: "Cases with Univariate and Multivariate Outliers Excluded"
#| code-fold: show
#| column: page-right

cases_with_outliers %>%
  format_numeric(digits = 2) %>%
  dplyr::mutate(outlier = beautify_names(outlier)) %>%
  dplyr::mutate(
    outlier = gsub("UniVariableiate", "Univariate", outlier, fixed = TRUE),
    outlier = gsub("MultiVariableiate", "Multivariate", outlier, fixed = TRUE)
  ) %>%
  as_kable(beautify_names(names(.))) %>%
  footnote() %>%
  add_style()
```

## Finalize Dataset for Analysis

```{r}
#| label: final-data-preparation
#| code-fold: show
#| column: body-outset-right

# collect all rows to remove
all_rows_to_remove <- cases_with_outliers$idnummer

# filter out outliers from main dataset
data_without_outliers <- main %>%
  dplyr::filter(!idnummer %in% all_rows_to_remove)

# verify consistency
rows_match <- nrow(data_without_outliers) == (nrow(main) - actual_removed)
ids_match <- identical(
  sort(setdiff(main$idnummer, data_without_outliers$idnummer)),
  sort(all_rows_to_remove)
)

if (!(rows_match && ids_match)) {
  stop("Rows removed do not match expected count.")
}
```

```{r}
#| label: final-data-summary
#| results: asis
#| code-fold: true
#| column: body-outset-right

# generate dataset refinement summary
overlap_text <- if (overlap_count > 0) {
  glue(" &minus; {overlap_count} overlap")
} else {
  ""
}

# identify all SES-related variables present in output
ses_vars_in_output <- c(ses_composites, partner_vars)
ses_vars_present <- ses_vars_in_output[ses_vars_in_output %in% names(data_without_outliers)]

final_summary <- glue(
  "<div class='callout-note'>
  <strong>Dataset Refinement Complete</strong><br><br>
  <strong>Input:</strong> {nrow(main)} observations from Step 3<br>
  <strong>Removed:</strong> {actual_removed} outliers ",
  "({rows_to_remove_uni} univariate + {rows_to_remove_multi} multivariate{overlap_text})<br>
  <strong>Output:</strong> {nrow(data_without_outliers)} observations ",
  "({round(100 * nrow(data_without_outliers) / nrow(main), 1)}% retained)<br><br>
  <strong>SES variables preserved:</strong><br>
  &bull; Parental composites: <code>{paste(ses_present, collapse = '</code>, <code>')}</code><br>
  &bull; Partner indicators: <code>{paste(structural_vars_present, collapse = '</code>, <code>')}</code>
  </div>"
)
cat(final_summary)
```

### Export Refined Dataset

```{r}
#| label: final-data-export
#| code-fold: show
#| column: body-outset-right

# save refined dataset for step 5
haven::write_sav(
  data_without_outliers,
  here::here("assets", "data", "saved", "data_without_outliers.sav")
)

# export outlier case IDs for documentation
outlier_ids_export <- cases_with_outliers %>%
  dplyr::select(idnummer, outlier) %>%
  dplyr::mutate(
    outlier_type = dplyr::case_when(
      grepl("Univariate", outlier) ~ "Univariate",
      grepl("Multivariate", outlier) ~ "Multivariate",
      TRUE ~ "Unknown"
    )
  )

readr::write_csv(
  outlier_ids_export,
  here::here("assets", "data", "saved", "outlier_cases_removed.csv")
)
```

## Descriptive Statistics After Outlier Removal

```{r}
#| label: descriptive-definitions-attrs
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "$\\ $", "$n$", "$\\bar{x}$", "$\\text{Min.}$", "$\\text{Max.}$",
  "$SD$", "$\\text{Skewness}$", "$\\text{Kurtosis}$", "$SE$",
  "$\\text{Normality}^{\\dagger}$"
)

footnote_text <- glue(
  "Shapiro-Wilk Test$^{{\\dagger}}$;",
  "values are normality test statistics (Shapiro-Wilk $W$ for $n \\leq 5000$, Lilliefors $D$ otherwise); ",
  "smaller $W$ (larger $D$) suggests stronger departure from normality;",
  "'&#x2014;' indicates variables with $\\leq 4$ unique values (normality test not applicable)"
)
```

```{r}
#| label: tbl-descriptive-stats-after-outlier-removal
#| tbl-cap: "Descriptive Statistics After Outlier Removal"
#| code-fold: show
#| column: page-right

data_without_outliers %>%
  dplyr::select(-idnummer) %>%
  setNames(beautify_names(names(.))) %>%
  describe(
    normality = TRUE,
    print_only = FALSE
  ) %>%
  dplyr::rowwise() %>%
  dplyr::ungroup() %>%
  dplyr::select(-p) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                                          | Content                                                           |
|-----------------------------------------------|-------------------------------------------------------------------|
| `assets/data/saved/data_without_outliers.sav` | Refined dataset with univariate and multivariate outliers removed |
| `assets/data/saved/outlier_cases_removed.csv` | ID and detection type for each excluded case                      |

**Next:** [Step 5 (Missing Data Imputation)](/analysis/data/imputation/index.qmd)
