---
title: "Interaction Effects & Johnson–Neyman Analysis"
subtitle: "Identifying Moderator Ranges for Significant Predictor Effects"
author: "Siard van den Bosch"
date: 2025-03-06
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 11};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Identify moderator ranges where gender × SES interaction effects reach significance |
| **Inputs**     | `adjusted_data.rds`, `unadjusted_fits.rds`, `adjusted_fits.rds` |
| **Outputs**    | none |
| **Method**     | Johnson–Neyman intervals on bootstrap-derived vcov; Freedman–Diaconis binning |
| **Requires**   | [Step 7 (Unadjusted Models)](/analysis/models/unadjusted/index.qmd), [Step 8 (Influence Diagnostics)](/analysis/models/assumptions/index.qmd), [Step 9 (Adjusted Models)](/analysis/models/adjusted/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- Johnson–Neyman over pick-a-point: delineates exact moderator boundaries rather than testing simple slopes at chosen values [@finsaas2021]
- Freedman–Diaconis binning: data-adaptive bin width robust to skewed moderator distributions [@bauer2005]
- Binary-moderator interactions skipped: JN intervals require a continuous or multi-level moderator [@johnson1936]
:::

::: {style="margin-block-end: 1rem;"}
:::

## Load Project Resources

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear environment
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions (consistent with steps 7-9)
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "as_kable", "footnote", "set_globals", "beautify_names", "structure_models",
  "sci_to_tex", "row_grouping", "md_to_html", "pval_str", "format_p_value",
  "dim_ci_rows", "add_style", "format_coef", "import_pipes", "numeric_df",
  "suppress_output", "named_list"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()
```

## Load Data and Models

```{r}
#| label: load-data-and-models
#| code-fold: show
#| column: body-outset-right

# load adjusted data from step 8 (sensitivity dataset)
adjusted_data_path <- here::here("assets", "data", "saved", "adjusted_data.rds")
if (!file.exists(adjusted_data_path)) {
  stop("Missing input from Step 8: ", adjusted_data_path)
}
source_data <- readRDS(adjusted_data_path)

# load model fits from steps 7 and 9
unadjusted_fits_path <- here::here("assets", "models", "unadjusted_fits.rds")
adjusted_fits_path <- here::here("assets", "models", "adjusted_fits.rds")

if (!file.exists(unadjusted_fits_path)) {
  stop("Missing input from Step 7: ", unadjusted_fits_path)
}
if (!file.exists(adjusted_fits_path)) {
  stop("Missing input from Step 9: ", adjusted_fits_path)
}

# load fits and structure into nested list (unadjusted/adjusted × ols/robust)
unadjusted_fits <- readRDS(unadjusted_fits_path)
adjusted_fits <- readRDS(adjusted_fits_path)
model_fits <- structure_models(unadjusted_fits, adjusted_fits)

# assign model class to robust fits for compatibility with interactions package
lmrobs <- model_fits$adjusted$robust %>%
  purrr::map(~ `class<-`(.x, c("model", class(.x))))
```

```{r}
#| label: verify-model-structure
#| results: asis
#| code-fold: true
#| column: body-outset-right

# extract model identifiers for verification
model_ids <- names(model_fits$adjusted$robust) %>%
  stringr::str_extract("[0-9]+$") %>%
  as.integer() %>%
  sort()

n_models <- length(model_ids)
model_range <- paste0("Models ", min(model_ids), "–", max(model_ids))
n_obs <- nrow(source_data)

# identify models with second-order interactions (eligible for JN analysis)
models_with_interactions <- model_fits$adjusted$robust %>%
  purrr::map_lgl(function(m) {
    if (is.null(m$fit)) {
      return(FALSE)
    }
    terms_ <- terms(m$fit)
    order_ <- attr(terms_, "order")
    any(order_ == 2)
  })

interaction_model_ids <- names(model_fits$adjusted$robust)[models_with_interactions] %>%
  stringr::str_extract("[0-9]+$") %>%
  as.integer() %>%
  sort()

# binary moderator patterns (partner status is binary: 0/1)
binary_patterns <- c("partner:education_partner", "partner:isei08_partner")

# identify models with only binary-moderator interactions (excluded from JN)
models_binary_only <- model_fits$adjusted$robust %>%
  purrr::imap_chr(function(m, name) {
    if (is.null(m$fit)) {
      return(NA_character_)
    }
    terms_ <- terms(m$fit)
    labels_ <- attr(terms_, "term.labels")[attr(terms_, "order") == 2]
    if (length(labels_) == 0) {
      return(NA_character_)
    }
    if (all(labels_ %in% binary_patterns)) gsub("model_", "", name) else NA_character_
  }) %>%
  na.omit() %>%
  as.character()

# models eligible for JN (have non-binary interactions)
jn_eligible <- setdiff(interaction_model_ids, as.integer(models_binary_only))

# verification callout
verify_msg <- glue::glue(
  "<div class='callout-note'>
<strong>Step 7/9 Model Verification</strong><br><br>
<strong>Models loaded:</strong> {n_models} models ({model_range})<br>
<strong>Analysis dataset:</strong> {n_obs} observations (sensitivity sample from Step 8)<br>
<strong>Models with interactions:</strong> {paste(interaction_model_ids, collapse = ', ')}<br>
<strong>JN analysis applicable to:</strong> Models {paste(jn_eligible, collapse = ', ')}
{if (length(models_binary_only) > 0) paste0('<br><em>Models ', paste(models_binary_only, collapse = ', '), ' skipped (binary moderator only)</em>') else ''}
</div>"
)
cat(verify_msg)

if (length(jn_eligible) == 0) {
  warning("No models with continuous moderator interactions. JN analysis not applicable.")
}
```

## Johnson–Neyman Functions

```{r}
#| label: calculate-optimal-interval-function
#| code-fold: true
#| column: body-outset-right

# calculate optimal bin width for moderator variable
calculate_optimal_interval <- function(data, method = c("fd", "scott", "sturges")) {
  method <- match.arg(method)

  if (is.null(data) || nrow(data) == 0) {
    return(NULL)
  }

  modx <- data$modx
  modx_range <- range(modx, na.rm = TRUE)
  total_range <- diff(modx_range)

  if (total_range < .Machine$double.eps) {
    warning("Moderator has no variance; cannot compute interval.")
    return(NULL)
  }

  interval <- switch(method,
    fd = 2 * IQR(modx, na.rm = TRUE) / (nrow(data)^(1 / 3)),
    scott = 3.5 * sd(modx, na.rm = TRUE) / (nrow(data)^(1 / 3)),
    sturges = total_range / (log2(nrow(data)) + 1)
  )

  if (is.na(interval) || interval < .Machine$double.eps) {
    interval <- total_range / 10
  }

  n_bins <- total_range / interval
  if (n_bins < 5) interval <- total_range / 5
  if (n_bins > 20) interval <- total_range / 20

  magnitude <- 10^floor(log10(interval))
  round(interval / magnitude) * magnitude
}
```

```{r}
#| label: validate-jn-intervals-input
#| code-fold: true
#| column: body-outset-right

# validate inputs for JN interval summarization
validate_jn_intervals_input <- function(x, interval = NULL) {
  if (is.null(x)) {
    warning("Input to summarize_jn_intervals is NULL.")
    return(NULL)
  }

  required_columns <- c("modx", "slope", "conf.low", "conf.high", "sig")

  validated <- purrr::map(x, function(elem) {
    if (is.null(elem) || !is.data.frame(elem)) {
      warning("Skipping NULL or non-data frame element.")
      return(NULL)
    }

    if (!all(required_columns %in% colnames(elem))) {
      warning("Required columns are missing from data. Skipping.")
      return(NULL)
    }

    int_terms <- attr(elem, "terms")
    if (is.null(int_terms) || length(int_terms) != 2) {
      warning("Invalid or missing terms attribute. Skipping.")
      return(NULL)
    }

    elem_interval <- interval
    if (is.null(elem_interval)) {
      elem_interval <- calculate_optimal_interval(elem, method = "fd")
      if (is.null(elem_interval)) {
        warning("Interval calculation failed. Skipping.")
        return(NULL)
      }
    }

    list(data = elem, interval = elem_interval, terms = int_terms)
  })

  purrr::compact(validated)
}
```

```{r}
#| label: summarize-jn-intervals-function
#| code-fold: true
#| column: body-outset-right

# summarize JN analysis results into interpretable intervals
summarize_jn_intervals <- function(jn_result, interval = NULL) {
  x <- jn_result$data
  model_name <- deparse(jn_result$call$model)

  validated_inputs <- validate_jn_intervals_input(x, interval)
  if (is.null(validated_inputs) || length(validated_inputs) == 0) {
    return(NULL)
  }

  summaries <- purrr::map(validated_inputs, function(input) {
    input$data %>%
      dplyr::mutate(modx_grp = round(modx / input$interval) * input$interval) %>%
      dplyr::group_by(modx_grp) %>%
      dplyr::summarize(
        avg_slope = mean(slope, na.rm = TRUE),
        min_conf = min(conf.low, na.rm = TRUE),
        max_conf = max(conf.high, na.rm = TRUE),
        sig_count = sum(sig == "Significant", na.rm = TRUE),
        n_obs = dplyr::n(),
        sig_perc = ifelse(n_obs > 0, round((sig_count / n_obs) * 100, 1), NA_real_),
        .groups = "drop"
      ) %>%
      dplyr::arrange(modx_grp) %>%
      dplyr::mutate(
        term1 = beautify_names(input$terms[[1]]),
        term2 = beautify_names(input$terms[[2]]),
        int_size = input$interval
      ) %>%
      `attr<-`("n_groups", nrow(.)) %>%
      `attr<-`("int_size", input$interval)
  })

  summaries %>%
    purrr::compact() %>%
    dplyr::bind_rows() %>%
    dplyr::mutate(model = gsub("model_", "", model_name)) %>%
    dplyr::select(model, term1, term2, int_size, dplyr::everything())
}
```

```{r}
#| label: second-order-interactions-function
#| code-fold: true
#| column: body-outset-right

# extract second-order interaction terms from model
second_order_interactions <- function(fit) {
  terms_ <- terms(fit)
  order_ <- attr(terms_, "order")

  inter_terms <- attr(terms_, "term.labels")
  inter_terms <- inter_terms[order_ == 2]

  strsplit(inter_terms, ":")
}
```

```{r}
#| label: plot-utilities
#| code-fold: true
#| column: body-outset-right

# clean linetype mappings in ggplot layers
clean_layer_linetypes <- function(plot) {
  if (!inherits(plot, "ggplot")) {
    stop("The input must be a ggplot object.")
  }

  plot$layers <- purrr::map(plot$layers, function(layer) {
    .linetype <- layer$mapping$linetype
    if (!is.null(.linetype)) {
      layer$mapping$linetype <- rlang::quo(
        stringr::str_squish(rlang::as_name(.linetype))
      )
    }
    layer
  })

  plot
}

# format variable names with smaller parenthetical text for plot labels
shrink_parens <- function(txt) {
  txt <- beautify_names(txt, as_HTML = FALSE)
  match_res <- regexpr("\\s+\\(([^)]+)\\)", txt, perl = TRUE)

  if (match_res < 1) {
    return(bquote(bold(.(txt))))
  }

  in_txt <- regmatches(txt, match_res)
  bef_txt <- substr(txt, 1, match_res - 1)
  aft_start <- match_res + attr(match_res, "match.length")
  aft_txt <- if (aft_start <= nchar(txt)) {
    substr(txt, aft_start, nchar(txt))
  } else {
    ""
  }

  bquote(bold(.(bef_txt)) ~ bold(scriptstyle(.(in_txt))) ~ bold(.(aft_txt)))
}
```

```{r}
#| label: jn-eval-function
#| code-fold: true
#| column: body-outset-right

# validate inputs and extract variance-covariance matrix for JN analysis
jn_eval <- function(model, pred_var, mod_var) {
  df <- model.frame(model$fit)

  if (!(pred_var %in% names(df))) {
    stop(glue::glue("Predictor variable {pred_var} not found"))
  }
  if (!(mod_var %in% names(df))) {
    stop(glue::glue("Moderator variable {mod_var} not found"))
  }
  if (is.null(model$bootstrap$ests)) {
    stop("No bootstrap estimates found in model; bootstrapping not performed")
  }

  boot_mat <- as.matrix(model$bootstrap$ests)
  vcov_mat <- tryCatch(
    cov(boot_mat),
    error = function(err) {
      stop(glue::glue("Error calculating vcov matrix: {err$message}"))
    }
  )

  if (interactive()) {
    message(glue::glue("Variables used in interaction:\n{pred_var}\n{mod_var}"))
  }

  vcov_mat
}
```

```{r}
#| label: jn-analysis
#| code-fold: show
#| column: body-outset-right

# test whether a variable is effectively binary (at most 2 unique non-NA values)
is_binary_var <- function(v, df) {
  vals <- stats::na.omit(df[[v]])
  length(unique(vals)) <= 2L
}

# perform Johnson-Neyman analysis for all second-order interactions
jn_analysis <- function(model, dep_var = "DHD-index", alpha = 0.05) {
  call_expr <- match.call()
  int_terms <- second_order_interactions(model$fit)

  if (length(int_terms) == 0) {
    message("No second-order interactions found in the model")
    return(NULL)
  }

  df <- model.frame(model$fit)

  results <- purrr::map(int_terms, function(term) {
    # orient predictor vs moderator by type, not by alphabetical position
    bin <- vapply(term, is_binary_var, logical(1), df = df)

    if (sum(bin) == 2L) {
      # both binary: JN is not informative, skip
      message(glue::glue(
        "Skipping Johnson-Neyman analysis for {paste(term, collapse = ':')} ",
        "(both variables binary; no continuous moderator)."
      ))
      return(NULL)
    } else if (sum(bin) == 1L) {
      # one binary, one continuous: binary is predictor, continuous moderates
      pred_var <- term[bin]
      mod_var <- term[!bin]
    } else {
      # both continuous: fall back to term order (either orientation valid)
      mod_var <- term[[1]]
      pred_var <- term[[2]]
    }

    # compute bootstrap-based vcov matrix
    vcov_mat <- jn_eval(model, pred_var, mod_var)

    # run JN analysis via interactions package
    jn_res <- tryCatch(
      interactions::johnson_neyman(
        model = model$fit,
        pred = !!rlang::sym(pred_var),
        modx = !!rlang::sym(mod_var),
        vmat = vcov_mat,
        alpha = alpha,
        plot = FALSE,
        df = "residual"
      ),
      error = function(err) {
        message(glue::glue(
          "Johnson-Neyman analysis failed for {pred_var} and {mod_var}: {err$message}"
        ))
        NULL
      }
    )

    if (is.null(jn_res)) {
      return(NULL)
    }

    # standardize column names from interactions output
    jn_data <- jn_res$cbands %>%
      dplyr::rename_with(~"slope", dplyr::starts_with("Slope of ")) %>%
      dplyr::rename(
        modx = !!rlang::sym(mod_var),
        conf.low = Lower,
        conf.high = Upper,
        sig = Significance
      )

    list(
      data = jn_data,
      raw_plot = clean_layer_linetypes(jn_res$plot),
      pred_var = pred_var,
      mod_var = mod_var,
      dep_var = dep_var
    )
  })

  # filter to valid results and attach terms
  valid_idx <- purrr::map_lgl(results, ~ !is.null(.x))
  valid_terms <- int_terms[valid_idx]
  valid_out <- purrr::compact(results)
  jn_data_list <- purrr::compact(purrr::map(valid_out, ~ .x$data))

  for (i in seq_along(jn_data_list)) {
    attr(jn_data_list[[i]], "terms") <- valid_terms[[i]]
  }

  res <- list(
    data = jn_data_list,
    raw_plots = valid_out,
    model = model,
    call = call_expr
  )
  class(res) <- "jn_result"
  res
}
```

```{r}
#| label: jn-result-methods
#| code-fold: true
#| column: body-outset-right

# print method for jn_result objects
print.jn_result <- function(x, ...) {
  cat("Johnson-Neyman Analysis Result\n")
  cat("Number of interactions analyzed:", length(x$data), "\n")
  invisible(x)
}

# plot method for jn_result objects
plot.jn_result <- function(x, plot_significant_only = FALSE, ...) {
  raw_plots <- x$raw_plots

  if (length(raw_plots) == 0) {
    message("No plots available")
    return(invisible(NULL))
  }

  model_frame <- model.frame(x$model$fit)

  has_significant <- function(jn_data) {
    any(jn_data$sig == "Significant")
  }

  if (plot_significant_only) {
    raw_plots <- purrr::keep(raw_plots, ~ has_significant(.x$data))
  }

  if (length(raw_plots) == 0) {
    message("No plots with significant ranges to display")
    return(invisible(NULL))
  }

  # generate customized plots
  final_plots <- purrr::map(raw_plots, function(res) {
    slope_label <- bquote(
      atop(
        "Slope of " ~ .(shrink_parens(res$pred_var)) ~
          " on " ~ bold("DHD-index"),
        "for values of " ~ .(shrink_parens(res$mod_var))
      )
    )

    p <- res$raw_plot +
      ggplot2::labs(
        title = NULL,
        x = shrink_parens(res$mod_var),
        y = slope_label,
        fill = NULL,
        color = NULL
      ) +
      ggplot2::scale_fill_manual(
        values = c(
          "Data range" = "black",
          "Insignificant" = "grey",
          "Significant" = "#21908C"
        ),
        labels = c(
          "Data range" = "Data range",
          "Insignificant" = bquote(italic("n.s.")),
          "Significant" = bquote(italic("p < .05"))
        ),
        breaks = c("Data range", "Insignificant", "Significant"),
        guide = ggplot2::guide_legend(order = 1, title = NULL)
      ) +
      ggplot2::scale_color_manual(
        values = c("Insignificant" = "grey", "Significant" = "#21908C"),
        guide = "none"
      ) +
      garamond_theme +
      ggplot2::theme(
        axis.title.y = ggplot2::element_text(size = 11, hjust = 0.5),
        axis.title.x = ggplot2::element_text(
          size = 12,
          margin = ggplot2::margin(t = 10)
        ),
        legend.key.size = grid::unit(0.4, "cm"),
        legend.key.width = grid::unit(0.4, "cm"),
        legend.key.height = grid::unit(0.4, "cm"),
        legend.spacing.x = grid::unit(0.2, "cm"),
        legend.text = ggplot2::element_text(size = 8, hjust = 0.5, vjust = 0.5),
        legend.box.spacing = grid::unit(0.3, "cm"),
        legend.box.margin = ggplot2::margin(t = 0, r = 5, b = 5, l = 5),
        plot.margin = ggplot2::margin(t = 50, r = 15, b = 30, l = 15)
      )

    # add rug plot showing moderator distribution
    if (res$mod_var %in% names(model_frame)) {
      mod_values <- model_frame[[res$mod_var]]
      n_unique <- length(unique(mod_values))

      # for binary or low-cardinality moderators, use simple single-color rug
      if (n_unique <= 5) {
        p <- p +
          ggplot2::geom_rug(
            data = model_frame,
            ggplot2::aes(x = .data[[res$mod_var]]),
            sides = "b",
            color = "#440154",
            alpha = 0.6,
            linewidth = 0.4,
            length = grid::unit(0.02, "npc"),
            outside = FALSE,
            inherit.aes = FALSE
          ) +
          ggplot2::coord_cartesian(clip = "off") +
          ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = 0.05))
      } else {
        # for continuous moderators, use viridis color scale
        p <- p +
          ggnewscale::new_scale_color() +
          ggplot2::geom_rug(
            data = model_frame,
            ggplot2::aes(
              x = .data[[res$mod_var]],
              color = .data[[res$mod_var]]
            ),
            sides = "b",
            alpha = 0.8,
            linewidth = 0.5,
            length = grid::unit(0.02, "npc"),
            outside = FALSE,
            inherit.aes = FALSE
          ) +
          ggplot2::scale_color_viridis_c(
            option = "viridis",
            name = beautify_names(res$mod_var, as_HTML = FALSE),
            guide = ggplot2::guide_colorbar(
              title.position = "top",
              barheight = grid::unit(0.4, "cm"),
              barwidth = grid::unit(2.5, "cm"),
              order = 2
            )
          ) +
          ggplot2::coord_cartesian(clip = "off") +
          ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = 0.05))
      }
    } else {
      warning(glue::glue(
        "Moderator variable '{res$mod_var}' not found in model frame. ",
        "Rug plot not added."
      ))
    }

    p
  })

  old_par <- graphics::par(mfrow = c(ceiling(length(final_plots) / 2), 2))
  on.exit(graphics::par(old_par))

  purrr::walk(final_plots, print)
  invisible(final_plots)
}
```

## JN Results

```{r}
#| label: attach-model-fits
#| code-fold: show
#| column: body-outset-right

# attach robust model fits for formula-style access
robust_models <- model_fits$adjusted$robust
if (!"robust_models" %in% search()) {
  attach(robust_models)
}

# extract sample size for footnotes
n_obs <- nrow(model.frame(model_fits$adjusted$robust$model_1$fit))
```

```{r}
#| label: jn-table-attributes
#| code-fold: true
#| column: body-outset-right

# column names for JN interval tables
col_names <- c(
  "Predictor", "$\\Delta x$", "Modx Group",
  "$\\overline{\\beta}$", "$\\text{CI}_{\\text{low}}$",
  "$\\text{CI}_{\\text{high}}$", "$\\Sigma \\text{ Sig.}$",
  "$\\Sigma \\text{ Total}$", "$\\% \\text{Sig.}$"
)

# footnote text for JN interval tables
footnote_text <- glue::glue(
  "$N={n_obs}$ (sensitivity sample excluding influential observations). ",
  "For each moderator bin ($\\Delta x$), mean slopes ($\\overline{{\\beta}}$) ",
  "and 95\\% CIs are shown. $\\Sigma$ Sig. counts points within the bin where ",
  "$p \\lt .05$; \\% Sig. = $\\Sigma$ Sig. / $\\Sigma$ Total. ",
  "Greyed rows indicate CIs spanning zero."
)
```

### Model 4: Gender × SES (Composite)

```{r}
#| label: jn-model-4-plot
#| fig-cap: "Johnson–Neyman Regions (Model 4 – Composite Gender × SES)"
#| fig-subcap: true
#| layout-ncol: 3
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

m4 <- jn_analysis(model_4) %T>%
  plot(plot_significant_only = FALSE) %>%
  summarize_jn_intervals()
```

```{r}
#| label: jn-model-4-df
#| tbl-cap: "Johnson–Neyman Intervals (Model 4)"
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

if (!is.null(m4) && nrow(m4) > 0) {
  m4 %>%
    dplyr::select(-term2, -model) %>%
    as_kable() %>%
    dim_ci_rows(m4, opacity = 0.25) %>%
    kableExtra::pack_rows(index = row_grouping(m4$term2), escape = FALSE) %>%
    footnote() %>%
    add_style()
} else {
  cat("No JN intervals available for Model 4 (no significant interactions or missing bootstrap estimates).")
}
```

### Model 9: Gender × SES (Disaggregated)

```{r}
#| label: jn-model-9-plot
#| fig-cap: "Johnson–Neyman Regions (Model 9 – Disaggregated Gender × SES)"
#| fig-subcap: true
#| layout-ncol: 3
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

m9 <- jn_analysis(model_9) %T>%
  plot(plot_significant_only = FALSE) %>%
  summarize_jn_intervals()
```

```{r}
#| label: jn-model-9-df
#| tbl-cap: "Johnson–Neyman Intervals (Model 9)"
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

if (!is.null(m9) && nrow(m9) > 0) {
  m9 %>%
    dplyr::select(-term2, -model) %>%
    as_kable() %>%
    dim_ci_rows(m9, opacity = 0.25) %>%
    kableExtra::pack_rows(index = row_grouping(m9$term2), escape = FALSE) %>%
    footnote() %>%
    add_style()
} else {
  cat("No JN intervals available for Model 9 (no significant interactions or missing bootstrap estimates).")
}
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File   | Content                                                    |
| ------ | ---------------------------------------------------------- |
| none   | step renders inline tables and figures; no files persisted |

**Next:** [Step 12 (Bootstrapped Validation)](/analysis/models/goodness-of-fit/performance-metrics/index.qmd)
