---
title: "Data Cleaning & Recoding"
subtitle: "Transforming Variables for Consistency and Imputation Readiness"
author: "Siard van den Bosch"
date: 2024-08-26
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 3};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Recode variables, screen missingness, and construct SES composites for imputation |
| **Inputs**     | `main_corrected.sav` |
| **Outputs**    | `data_for_imputation.sav` |
| **Method**     | Variable recoding; MCAR diagnostic; psychometric scale evaluation |
| **Requires**   | [Step 2 (ISCO-08 Standardization)](/analysis/data/preparation/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- Critical-missing exclusion: drop cases missing both DHD index and kilocalories rather than impute pivotal variables [@enders2010]
- Retention threshold: 20% per-case missingness cap balances sample retention against imputation quality [@schafer2002]
- Structural missingness: partner indicators set to NA for unpartnered respondents to separate from MAR [@little1988]
- Domain-specific SES: education and occupation index distinct mechanisms (cultural capital vs. material resources) [@galobardes2006]
- Scale validity criteria: $\alpha$, AVE, Fornell-Larcker, and HTMT follow established psychometric frameworks [@fornell1981]
:::

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear global env and force full garbage collection
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))

# detach non-base packages to avoid namespace interference
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions for data processing and analysis
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "md_to_html", "format_coef", "as_kable", "describe", "sci_to_tex",
  "add_style", "footnote", "order_array", "numeric_df", "update_isco08r",
  "brc14_map", "import_pipes", "update_value_label", "labels_to_string",
  "format_p_value", "set_globals", "isco08_to_jobtitle", "update_brc14",
  "frequency_table", "beautify_names", "summarize_na", "suppress_output"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()

# load corrected dataset from step 2 and adjust variable types
main <- haven::read_sav(
  here::here("assets", "data", "saved", "main_corrected.sav"),
  user_na = TRUE
) %>%
  dplyr::mutate(dplyr::across(dplyr::starts_with("brc14"), as.integer)) %>%
  dplyr::select(-dplyr::matches("(_cleaned|_init)$|_old"))
```

## Select Variables for Analysis

Excludes character-type, derived, and previously-recoded variables; retains identifiers and analytic indicators.

```{r}
#| label: filter-select-variables
#| tbl-cap: Variables Excluded from Imputation
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# standardize variable names to lowercase
names(main) %<>% tolower()

# summarize missing values for each variable
missing_values_summary <- summarize_na(main)

# function to determine variable type from character representation
# Update this function in your Quarto file
get_variable_type <- function(x, ...) {
  # Convert haven_labelled to plain numeric or character first
  if (inherits(x, "haven_labelled")) {
    x <- haven::zap_labels(x)
  }
  class(type.convert(as.character(x), as.is = TRUE))
}

# derived variables to exclude from imputation (identifiers, computed composites)
derived_variables <- c(
  "idnummer", "crowding", "hh_comp", "hh_members", "eq_size",
  "income_hh_num", "income_hh_eq", "children_hh_yes"
)

# create dataframe with variable types and exclusion flags
column_types <- stack(sapply(main, get_variable_type))[2:1]
names(column_types) <- c("variable", "character")

# patterns for identifying recoded or classification variables
exclusion_patterns <- paste(
  ".*_(cat|rec)$",
  "^ses_index.*",
  "^isco[2008]+[^r].*",
  "^brc14[_a-z]*",
  "^skill_level",
  sep = "|"
)

# flag variables as character, recoded, or derived
column_types[c("character", "recoded", "derived")] <- lapply(list(
  column_types$character == "character",
  grepl(exclusion_patterns, column_types$variable),
  column_types$variable %in% derived_variables
), as.integer)

# determine which variables to keep (no exclusion flags set)
column_types$keep <- rowSums(column_types[, -1]) == 0

# merge with missing values summary for complete variable info
variable_info <- column_types %>%
  dplyr::mutate(dplyr::across(
    -variable,
    ~ c("\u00b7", "\u00d7")[as.numeric(.) + 1]
  )) %>%
  dplyr::inner_join(missing_values_summary, by = "variable")

# mark occupational classification variables to keep despite exclusion flags;
# lf_status_* are character and would otherwise be dropped by the character flag
variable_info[grep("isco08r|brc14|isei08|siops08|egp11|lf_status", variable_info$variable), "keep"] <- "\u00d7"

# sort for display: kept variables first, then by exclusion reason
d <- dplyr::desc
variable_info %<>%
  dplyr::mutate(kd = as.integer(paste0(keep, derived) == "\u00d7\u00d7")) %>%
  dplyr::arrange(d(keep), kd, d(derived), d(recoded)) %>%
  dplyr::select(-kd)

# subset data to keep only selected variables plus identifier
data_cleaned <- main %>%
  subset(select = c(
    "idnummer",
    variable_info$variable[variable_info$keep == "\u00d7"]
  ))
```


```{r}
#| label: filter-select-variables-attrs
#| tbl-cap: Variables Excluded from Imputation
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "Variable", "Character", "Recoded", "Derived",
  "Keep", "$\\textit{N}_{missing}$", "$\\%_{missing}$"
)

footnote_text <- glue(
  "$N = {nrow(main)}$; $n = {nrow(variable_info)}$; ",
  "The symbol '\u00d7' indicates that the variable meets the exclusion criteria ",
  "specified in the column header. Variables that were recoded, character type, ",
  "or derived were excluded from imputation."
)

# display exclusion summary table
variable_info %>%
  as_kable(omit_leading_zero = FALSE) %>%
  kableExtra::add_header_above(
    c(" " = 1, "Reasons for Exclusion" = 3, " " = 3)
  ) %>%
  footnote() %>%
  add_style()
```

## Recode Variables for Analysis

Transformations enforce consistent directionality, with higher values denoting greater exposure or difficulty.

### Number of Children in Household

```{r}
#| label: tbl-children-hh
#| tbl-cap: "Distribution of the Number of Children in the Household"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "children_hh")

# recode to start from zero (original: 1 = 0 children, 2 = 1 child, etc.)
data_cleaned %<>%
  dplyr::mutate(num_children_hh = children_hh - 1) %T>%
  frequency_table("num_children_hh")
```

### Smoking Habit

```{r}
#| label: tbl-smoking
#| tbl-cap: "Distribution of Smoking Habit"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "q63")

# recode smoking status: 0 = never, 1 = former, 2 = current
data_cleaned %<>%
  dplyr::mutate(smoking = labelled::labelled(
    as.numeric(q63) - 1,
    c("No, never did" = 0, "No, but used to" = 1, "Yes" = 2)
  )) %T>%
  frequency_table("smoking")
```

### Week Normality Indicator

```{r}
#| label: tbl-normal-week
#| tbl-cap: "Normality of Surveyed Week"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "q79")

# recode to binary: 1 = normal week, 0 = not normal (original: 2 = not normal)
data_cleaned %<>%
  dplyr::mutate(normal_week = labelled::labelled(
    as.integer(q79 != 2), c("No" = 0, "Yes" = 1)
  )) %T>%
  frequency_table("normal_week")

# validate recoding produced expected values only
if (any(!data_cleaned$normal_week %in% c(0, 1, NA), na.rm = TRUE)) {
  warning("Unexpected values found in normal_week after recoding")
}
```

### Home Ownership Status

```{r}
#| label: tbl-home-ownership
#| tbl-cap: "Home Ownership Status"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "q35")

# recode to binary: 1 = owner, 0 = not owner (original: 1 = owner)
data_cleaned %<>%
  dplyr::mutate(home_owner = labelled::labelled(
    as.integer(q35 == 1), c("No" = 0, "Yes" = 1)
  )) %T>%
  frequency_table("home_owner")
```

### Gender of Respondent

```{r}
#| label: tbl-gender
#| tbl-cap: "Gender (Respondent is Woman)"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "gender")

# recode to binary: 1 = woman, 0 = man (original: 1 = woman)
data_cleaned %<>%
  dplyr::mutate(woman = labelled::labelled(
    as.integer(gender == 1), c("Man" = 0, "Woman" = 1)
  )) %T>%
  frequency_table("woman")
```

### Difficulty Living from Income

```{r}
#| label: tbl-difficulty-living-income
#| tbl-cap: "Difficulty Living from Income"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "diff_live_income")

# reverse code: 0 = very easy (no difficulty) to 3 = great difficulty
data_cleaned %<>%
  dplyr::mutate(diff_live_income = labelled::labelled(
    4 - as.integer(diff_live_income),
    c(
      "Very easy (no difficulty)" = 0, "Relatively easy" = 1,
      "Some difficulty" = 2, "Great difficulty" = 3
    )
  )) %T>%
  frequency_table("diff_live_income")

# validate recoding produced expected values only (missing excluded)
if (any(!na.omit(as.integer(data_cleaned$diff_live_income)) %in% 0:3)) {
  warning("Unexpected values after recoding")
}
```

### Difficulty Paying Regular Bills

```{r}
#| label: tbl-difficulty-paying-bills
#| tbl-cap: "Difficulty Paying Regular Bills"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "diff_pay_bills")

# recode to 0-2 scale: 0 = no difficulty, 1 = some, 2 = great difficulty
data_cleaned %<>%
  dplyr::mutate(diff_pay_bills = labelled::labelled(
    as.integer(diff_pay_bills) - 1,
    c(
      "No, no difficulty at all" = 0,
      "Some difficulty" = 1,
      "Yes, great difficulty" = 2
    )
  )) %T>%
  frequency_table("diff_pay_bills")

# validate recoding produced expected values only (missing excluded)
if (any(!na.omit(as.integer(data_cleaned$diff_pay_bills)) %in% 0:2)) {
  warning("Unexpected values after recoding")
}
```

### Education Level

```{r}
#| label: education-recoding
#| code-fold: show
#| column: body-outset-right

# education labels for recoded scale
# 0 = no education, 6 = higher professional
education_labels <- c(
  "no education" = 0,
  "primary school" = 1,
  "lower vocational education" = 2,
  "general secondary education" = 3,
  "secondary vocational education" = 4,
  "higher general secondary education" = 5,
  "higher professional education" = 6
)

# recode all education variables to start from 0
data_cleaned <- data_cleaned %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::matches("^education($|_)"),
      ~ {
        x <- as.numeric(.)
        x[!is.na(x)] <- x[!is.na(x)] - 1
        out <- labelled::labelled(x, education_labels)
        attr(out, "label") <- attr(., "label")
        out
      }
    )
  )
```

```{r}
#| label: tbl-respondent-education
#| tbl-cap: "Respondent's Education Level"
#| results: asis
#| code-fold: true
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "education")
invisible(data_cleaned %T>% frequency_table("education"))
```

```{r}
#| label: tbl-mother-education
#| tbl-cap: "Mother's Education Level"
#| results: asis
#| code-fold: true
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "education_mother")
invisible(data_cleaned %T>% frequency_table("education_mother"))
```

```{r}
#| label: tbl-father-education
#| tbl-cap: "Father's Education Level"
#| results: asis
#| code-fold: true
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "education_father")
invisible(data_cleaned %T>% frequency_table("education_father"))
```

### Partner Status

```{r}
#| label: tbl-partner-status
#| tbl-cap: "Partner Status"
#| results: asis
#| code-fold: show
#| column: body-outset-right

footnote_text <- labels_to_string(data_cleaned, "partneryn")

# recode to binary: 1 = has partner, 0 = no partner
data_cleaned <- data_cleaned %>%
  dplyr::mutate(partner = haven::labelled(
    as.integer(partneryn == 1),
    c("No" = 0, "Yes" = 1)
  )) %T>%
  frequency_table("partner")
```

```{r}
#| label: drop-original-variables-after-recoding
#| code-fold: true
#| column: body-outset-right

# remove original variables post-recoding to avoid redundancy
data_cleaned %<>%
  dplyr::select(-children_hh, -q63, -q79, -q35, -gender, -partneryn)
```

### Handle Partner-Specific Variables

```{r}
#| label: tbl-partner-education
#| tbl-cap: "Partner's Education Level"
#| results: asis
#| code-fold: show
#| column: body-outset-right

# set partner education and occupation to NA for respondents without partners
data_cleaned %<>%
  dplyr::mutate(dplyr::across(
    c(isei08_partner, education_partner),
    ~ dplyr::if_else(partner == 0, NA_real_, .)
  )) %>%
  labelled::drop_unused_value_labels()

# update value label for structural missingness
update_value_label(data_cleaned, "education_partner", NA_real_, "No partner")

footnote_text <- glue(
  "{labels_to_string(data_cleaned, 'education_partner')}<br><i>Note</i>: ",
  "Missing values for partner's education were recoded to 'No partner' ",
  "for respondents without a partner."
)

invisible(data_cleaned %T>% frequency_table("education_partner"))
```

```{r}
#| label: define-classification-update-functions
#| code-fold: show
#| column: body-outset-right

# update BRC14 variable labels from mapping table
update_brc14 <- function(column_name, data, mapping) {
  values <- unique(stats::na.omit(data[[column_name]]))
  labels <- sapply(values, function(x) {
    unique(mapping$BRC2014beroepsgroep_label[
      as.numeric(mapping$BRC2014beroepsgroep) == x
    ])
  })
  values_named <- setNames(values, labels)
  for (i in seq_along(values_named)) {
    eval.parent(substitute(
      update_value_label(
        data,
        column_name,
        as.numeric(values_named[i]),
        names(values_named)[i]
      )
    ))
  }
}

# label lookup for reconciled isco08r codes; labels come from the brc2014
# crosswalk because the job-title package lookup does not know the z.n.d. and
# aggregate codes now present (e.g. 0000, 1000, 2300, 6100, 7000, 8300)
isco08_label_map <- readr::read_csv(
  here::here("assets", "data", "brc2014_isco08_brc_crosswalk.csv"),
  col_types = readr::cols(.default = readr::col_character()),
  progress = FALSE
) %>%
  dplyr::transmute(
    isco08 = stringr::str_pad(trimws(ISCO2008unitgroup), 4, pad = "0"),
    label = ISCO2008unitgrouplabel
  ) %>%
  dplyr::distinct()

# return the brc2014 label for numeric isco08r values; codes are matched as
# four-character zero-padded text; an unknown code stops the render
isco08_to_label <- function(x) {
  code <- sprintf("%04d", as.integer(x))
  label <- isco08_label_map$label[match(code, isco08_label_map$isco08)]
  stopifnot(!anyNA(label))
  label
}

# update ISCO-08 variable labels from the brc2014 crosswalk lookup
update_isco08r <- function(column_name, data) {
  values <- unique(stats::na.omit(data[[column_name]]))
  labels <- isco08_to_label(values)
  values_named <- setNames(values, labels)
  for (i in seq_along(values_named)) {
    eval.parent(substitute(
      update_value_label(
        data,
        column_name,
        as.numeric(values_named[i]),
        names(values_named)[i]
      )
    ))
  }
}
```

```{r}
#| label: update-occupational-classifications
#| code-fold: show
#| column: body-outset-right

# set partner occupation to NA for respondents without partners
data_cleaned$isco08r_partner[data_cleaned$partner == 0] <- NA

# update value labels for ISCO-08 classifications
update_isco08r("isco08r_partner", data_cleaned)
update_value_label(data_cleaned, "isco08r_partner", NA, "No partner")
update_isco08r("isco08r", data_cleaned)
update_isco08r("isco08r_father", data_cleaned)
update_isco08r("isco08r_mother", data_cleaned)

# update value labels for BRC14 classifications
brc14_map_data <- brc14_map()
update_brc14("brc14", data_cleaned, brc14_map_data)
update_brc14("brc14_mother", data_cleaned, brc14_map_data)
update_brc14("brc14_father", data_cleaned, brc14_map_data)
update_brc14("brc14_partner", data_cleaned, brc14_map_data)

# set partner BRC14 and ISEI-08 to NA for respondents without partners
data_cleaned$brc14_partner[data_cleaned$partner == 0] <- NA
update_value_label(data_cleaned, "brc14_partner", NA, "No partner")
data_cleaned$isei08_partner[data_cleaned$partner == 0] <- NA
```

## Assess Missing Data Patterns

Little's MCAR test diagnoses whether missingness deviates from completely random.

```{r}
#| label: non-deterministic-data
#| code-fold: show
#| column: body-outset-right

# select variables with non-deterministic missingness for MCAR test
non_deterministic_data <- data_cleaned %>%
  dplyr::select(where(~ sum(is.na(.)) > 0 & sum(is.na(.)) < nrow(data_cleaned)))

if (ncol(non_deterministic_data) > 1) {
  # perform MCAR test with error handling
  mcar_test_result <- tryCatch(
    naniar::mcar_test(non_deterministic_data),
    error = function(e) {
      message("Error in MCAR test: ", e$message)
      return(NULL)
    }
  )
  if (!is.null(mcar_test_result)) {
    mcar_test_result %<>%
      dplyr::mutate(p.value = format_p_value(p.value))
  } else {
    message("MCAR test could not be completed. Review the missing data structure.")
  }
} else {
  message("Insufficient variability in missing data patterns for MCAR test.")
}
```


```{r}
#| label: tbl-mcar-test
#| tbl-cap: "Results of Little's MCAR Tests"
#| code-fold: true
#| column: body-outset-right

col_names <- c("$\\chi^2$", "$df$", "$p$", "Missing Patterns$^{1}$")

footnote_text <- glue(
  "$n={nrow(data_cleaned)}$;",
  "a $p$-value $\\leq$ 0.001 indicates that data are not missing completely at random;",
  "$^{1}$ Represents the number of missing data patterns."
)

mcar_test_result %>%
  as_kable(align = "lllr") %>%
  footnote() %>%
  add_style()
```

### Exclude Cases with Critical Missing Data

```{r}
#| label: handle-missing-dhd-values
#| code-fold: show
#| column: body-outset-right

# identify cases missing both outcome and key predictor
cases_with_missing_dhd_values <- which(
  is.na(data_cleaned$dhd_index) & is.na(data_cleaned$dhd_kcal)
)

percentage_missing <- length(cases_with_missing_dhd_values) / nrow(data_cleaned)

# report exclusion statistics
sentence_missing_dhd <- glue(
  "Out of the total {nrow(data_cleaned)} respondents, ",
  "{length(cases_with_missing_dhd_values)} had ",
  "missing values for both the DHD index and DHD ",
  "kcal ({round(percentage_missing * 100, 2)}%). ",
  "These cases were excluded from further analysis."
)

htmltools::p(sentence_missing_dhd)
```

### Summarize Remaining Missing Data

```{r}
#| label: tbl-proportion-missing-values
#| tbl-cap: "Overview of Missing Values After Excluding Cases Without DHD Values"
#| code-fold: show
#| column: body-outset-right

# helper function to summarize missingness distribution
summarize_missingness <- function(x) {
  counts <- as.numeric(x)
  props <- paste0(round(as.numeric(names(x)) * 100, 2), "%")
  parts <- paste0(counts[-1], " (", props[-1], ")")
  sentence <- glue(
    "Out of the total {sum(counts)} ",
    "respondents, {counts[1]} had no missing values, ",
    "followed by {paste(parts, collapse = ', ')} ",
    "with increasing proportions of missing values."
  )
  htmltools::HTML(as.character(htmltools::p(sentence)))
}

# subset data excluding cases with missing DHD values
data_subset <- if (length(cases_with_missing_dhd_values) > 0) {
  data_cleaned[-cases_with_missing_dhd_values, ]
} else {
  data_cleaned
}

# define table display parameters
col_names <- c("Variable", rep(c(
  "$n_{\\text{miss}}$",
  "$\\%_{\\text{miss}}$"
), 3))

footnote_text <- glue(
  "Missing values include both system-defined and ",
  "user-defined values, ordered by highest missingness."
)

# summarize missingness stratified by partner status
data_subset %>%
  dplyr::select(-idnummer) %$%
  list(
    No_Partner = dplyr::filter(., partner == 0),
    Partner = dplyr::filter(., partner == 1), .
  ) %>%
  purrr::map2(.y = names(.), ~ summarize_na(.x) %>%
    setNames(c("var", paste0(names(.)[-1], "_", .y)))) %>%
  Reduce(f = function(x, y) dplyr::full_join(x, y, by = "var")) %>%
  dplyr::mutate(var = beautify_names(var)) %>%
  as_kable() %>%
  kableExtra::add_header_above(
    c(" " = 1, "No Partner" = 2, "Partner" = 2, "Total" = 2)
  ) %>%
  footnote() %>%
  add_style()
```

### Compute Proportional Missingness

```{r}
#| label: prop-missing-values
#| code-fold: show
#| column: body-outset-right

# calculate proportion of missing values for key analysis variables
# the reconciled score and status columns stay out of the ratio: siops08 and
# egp11 would triple-count the occupational missingness already carried by
# isei08, and lf_status is character with no NA; excluding them keeps the
# 20% rule identical to the pre-reconciliation pipeline
data_subset$prop_miss <- data_subset %>%
  dplyr::select(
    dplyr::matches("_(par|mot|fat)|^dhd|dif|inc|edu|isei", ignore.case = TRUE)
  ) %>%
  dplyr::select(-dplyr::matches("brc14|isco08|siops08|egp11|lf_status")) %>%
  asplit(1) %>%
  sapply(function(x) mean(is.na(x)))

# tabulate proportions of missing values
prop_miss_table <- table(data_subset$prop_miss)

# display summary of missingness proportions
summarize_missingness(prop_miss_table)

# determine cutoff for imputation based on missingness proportion (>20%)
# note: structural partner NAs count toward prop_miss, so non-partnered
# respondents tolerate fewer additional missing values than partnered
# respondents before exclusion
cut_off <- min(unique(data_subset$prop_miss[data_subset$prop_miss > 0.2]))

# filter data for imputation, excluding high missingness cases
data_for_imputation <- data_subset %>%
  dplyr::filter(prop_miss < cut_off) %>%
  dplyr::select(-prop_miss)
```

## Generate Descriptive Statistics

```{r}
#| label: descriptive-stats-col-names
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "$\\ $", "$n$", "$\\bar{x}$", "$\\text{Min.}$", "$\\text{Max.}$",
  "$SD$", "$\\text{Skewness}$", "$\\text{Kurtosis}$", "$SE$",
  "$\\text{Normality}^{\\dagger}$"
)

footnote_text <- glue(
  "Shapiro-Wilk Test$^{{\\dagger}}$;",
  "values are normality test statistics (Shapiro-Wilk $W$ for $n \\leq 5000$, Lilliefors $D$ otherwise); ",
  "smaller $W$ (larger $D$) suggests stronger departure from normality;",
  "'&#x2014;' indicates variables with $\\leq 4$ unique values (normality test not applicable)"
)
```

```{r}
#| label: tbl-descriptive-stats-before-outlier-removal
#| tbl-cap: "Descriptive Statistics Before Outlier Removal"
#| code-fold: show
#| column: body-outset-right

# generate descriptive statistics for imputation dataset
data_for_imputation %>%
  dplyr::select(-idnummer) %>%
  setNames(beautify_names(names(.))) %>%
  describe(
    normality = TRUE,
    print_only = FALSE
  ) %>%
  dplyr::rowwise() %>%
  dplyr::ungroup() %>%
  dplyr::select(-p) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

## Evaluate SES Composite Scale Validity {#sec-ses-validity}

Compares composite and domain-specific operationalizations across reliability, factor structure, and discriminant validity criteria.

### Prepare SES Indicators

```{r}
#| label: ses-indicator-preparation
#| code-fold: show
#| column: body-outset-right

# extract SES indicators for psychometric evaluation
ses_indicators <- data_for_imputation %>%
  dplyr::select(
    edu_father = education_father,
    edu_mother = education_mother,
    occ_father = isei08_father,
    occ_mother = isei08_mother,
    edu_partner = education_partner,
    occ_partner = isei08_partner,
    partner
  )

# create analysis subsets for different scale configurations
parental_indicators <- ses_indicators %>%
  dplyr::select(edu_father, edu_mother, occ_father, occ_mother)

# partner indicators restricted to partnered respondents
partner_indicators <- ses_indicators %>%
  dplyr::filter(partner == 1) %>%
  dplyr::select(edu_partner, occ_partner)

# combined indicators for full model testing (partnered respondents only)
combined_indicators <- ses_indicators %>%
  dplyr::filter(partner == 1) %>%
  dplyr::select(-partner)
```

```{r}
#| label: tbl-ses-sample-sizes
#| tbl-cap: "Sample Sizes for SES Scale Validation"
#| code-fold: true
#| column: body-outset-right

sample_info <- tibble::tibble(
  Subset = c("Total sample", "Partnered respondents", "Parental complete cases"),
  n = c(
    nrow(ses_indicators),
    sum(ses_indicators$partner == 1, na.rm = TRUE),
    sum(stats::complete.cases(parental_indicators))
  )
)

footnote_text <- "Sample sizes for validity and reliability analyses."

sample_info %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Descriptive Statistics of SES Indicators

```{r}
#| label: ses-descriptives-tbl-labels
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "Indicator", "$n$", "$n_{miss}$", "$\\%_{miss}$",
  "$\\bar{x}$", "$SD$", "Min", "Max", "Skew", "Kurt", "Domain"
)

footnote_text <- glue(
  "$N = {nrow(parental_indicators)}$; ",
  "Education = educational attainment (ordinal, 0-6); ",
  "Occupation = ISEI-08 occupational status (continuous, 16-90)."
)
```

```{r}
#| label: tbl-ses-descriptives
#| tbl-cap: "Descriptive Statistics of Parental SES Indicators"
#| code-fold: show
#| column: body-outset-right

# function to compute descriptive statistics for SES indicators
compute_ses_descriptives <- function(data) {
  data %>%
    tidyr::pivot_longer(
      cols = dplyr::everything(),
      names_to = "indicator",
      values_to = "value"
    ) %>%
    dplyr::group_by(indicator) %>%
    dplyr::summarise(
      n = sum(!is.na(value)),
      n_miss = sum(is.na(value)),
      pct_miss = round(100 * mean(is.na(value)), 1),
      mean = round(mean(value, na.rm = TRUE), 2),
      sd = round(sd(value, na.rm = TRUE), 2),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      skew = round(moments::skewness(value, na.rm = TRUE), 2),
      kurt = round(moments::kurtosis(value, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      domain = dplyr::case_when(
        grepl("edu", indicator) ~ "Education",
        grepl("occ", indicator) ~ "Occupation",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::arrange(domain, indicator)
}

compute_ses_descriptives(parental_indicators) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Correlation Structure

```{r}
#| label: tbl-ses-correlations
#| tbl-cap: "Correlation Matrix of Parental SES Indicators"
#| code-fold: show
#| column: body-outset-right

cor_parental <- cor(parental_indicators, use = "pairwise.complete.obs")

footnote_text <- glue(
  "Pearson correlations with pairwise deletion; ",
  "within-domain correlations appear in diagonal quadrants; ",
  "cross-domain correlations indicate degree of construct overlap."
)

as.data.frame(round(cor_parental, 3)) %>%
  tibble::rownames_to_column("Indicator") %>%
  as_kable(align = "lrrrr") %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: tbl-correlation-diagnostics
#| tbl-cap: "Correlation Pattern Diagnostics"
#| code-fold: show
#| column: body-outset-right

# extract correlation patterns for domain diagnostics
within_edu <- cor_parental["edu_father", "edu_mother"]
within_occ <- cor_parental["occ_father", "occ_mother"]
cross_domain <- mean(c(
  cor_parental["edu_father", "occ_father"],
  cor_parental["edu_father", "occ_mother"],
  cor_parental["edu_mother", "occ_father"],
  cor_parental["edu_mother", "occ_mother"]
))

# determine if cross-domain correlations are substantially lower
cross_lower_than_within <- cross_domain < min(within_edu, within_occ) - 0.10

cor_diagnostics <- tibble::tibble(
  Pattern = c(
    "Within-Education (Father \u2194 Mother)",
    "Within-Occupation (Father \u2194 Mother)",
    "Cross-Domain (Education \u2194 Occupation)"
  ),
  r = round(c(within_edu, within_occ, cross_domain), 3),
  Interpretation = c(
    "Parental homogamy in education",
    "Parental homogamy in occupation",
    dplyr::if_else(cross_lower_than_within,
      "Lower than within-domain (supports distinction)",
      "Comparable to within-domain (suggests overlap)"
    )
  )
)

# conditional footnote based on observed correlation pattern
footnote_text <- dplyr::if_else(
  cross_lower_than_within,
  glue(
    "Cross-domain correlations are substantially lower than within-domain ",
    "correlations, suggesting education and occupation tap distinct constructs."
  ),
  glue(
    "Cross-domain correlations are comparable to or exceed within-domain ",
    "correlations, suggesting substantial construct overlap."
  )
)

cor_diagnostics %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Reliability Analysis

```{r}
#| label: reliability-computation
#| code-fold: show
#| column: body-outset-right

# complete cases for reliability analysis
parental_complete <- parental_indicators %>%
  tidyr::drop_na()

# cronbach's alpha for parental SES composite (4 indicators)
alpha_parental <- psych::alpha(
  parental_complete,
  check.keys = TRUE,
  warnings = FALSE
)

# mcdonald's omega (hierarchical) with error handling
omega_parental <- tryCatch(
  {
    psych::omega(parental_complete, nfactors = 1, fm = "ml", plot = FALSE)
  },
  error = function(e) NULL
)

# spearman-brown for 2-item education scale
parental_edu_only <- parental_indicators %>%
  dplyr::select(edu_father, edu_mother) %>%
  tidyr::drop_na()

r_edu <- cor(parental_edu_only$edu_father, parental_edu_only$edu_mother)
sb_edu <- (2 * r_edu) / (1 + r_edu)

# spearman-brown for 2-item occupation scale
parental_occ_only <- parental_indicators %>%
  dplyr::select(occ_father, occ_mother) %>%
  tidyr::drop_na()

r_occ <- cor(parental_occ_only$occ_father, parental_occ_only$occ_mother)
sb_occ <- (2 * r_occ) / (1 + r_occ)

# partner SES reliability (2 items)
partner_complete <- partner_indicators %>%
  tidyr::drop_na()

r_partner <- cor(partner_complete$edu_partner, partner_complete$occ_partner)
sb_partner <- (2 * r_partner) / (1 + r_partner)
```

```{r}
#| label: tbl-reliability-composite
#| tbl-cap: "Reliability of Parental SES Composite Scale (4 Indicators)"
#| code-fold: true
#| column: body-outset-right

col_names <- c("Statistic", "Value", "Threshold", "Met")

reliability_composite <- data.frame(
  Statistic = c(
    "Cronbach's $\\alpha$ (standardized)",
    "Cronbach's $\\alpha$ (raw)",
    "McDonald's $\\omega$ total",
    "Average inter-item $r$",
    "Complete cases"
  ),
  Value = c(
    round(alpha_parental$total$std.alpha, 3),
    round(alpha_parental$total$raw_alpha, 3),
    if (!is.null(omega_parental)) round(omega_parental$omega.tot, 3) else NA,
    round(alpha_parental$total$average_r, 3),
    nrow(parental_complete)
  ),
  Threshold = c("$\\geq 0.70$", "$\\geq 0.70$", "$\\geq 0.70$", "$\\geq 0.30$", "\u2014"),
  Met = c(
    if (alpha_parental$total$std.alpha >= 0.70) "$\\checkmark$" else "$\\times$",
    if (alpha_parental$total$raw_alpha >= 0.70) "$\\checkmark$" else "$\\times$",
    if (!is.null(omega_parental) && omega_parental$omega.tot >= 0.70) "$\\checkmark$" else "$\\times$",
    if (alpha_parental$total$average_r >= 0.30) "$\\checkmark$" else "$\\times$",
    "\u2014"
  )
)

reliability_composite$Statistic %<>% md_to_html()
reliability_composite$Threshold %<>% md_to_html()
reliability_composite$Met %<>% md_to_html()

footnote_text <- glue(
  "Reliability thresholds based on Nunnally and Bernstein (1994); ",
  "$\\alpha \\geq 0.70$ indicates acceptable internal consistency."
)

reliability_composite %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: tbl-item-statistics
#| tbl-cap: "Item-Total Statistics for Parental SES Composite"
#| code-fold: show
#| column: body-outset-right

col_names <- c("Indicator", "$r_{\\text{drop}}$", "$\\alpha_{\\text{if dropped}}$")

item_stats <- data.frame(
  indicator = rownames(alpha_parental$item.stats),
  r_drop = round(alpha_parental$item.stats$r.drop, 3),
  alpha_drop = round(alpha_parental$alpha.drop$std.alpha, 3)
)

footnote_text <- glue(
  "$r_{{drop}}$ = corrected item-total correlation; ",
  "$\\alpha_{{if dropped}}$ = scale reliability if item removed; ",
  "items with $r_{{drop}} < 0.30$ may not belong to the construct."
)

item_stats %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: tbl-reliability-summary
#| tbl-cap: "Reliability Summary Across Scale Specifications"
#| code-fold: show
#| column: body-outset-right

col_names <- c(
  "Scale", "$k$", "$n$", "$\\alpha$ / SB",
  "Avg. $r$", "Threshold Met"
)

reliability_summary <- data.frame(
  Scale = c(
    "Parental SES (composite)",
    "Parental Education",
    "Parental Occupation",
    "Partner SES (composite)"
  ),
  k = c(4, 2, 2, 2),
  n = c(
    nrow(parental_complete),
    nrow(parental_edu_only),
    nrow(parental_occ_only),
    nrow(partner_complete)
  ),
  alpha = round(c(
    alpha_parental$total$std.alpha,
    sb_edu,
    sb_occ,
    sb_partner
  ), 3),
  avg_r = round(c(
    alpha_parental$total$average_r,
    r_edu,
    r_occ,
    r_partner
  ), 3),
  met = c(
    if (alpha_parental$total$std.alpha >= 0.70) "$\\checkmark$" else "$\\times$",
    if (sb_edu >= 0.70) "$\\checkmark$" else "$\\times$",
    if (sb_occ >= 0.70) "$\\checkmark$" else "$\\times$",
    if (sb_partner >= 0.70) "$\\checkmark$" else "$\\times$"
  )
)

reliability_summary$met %<>% md_to_html()

footnote_text <- glue(
  "$k$ = number of indicators; ",
  "for 2-item scales, Spearman-Brown coefficient is reported instead of $\\alpha$; ",
  "threshold: $\\alpha \\geq 0.70$ (Nunnally and Bernstein 1994)."
)

reliability_summary %>%
  as_kable(escape = FALSE) %>%
  footnote() %>%
  add_style()
```

### Confirmatory Factor Analysis

```{r}
#| label: cfa-model-specifications
#| code-fold: show
#| column: body-outset-right

# model 1: single-factor composite
model_1_composite <- "
  parental_ses =~ edu_father + edu_mother + occ_father + occ_mother
"

# model 2: two-factor domain-specific
model_2_domains <- "
  education_parental =~ edu_father + edu_mother
  isei08_parental =~ occ_father + occ_mother
  education_parental ~~ isei08_parental
"
```

```{r}
#| label: cfa-estimation
#| code-fold: show
#| column: body-outset-right

# fit model 1: single factor
fit_m1 <- lavaan::cfa(
  model_1_composite,
  data = parental_indicators,
  estimator = "MLR",
  missing = "fiml",
  std.lv = TRUE
)

# fit model 2: two factors
fit_m2 <- lavaan::cfa(
  model_2_domains,
  data = parental_indicators,
  estimator = "MLR",
  missing = "fiml",
  std.lv = TRUE
)
```

```{r}
#| label: cfa-fit-extraction
#| code-fold: show
#| column: body-outset-right

# function to extract fit indices into standardized format
extract_fit <- function(fit, name) {
  if (is.null(fit)) {
    return(tibble::tibble(
      Model = name, chi_sq = NA, df = NA, p = NA,
      CFI = NA, TLI = NA, RMSEA = NA, SRMR = NA, AIC = NA
    ))
  }
  fm <- lavaan::fitMeasures(
    fit, c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", "srmr", "aic")
  )
  tibble::tibble(
    Model = name,
    chi_sq = round(fm["chisq"], 2),
    df = fm["df"],
    p = round(fm["pvalue"], 4),
    CFI = round(fm["cfi"], 3),
    TLI = round(fm["tli"], 3),
    RMSEA = round(fm["rmsea"], 3),
    SRMR = round(fm["srmr"], 3),
    AIC = round(fm["aic"], 1)
  )
}

# extract fit indices for model evaluation
fit_m1_cfi <- lavaan::fitMeasures(fit_m1, "cfi")
fit_m1_rmsea <- lavaan::fitMeasures(fit_m1, "rmsea")
fit_m2_cfi <- lavaan::fitMeasures(fit_m2, "cfi")
fit_m2_rmsea <- lavaan::fitMeasures(fit_m2, "rmsea")

# define fit acceptability thresholds
cfi_acceptable <- 0.90
rmsea_acceptable <- 0.08

# evaluate model fit against thresholds
m1_fit_acceptable <- fit_m1_cfi >= cfi_acceptable & fit_m1_rmsea <= rmsea_acceptable
m2_fit_acceptable <- fit_m2_cfi >= cfi_acceptable & fit_m2_rmsea <= rmsea_acceptable
any_fit_acceptable <- m1_fit_acceptable | m2_fit_acceptable
```

```{r}
#| label: tbl-cfa-fit-comparison
#| tbl-cap: "CFA Model Fit Comparison"
#| code-fold: show
#| column: body-outset-right

fit_comparison <- dplyr::bind_rows(
  extract_fit(fit_m1, "1. Single-Factor (Composite)"),
  extract_fit(fit_m2, "2. Two-Factor (Domains)")
)

col_names <- c(
  "Model", "$\\chi^2$", "$df$", "$p$",
  "CFI", "TLI", "RMSEA", "SRMR", "AIC"
)

footnote_text <- glue(
  "Estimation: MLR (robust to non-normality); ",
  "missing data: FIML; ",
  "fit thresholds: CFI/TLI $\\geq$ 0.95 (good), $\\geq$ 0.90 (acceptable); ",
  "RMSEA $\\leq$ 0.06 (good), $\\leq$ 0.08 (acceptable)."
)

fit_comparison %>%
  as_kable(escape = FALSE) %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: model-fit-warning
#| results: asis
#| code-fold: true
#| column: body-outset-right

# generate warning callout if model fit is unacceptable
if (!any_fit_acceptable) {
  warning_html <- glue(
    "<div class='callout-warning'>
    <strong>Model Fit Warning</strong><br><br>
    Neither measurement model achieves acceptable fit (CFI $\\geq$ 0.90, ",
    "RMSEA $\\leq$ 0.08). The single-factor model shows CFI = {round(fit_m1_cfi, 3)} ",
    "and RMSEA = {round(fit_m1_rmsea, 3)}; the two-factor model shows ",
    "CFI = {round(fit_m2_cfi, 3)} and RMSEA = {round(fit_m2_rmsea, 3)}. ",
    "Observed variable composites may be more appropriate than latent factor scores.
    </div>"
  )
  htmltools::HTML(warning_html)
}
```

```{r}
#| label: tbl-chi-diff-test
#| tbl-cap: "Chi-Square Difference Test: Single vs. Two-Factor Model"
#| code-fold: show
#| column: body-outset-right

# nested model comparison via chi-square difference test
chi_diff <- lavaan::anova(fit_m1, fit_m2)

chi_diff_tbl <- tibble::tibble(
  Comparison = "Model 2 vs. Model 1",
  Delta_chi_sq = round(chi_diff$`Chisq diff`[2], 2),
  Delta_df = chi_diff$`Df diff`[2],
  p = round(chi_diff$`Pr(>Chisq)`[2], 4)
)

footnote_text <- glue(
  "A significant $\\chi^2$ difference ($p < .05$) indicates the two-factor ",
  "model fits significantly better. With df = 1, interpret cautiously."
)

chi_diff_tbl %>%
  as_kable(escape = FALSE) %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: tbl-cfa-loadings
#| tbl-cap: "Standardized Factor Loadings"
#| code-fold: show
#| column: body-outset-right

# extract and combine loadings from both models
loadings_m1 <- lavaan::standardizedSolution(fit_m1) %>%
  dplyr::filter(op == "=~") %>%
  dplyr::transmute(
    Model = "Single-Factor",
    Factor = lhs,
    Indicator = rhs,
    Loading = round(est.std, 3),
    SE = round(se, 3),
    p = pvalue
  )

loadings_m2 <- lavaan::standardizedSolution(fit_m2) %>%
  dplyr::filter(op == "=~") %>%
  dplyr::transmute(
    Model = "Two-Factor",
    Factor = lhs,
    Indicator = rhs,
    Loading = round(est.std, 3),
    SE = round(se, 3),
    p = pvalue
  )

loadings_combined <- dplyr::bind_rows(loadings_m1, loadings_m2) %>%
  dplyr::mutate(
    Sig = dplyr::case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  dplyr::select(-p)

col_names <- c("Model", "Factor", "Indicator", "$\\lambda$", "$SE$", "Sig.")

footnote_text <- "*** $p < .001$; ** $p < .01$; * $p < .05$."

loadings_combined %>%
  as_kable(escape = FALSE) %>%
  footnote() %>%
  add_style()
```

### Convergent and Discriminant Validity

```{r}
#| label: validity-computation
#| code-fold: show
#| column: body-outset-right

# function to compute average variance extracted (AVE)
compute_ave <- function(fit, factor_name) {
  loadings <- lavaan::standardizedSolution(fit) %>%
    dplyr::filter(op == "=~", lhs == factor_name) %>%
    dplyr::pull(est.std)
  mean(loadings^2)
}

# function to compute composite reliability (CR)
compute_cr <- function(fit, factor_name) {
  loadings <- lavaan::standardizedSolution(fit) %>%
    dplyr::filter(op == "=~", lhs == factor_name) %>%
    dplyr::pull(est.std)
  sum_l <- sum(loadings)
  sum_e <- sum(1 - loadings^2)
  (sum_l^2) / (sum_l^2 + sum_e)
}

# single-factor model metrics
ave_composite <- compute_ave(fit_m1, "parental_ses")
cr_composite <- compute_cr(fit_m1, "parental_ses")

# two-factor model metrics
ave_edu <- compute_ave(fit_m2, "education_parental")
ave_occ <- compute_ave(fit_m2, "isei08_parental")
cr_edu <- compute_cr(fit_m2, "education_parental")
cr_occ <- compute_cr(fit_m2, "isei08_parental")

# factor correlation from two-factor model
factor_cor <- lavaan::standardizedSolution(fit_m2) %>%
  dplyr::filter(op == "~~", lhs == "education_parental", rhs == "isei08_parental") %>%
  dplyr::pull(est.std)

# fornell-larcker criterion: squared correlation < AVE for each factor
fl_edu <- factor_cor^2 < ave_edu
fl_occ <- factor_cor^2 < ave_occ

# htmt ratio using observed correlations
htmt_num <- mean(c(
  cor_parental["edu_father", "occ_father"],
  cor_parental["edu_father", "occ_mother"],
  cor_parental["edu_mother", "occ_father"],
  cor_parental["edu_mother", "occ_mother"]
))
htmt_denom <- sqrt(
  cor_parental["edu_father", "edu_mother"] *
    cor_parental["occ_father", "occ_mother"]
)
htmt <- htmt_num / htmt_denom

# flag extreme factor correlation (potential model identification issue)
factor_cor_extreme <- abs(factor_cor) > 0.95
```

```{r}
#| label: factor-correlation-warning
#| results: asis
#| code-fold: true
#| column: body-outset-right

# generate warning callout if factor correlation is extreme
if (factor_cor_extreme) {
  fc_warning <- glue(
    "<div class='callout-warning'>
    <strong>Factor Correlation Warning</strong><br><br>
    The factor correlation between parental education and parental occupation ",
    "is extremely high ($\\phi$ = {round(factor_cor, 3)}), approaching unity. ",
    "With only two indicators per factor, discriminant validity tests are unreliable.
    </div>"
  )
  htmltools::HTML(fc_warning)
}
```

```{r}
#| label: tbl-validity-metrics
#| tbl-cap: "Convergent Validity: AVE and Composite Reliability"
#| code-fold: show
#| column: body-outset-right

col_names <- c("Model", "Factor", "AVE", "CR", "AVE Met", "CR Met")

validity_metrics <- data.frame(
  Model = c("Single-Factor", "Two-Factor", "Two-Factor"),
  Factor = c("Parental SES", "Parental Education", "Parental Occupation"),
  AVE = round(c(ave_composite, ave_edu, ave_occ), 3),
  CR = round(c(cr_composite, cr_edu, cr_occ), 3),
  AVE_met = c(
    if (ave_composite >= 0.50) "$\\checkmark$" else "$\\times$",
    if (ave_edu >= 0.50) "$\\checkmark$" else "$\\times$",
    if (ave_occ >= 0.50) "$\\checkmark$" else "$\\times$"
  ),
  CR_met = c(
    if (cr_composite >= 0.70) "$\\checkmark$" else "$\\times$",
    if (cr_edu >= 0.70) "$\\checkmark$" else "$\\times$",
    if (cr_occ >= 0.70) "$\\checkmark$" else "$\\times$"
  )
)

validity_metrics$AVE_met %<>% md_to_html()
validity_metrics$CR_met %<>% md_to_html()

footnote_text <- glue(
  "AVE = Average Variance Extracted (threshold $\\geq 0.50$); ",
  "CR = Composite Reliability (threshold $\\geq 0.70$) (Fornell and Larcker 1981)."
)

validity_metrics %>%
  as_kable(escape = FALSE) %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: discriminant-validity-attrs
#| code-fold: true
#| column: body-outset-right

col_names <- c("Criterion", "Value", "Threshold", "Conclusion")

footnote_text <- glue(
  "Fornell-Larcker: discriminant validity established when $\\phi^2 < AVE$ (Fornell and Larcker 1981); ",
  "HTMT $< 0.85$ indicates discriminant validity (Henseler, Ringle, and Sarstedt 2015); ",
  "'Unreliable' indicates model identification issues."
)
```

```{r}
#| label: tbl-discriminant-validity
#| tbl-cap: "Discriminant Validity Assessment"
#| code-fold: show
#| column: body-outset-right

# determine conclusions with appropriate caveats for unreliable estimates
fl_edu_conclusion <- dplyr::case_when(
  factor_cor_extreme ~ "Unreliable",
  fl_edu ~ "Distinct",
  TRUE ~ "Overlap"
)
fl_occ_conclusion <- dplyr::case_when(
  factor_cor_extreme ~ "Unreliable",
  fl_occ ~ "Distinct",
  TRUE ~ "Overlap"
)
htmt_conclusion <- if (htmt < 0.85) "Distinct" else "Overlap"

discriminant_tbl <- data.frame(
  Criterion = c(
    "Factor correlation ($\\phi$)",
    "Squared correlation ($\\phi^2$)",
    "AVE (Education)",
    "AVE (Occupation)",
    "Fornell-Larcker: $\\phi^2 < AVE_{Edu}$",
    "Fornell-Larcker: $\\phi^2 < AVE_{Occ}$",
    "HTMT ratio"
  ),
  Value = c(
    round(factor_cor, 3),
    round(factor_cor^2, 3),
    round(ave_edu, 3),
    round(ave_occ, 3),
    if (fl_edu) "$\\checkmark$" else "$\\times$",
    if (fl_occ) "$\\checkmark$" else "$\\times$",
    round(htmt, 3)
  ),
  Threshold = c(
    "\u2014", "\u2014", "$\\geq 0.50$", "$\\geq 0.50$",
    "$\\phi^2 < AVE$", "$\\phi^2 < AVE$", "$< 0.85$"
  ),
  Conclusion = c(
    if (factor_cor_extreme) "Extreme (see warning)" else "\u2014",
    "\u2014", "\u2014", "\u2014",
    fl_edu_conclusion,
    fl_occ_conclusion,
    htmt_conclusion
  )
)

discriminant_tbl$Criterion %<>% md_to_html()
discriminant_tbl$Value %<>% md_to_html()
discriminant_tbl$Threshold %<>% md_to_html()

discriminant_tbl %>%
  as_kable(escape = FALSE) %>%
  footnote() %>%
  add_style()
```

### Synthesis and Recommendation

```{r}
#| label: validity-synthesis-attrs
#| code-fold: true
#| column: body-outset-right

col_names <- c("Criterion", "Composite SES", "Domain-Specific", "Threshold")

footnote_text <- glue(
  "FL = Fornell-Larcker criterion; ",
  "HTMT = Heterotrait-Monotrait ratio; ",
  "domain-specific values reported as Education / Occupation."
)
```


```{r}
#| label: tbl-validity-synthesis
#| tbl-cap: "Summary of Validity and Reliability Evidence"
#| results: asis
#| code-fold: show
#| column: body-outset-right

synthesis_tbl <- data.frame(
  Criterion = c(
    "Internal Consistency ($\\alpha$)",
    "Average Variance Extracted",
    "Composite Reliability",
    "Discriminant Validity (FL)",
    "Discriminant Validity (HTMT)",
    "CFA Model Fit (CFI)",
    "CFA Model Fit (RMSEA)"
  ),
  Composite = c(
    round(alpha_parental$total$std.alpha, 3),
    round(ave_composite, 3),
    round(cr_composite, 3),
    "N/A",
    "N/A",
    round(fit_m1_cfi, 3),
    round(fit_m1_rmsea, 3)
  ),
  Domain = c(
    paste0(round(sb_edu, 3), " / ", round(sb_occ, 3)),
    paste0(round(ave_edu, 3), " / ", round(ave_occ, 3)),
    paste0(round(cr_edu, 3), " / ", round(cr_occ, 3)),
    if (fl_edu && fl_occ && !factor_cor_extreme) "Supported" else "Not Supported",
    if (htmt < 0.85) "Supported" else "Not Supported",
    round(fit_m2_cfi, 3),
    round(fit_m2_rmsea, 3)
  ),
  Threshold = c(
    "$\\geq 0.70$",
    "$\\geq 0.50$",
    "$\\geq 0.70$",
    "$\\phi^2 < AVE$",
    "$< 0.85$",
    "$\\geq 0.90$",
    "$\\leq 0.08$"
  )
)

synthesis_tbl$Criterion %<>% md_to_html()
synthesis_tbl$Threshold %<>% md_to_html()

synthesis_tbl %>%
  as_kable(escape = FALSE) %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: ses-recommendation
#| results: asis
#| code-fold: true
#| column: body-outset-right

# comprehensive decision logic incorporating model fit and validity evidence
alpha_ok <- alpha_parental$total$std.alpha >= 0.70
ave_ok <- ave_composite >= 0.50
any_fit_acceptable <- m1_fit_acceptable || m2_fit_acceptable
discriminant_not_established <- !(fl_edu && fl_occ) | factor_cor_extreme

# determine recommendation based on full evidence pattern
if (!any_fit_acceptable) {
  decision <- glue(
    "<div class='callout-important'>
    <strong>Recommendation:</strong> Latent variable modeling is not supported.<br><br>
    Neither measurement model achieves acceptable fit ",
    "(single-factor: CFI = {round(fit_m1_cfi, 3)}, RMSEA = {round(fit_m1_rmsea, 3)}; ",
    "two-factor: CFI = {round(fit_m2_cfi, 3)}, RMSEA = {round(fit_m2_rmsea, 3)}). ",
    "Use observed variable composites:<br>
    <ul>
    <li><code>education_parental</code> = mean of father's and mother's education</li>
    <li><code>isei08_parental</code> = mean of father's and mother's ISEI-08</li>
    </ul>
    </div>"
  )
} else {
  decision <- if (alpha_ok && ave_ok && discriminant_not_established) {
    "Composite SES operationalization is acceptable."
  } else if (fl_edu && fl_occ && !factor_cor_extreme && htmt < 0.85) {
    "Domain-specific operationalization preferred."
  } else {
    "Evidence is mixed."
  }

  decision <- glue(
    "<div class='callout-note'>
    <strong>Recommendation:</strong> {decision}
    </div>"
  )
}

htmltools::HTML(decision)
```

### Compute Composite SES Indicators

The composites computed here are provisional; step 6 recomputes them after imputation.

```{r}
#| label: compute-ses-composites-attrs
#| tbl-cap: "Summary of SES Indicators"
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "Indicator", "$n$", "$\\bar{x}$", "$SD$", "Min", "Max", "Definition"
)

footnote_text <- glue(
  "$N = {nrow(data_for_imputation)}$; ",
  "parental composites computed using complete-case row means (na.rm = FALSE); ",
  "partner indicators (`education_partner`, `isei08_partner`) set to NA for respondents without partners."
)
```

```{r}
#| label: compute-ses-composites
#| tbl-cap: "Summary of SES Indicators"
#| results: asis
#| code-fold: show
#| column: body-outset-right

# compute domain-specific parental SES composites
# naming follows dimension_referent pattern (e.g., education_parental, isei08_parental)
data_for_imputation <- data_for_imputation %>%
  dplyr::mutate(
    education_parental = rowMeans(
      dplyr::pick(education_father, education_mother),
      na.rm = FALSE
    ),
    isei08_parental = rowMeans(
      dplyr::pick(isei08_father, isei08_mother),
      na.rm = FALSE
    )
  )

# helper function to update variable labels
update_variable_label <- function(data, column_name, label) {
  labelled::var_label(data[[column_name]]) <- label
  data
}

# apply variable labels for documentation
data_for_imputation <- data_for_imputation %>%
  update_variable_label("education_parental", "Parental Education (mean of father and mother)") %>%
  update_variable_label("isei08_parental", "Parental Occupation (mean ISEI-08 of father and mother)")

# generate summary statistics for parental composites and partner indicators
ses_composite_summary <- data_for_imputation %>%
  dplyr::select(education_parental, isei08_parental, education_partner, isei08_partner) %>%
  tidyr::pivot_longer(
    cols = dplyr::everything(),
    names_to = "indicator",
    values_to = "value"
  ) %>%
  dplyr::group_by(indicator) %>%
  dplyr::summarise(
    n = sum(!is.na(value)),
    mean = round(mean(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    min = round(min(value, na.rm = TRUE), 2),
    max = round(max(value, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    definition = dplyr::case_when(
      indicator == "education_parental" ~ "Mean of father's and mother's education (0-6)",
      indicator == "isei08_parental" ~ "Mean of father's and mother's ISEI-08",
      indicator == "education_partner" ~ "Partner's educational attainment (0-6)",
      indicator == "isei08_partner" ~ "Partner's ISEI-08 occupational status",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::arrange(dplyr::desc(grepl("parental", indicator)), indicator)

ses_composite_summary %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

```{r}
#| label: save-cleaned-data
#| code-fold: true
#| column: body-outset-right

# save final cleaned dataset with SES composites for step 4
haven::write_sav(
  data_for_imputation,
  here::here("assets", "data", "saved", "data_for_imputation.sav")
)
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                                        | Content                                                |
| ------------------------------------------- | ------------------------------------------------------ |
| `assets/data/saved/data_for_imputation.sav` | Recoded data with SES composites, ready for imputation |

**Next:** [Step 4 (Outliers Detection)](/analysis/data/outliers/index.qmd)
