---
title: "ISCO-08 Standardization and Validation"
subtitle: "Mapping to BRC14 and ISEI-08 Classifications"
author: "Siard van den Bosch"
date: 2023-08-31
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 2};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Validate ISCO-08 codes; derive BRC14 and ISEI-08 mappings |
| **Inputs**     | `main_initial.sav`, `isco08_codes.xlsx`, `brc2014.xls` |
| **Outputs**    | `main_corrected.sav`, `brc14_map.R`, `isco08_to_brc14.R`, `isco08_to_isei08.R` |
| **Method**     | Manual code correction; ISCO-08 to BRC14/ISEI-08 crosswalk |
| **Requires**   | [Step 1 (Setup)](/analysis/setup/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

::: {.callout-note icon=false}
## Rationale

- ISEI-08 scoring: continuous occupational status measure with validated cross-national comparability [@ganzeboom2010]
:::

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear global env and force full garbage collection
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))

# detach non-base packages to avoid namespace interference
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions for text processing and data handling
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "sci_to_tex", "md_to_html", "set_globals", "decapitalize",
  "save_function", "add_style", "suppress_output", "import_pipes",
  "numeric_df", "as_kable", "empty_to_na", "drop_all_na_rows",
  "named_list", "std_text", "beautify_names", "footnote"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()
```


## Define Constants

Respondent-type suffixes govern column naming throughout the step.

```{r}
#| label: define-constants
#| code-fold: true
#| column: body-outset-right

# suffixes map respondent roles to column name patterns
# empty string for respondent self, underscore-prefixed for others
suffixes <- c("partner", "mother", "father") %>%
  setNames(object = sprintf("_%s", .)) %>%
  c("respondent" = "", .)

# occupation column names derived from suffixes
occupation_columns <- paste0("occupation", unname(suffixes))
```

## Load Data

```{r}
#| label: load-data
#| message: false
#| code-fold: show
#| column: body-outset-right

# load initial dataset and rename ISCO-08 columns for clarity
main <- haven::read_sav(here::here("assets", "data", "main_initial.sav")) %>%
  dplyr::rename(
    isco08_old = BRC14,
    isco08_old_father = BRC14_father,
    isco08_old_mother = BRC14_mother,
    isco08_old_partner = BRC14_partner
  )

# import corrected ISCO-08 codes from external source
xlsx <- readxl::read_xlsx(
  here::here("assets", "data", "isco08_codes.xlsx"),
  sheet = 1
)
```

## Clean and Standardize Data

- *in*: `main`
- *out*: `main_with_copies` with `_init` and `_cleaned` occupation columns
- *rule*: set two invalid entries (`idnummer` 2023, 1854) to `NA`; standardize text via `std_text()`

```{r}
#| label: data-cleaning
#| code-fold: show
#| column: body-outset-right

# replace invalid occupation entries with NA (case-specific corrections)
main[main$idnummer == 2023, "occupation_mother"] <- NA
main[main$idnummer == 1854, "occupation_partner"] <- NA

# generate initial and cleaned copies of occupation data
main_with_copies <- main %>%
  dplyr::mutate(dplyr::across(
    tidyselect::all_of(occupation_columns),
    list(
      init = ~.,
      cleaned = ~ std_text(.)
    ),
    .names = "{col}_{fn}"
  ))
```

## Define Classification Mapping Functions

Three functions: `brc14_map()` retrieves the CBS crosswalk, `isco08_to_brc14()` returns BRC14 group codes, and `isco08_to_isei08()` returns ISEI-08 scores.

```{r}
#| label: define-mapping-functions
#| code-fold: show
#| column: body-outset-right

# fetch BRC2014 mapping from local file or download if absent
brc14_map <- function() {
  local_file <- here::here("assets", "data", "brc2014.xls")
  if (!file.exists(local_file)) {
    download_url <- glue(
      "https://www.cbs.nl/-/media/imported/",
      "onze-diensten/methoden/classificaties/",
      "documents/2015/09/brc2014.xls"
    )
    download.file(download_url, local_file, mode = "wb")
  }
  m <- readxl::read_excel(local_file, sheet = 2)

  # code columns are zero-padded text in the workbook; parse to numeric so
  # armed-forces codes (e.g. '0110') match numeric isco08 values
  m$ISCO2008unitgroup <- as.numeric(m$ISCO2008unitgroup)
  m$BRC2014beroepsgroep <- as.numeric(m$BRC2014beroepsgroep)
  stopifnot(!anyNA(m$ISCO2008unitgroup), !anyNA(m$BRC2014beroepsgroep))
  m
}

# convert ISCO-08 codes to BRC14 classifications
isco08_to_brc14 <- function(x) {
  m <- brc14_map()
  purrr::map_dbl(x, ~ m$BRC2014beroepsgroep[
    match(.x, m$ISCO2008unitgroup, nomatch = NA)
  ])
}

# convert ISCO-08 codes to ISEI-08 scores with error handling
isco08_to_isei08 <- function(x) {
  sapply(x, function(code) {
    if (is.na(code)) {
      return(NA)
    }
    tryCatch(
      ISCO08ConveRsions::isco08toisei08(
        formatC(as.integer(code), width = 4, flag = "0", format = "d")
      ),
      error = function(e) NA
    )
  })
}
```

## Preserve Mapping Functions

```{r}
#| label: save-mapping-functions
#| message: false
#| code-fold: show
#| column: body-outset-right

# suppress output during function saving
suppress_output({
  save_function(brc14_map)
  save_function(isco08_to_brc14)
  save_function(isco08_to_isei08)
})
```

## Integrate Corrected ISCO Codes

```{r}
#| label: define-rename-function
#| code-fold: show
#| column: body-outset-right

# standardize column names for merging across respondent types
rename_and_select <- function(data, suffix = "") {
  require(rlang, quietly = TRUE)
  suffix <- if (suffix == "") suffix else paste0("_", suffix)

  out <- data %>%
    dplyr::rename(
      !!paste0("isco08_old", suffix) := isco08_init,
      !!paste0("occupation", if (suffix == "") "" else suffix, "_cleaned") := occupation,
      !!paste0("isco08", suffix) := isco08,
      !!paste0("isco08_new", suffix) := isco08_new
    ) %>%
    dplyr::select(-source)

  names(out) <- gsub("[_]+", "_", names(out))
  return(out)
}
```

- *in*: `xlsx`, `main_with_copies`
- *out*: `groups_with_classifications` (with BRC14 and ISEI-08 columns)
- *rule*: justified corrections (`isco08_new`) coalesce over the curator's final codes (`isco08`); initial codes the curator nullified remain `NA`; first non-missing value per `idnummer`

```{r}
#| label: integrate-corrected-codes
#| message: false
#| code-fold: show
#| column: body-outset-right

# select relevant columns from corrected data
corrected <- xlsx %>%
  dplyr::select(idnummer, source, occupation, isco08, isco08_init, isco08_new)

# curator code columns are stored as text cells in the workbook; parse to
# numeric and fail loudly if any non-empty value is not a clean code
parse_code <- function(x) {
  chr <- dplyr::na_if(trimws(as.character(x)), "")
  out <- suppressWarnings(as.numeric(chr))
  stopifnot(identical(is.na(out), is.na(chr)))
  out
}

corrected <- corrected %>%
  dplyr::mutate(dplyr::across(c(isco08, isco08_init, isco08_new), parse_code))

# split corrected data by respondent type
groups <- dplyr::group_split(corrected, source)

# apply renaming and remove fully NA rows before merging
group_sources <- purrr::map_chr(groups, ~ as.character(.x$source[[1]]))
suffixes_ordered <- suffixes[group_sources]

groups <- setNames(
  object = purrr::map2(groups, unname(suffixes_ordered), rename_and_select),
  nm = unname(suffixes_ordered)
) %>%
  lapply(drop_all_na_rows)

# treat "" as missing, then take first non-missing value
pick_first_non_na <- function(x) {
  if (is.character(x)) x[x == ""] <- NA_character_
  y <- x[!is.na(x)]
  if (length(y) == 0) {
    return(x[1][NA]) # returns typed NA
  }
  y[[1]]
}

# quick diagnostics per group (which ids are duplicated)
dup_ids_by_group <- purrr::imap(
  groups,
  ~ .x %>%
    dplyr::count(idnummer, name = "n") %>%
    dplyr::filter(n > 1) %>%
    dplyr::mutate(source = .y) %>%
    dplyr::arrange(dplyr::desc(n), idnummer)
)

# collapse each group to one row per idnummer
groups_1row <- purrr::map(
  groups,
  ~ .x %>%
    dplyr::group_by(idnummer) %>%
    dplyr::summarise(
      dplyr::across(dplyr::everything(), pick_first_non_na),
      .groups = "drop"
    )
)

# flag ids where a column had >1 distinct non-missing value inside a group
conflicts_by_group <- purrr::imap(
  groups,
  ~ .x %>%
    dplyr::group_by(idnummer) %>%
    dplyr::summarise(
      dplyr::across(
        dplyr::everything(),
        ~ dplyr::n_distinct(.[!is.na(.)]) > 1
      ),
      .groups = "drop"
    ) %>%
    dplyr::filter(dplyr::if_any(-idnummer, isTRUE)) %>%
    dplyr::mutate(source = .y)
)

# join one-to-one to prevent many-to-many issues
groups <- groups_1row %>%
  Reduce(f = function(x, y) {
    dplyr::full_join(x, y, by = "idnummer", relationship = "one-to-one")
  })

# coalesce old and new ISCO codes for each respondent type
groups_coalesced <- groups %>%
  dplyr::mutate(
    isco08r = dplyr::coalesce(isco08_new, isco08),
    isco08r_father = dplyr::coalesce(isco08_new_father, isco08_father),
    isco08r_mother = dplyr::coalesce(isco08_new_mother, isco08_mother),
    isco08r_partner = dplyr::coalesce(isco08_new_partner, isco08_partner)
  )

# generate BRC14 and ISEI-08 classifications from final ISCO codes
groups_with_classifications <- groups_coalesced %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::matches("isco08r"),
      ~ isco08_to_brc14(.),
      .names = "{gsub('isco08r', 'brc14', .col)}"
    ),
    dplyr::across(
      dplyr::matches("isco08r"),
      ~ isco08_to_isei08(.),
      .names = "{gsub('isco08r', 'isei08', .col)}"
    )
  )
```

## Finalize Analytical Dataset

- *in*: `main_with_copies`, `groups_with_classifications`
- *out*: `main_corrected.sav`
- *rule*: empty strings to `NA`; occupation text set to `NA` where final ISCO-08 code is `NA`

```{r}
#| label: finalize-dataset
#| code-fold: show
#| column: body-outset-right

# replace empty strings with NA in character columns
remove_empty_strings <- function(data_frame) {
  dplyr::mutate(
    data_frame,
    dplyr::across(
      dplyr::where(is.character),
      ~ dplyr::if_else(.x == "", NA_character_, .x)
    )
  )
}

# subset relevant columns and clean empty strings
groups_subset <- groups_with_classifications %>%
  dplyr::select(
    idnummer,
    dplyr::all_of(paste0(occupation_columns, "_cleaned")),
    dplyr::matches("isco08r"),
    dplyr::matches("brc14", ignore.case = FALSE),
    dplyr::matches("isei08")
  ) %>%
  remove_empty_strings()

# merge with main dataset and standardize column names
main_cleaned <- main_with_copies %>%
  remove_empty_strings() %>%
  dplyr::select(-dplyr::any_of(c("source", "imputed"))) %>%
  dplyr::left_join(
    groups_subset %>%
      dplyr::select(-dplyr::all_of(paste0(occupation_columns, "_cleaned"))),
    by = "idnummer",
    relationship = "one-to-one"
  ) %>%
  setNames(decapitalize(names(.)))

# guard against silent row inflation from the merge
stopifnot(nrow(main_cleaned) == nrow(main_with_copies))

# ensure occupation aligns with corrected ISCO codes
for (suffix in unname(suffixes)) {
  isco_col <- paste0("isco08r", suffix)
  occ_col <- paste0("occupation", suffix)
  main_cleaned[[occ_col]][is.na(main_cleaned[[isco_col]])] <- NA
}
main <- main_cleaned

# export corrected dataset for subsequent use
haven::write_sav(main, here::here("assets", "data", "saved", "main_corrected.sav"))
```

## Visualize Classification Corrections

```{r}
#| label: visualize-corrections
#| fig-cap: "Comparison of the Initial and Corrected ISCO-08 Codes"
#| fig-subcap: true
#| layout-ncol: 2
#| layout-type: grid
#| message: false
#| code-fold: show
#| column: body-outset-right

# scatter plot with regression line to visualize code corrections
scatter <- function(x_var, y_var, title, point_size = 2) {
  # filter complete cases for plotting
  df_clean <- main %>%
    dplyr::select(dplyr::all_of(c(x_var, y_var))) %>%
    dplyr::filter(complete.cases(.))

  # compute linear model for trend visualization
  fit <- lm(as.formula(paste(y_var, "~", x_var)), data = df_clean)

  # add residuals and predictions to dataset
  df_clean$resids <- residuals(fit)
  df_clean$y_hat <- predict(fit)
  df_clean$resids_trans <- sqrt(abs(df_clean$resids))

  # generate scatter plot with color-coded deviations
  df_clean %>%
    ggplot2::ggplot(ggplot2::aes(
      x = .data[[x_var]],
      y = .data[[y_var]],
      color = resids_trans
    )) +
    ggplot2::geom_smooth(
      method = "lm",
      se = FALSE,
      color = "black",
      linewidth = 0.8,
      formula = y ~ x
    ) +
    ggplot2::geom_jitter(
      size = point_size,
      alpha = 0.6,
      width = 0.2,
      height = 0.2
    ) +
    ggplot2::scale_color_viridis_c(option = "plasma", name = "Deviation") +
    ggplot2::labs(
      title = title,
      x = "ISCO-08 (Corrected)",
      y = "ISCO-08 (Original)"
    )
}

# plot comparisons for each respondent type
scatter("isco08r", "isco08_old", "Respondent")
scatter("isco08r_partner", "isco08_old_partner", "Partner")
scatter("isco08r_father", "isco08_old_father", "Father")
scatter("isco08r_mother", "isco08_old_mother", "Mother")
```

## Summarize ISCO Corrections

Changed digits appear in red and unchanged digits in purple within each correction table.

```{r}
#| label: define-table-formatting
#| code-fold: true
#| column: body-outset-right

# apply HTML formatting to column names (reduce bracket font size)
format_brackets <- function(vec) {
  stringr::str_replace_all(vec, c(
    "\\(([^)]+)\\)" = "<span style='font-size: 0.8em; white-space: nowrap;'>(\\1)</span>",
    "ISCO-08" = "ISCO&#8209;08"
  ))
}

# column headers for ISCO corrections table
col_names <- format_brackets(c(
  "ID No.",
  "Occupation (as reported)$\\ ^1$",
  "ISCO-08 (original)",
  "ISCO-08 Group (original)",
  "ISCO-08 (corrected)",
  "ISCO-08 Group (corrected)",
  "Justification"
))
```

```{r}
#| label: define-highlight-functions
#| code-fold: show
#| column: body-outset-right

# split strings into characters and pad for alignment
split_and_pad <- function(x, max_len) {
  chars <- unlist(strsplit(as.character(x), ""))
  pad <- rep("_", max_len - nchar(x))
  c(chars, pad)
}

# highlight differing digits in corrected codes
highlight_diff_digits <- function(a, b) {
  # return b if either input is NA
  if (anyNA(c(a, b))) {
    return(as.character(b))
  }

  # determine maximum length for padding
  max_len <- max(nchar(a), nchar(b))

  # split and pad strings to uniform length
  a_digs <- split_and_pad(a, max_len)
  b_digs <- split_and_pad(b, max_len)

  # define colors for unchanged and changed digits
  clr <- viridis::plasma(20, 1, 0)[seq(4, 12, length.out = 2)]

  # format digits with HTML color coding
  formatted <- paste0(
    mapply(function(init, new) {
      sprintf(
        '<span style="color:%s;">%s</span>',
        ifelse(init == new, clr[1], clr[2]),
        new
      )
    }, a_digs, b_digs),
    collapse = ""
  )

  # return b if formatting fails
  if (is.na(formatted)) {
    return(as.character(b))
  }
  return(formatted)
}
```

```{r}
#| label: define-summarize-function
#| code-fold: show
#| column: body-outset-right

# generate summary table of ISCO corrections with explicit source name parameter
summarize_isco_corrections <- function(x, source_name = NULL) {
  # use provided source name or infer from variable name
  source_name <- source_name %||% deparse(substitute(x))

  # calculate total rows and justified corrections
  tot_rows <- nrow(x)
  just_df <- dplyr::filter(x, !is.na(isco08_new))
  just_rows <- nrow(just_df)
  pct_just <- round(just_rows / tot_rows * 100)

  footnote_text <- glue(
    "Correction records: $N={just_rows}$ ({pct_just}% of the {tot_rows} cases in ",
    "{source_name}). Occupation titles remain in lowercase for consistency with ",
    "reported data. Revised ISCO-08 codes reflect updates from original ",
    "classifications, supported by a rationale for each change. Red digits ",
    "highlight code modifications, while unchanged digits appear in purple."
  )

  # format table with styling and footnote
  just_df %>%
    as_kable() %>%
    kableExtra::column_spec(seq(ncol(just_df)), extra_css = "vertical-align: top;") %>%
    footnote() %>%
    add_style()
}
```

```{r}
#| label: prepare-correction-tables
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# process corrected data and split by respondent type
xlsx_grouped <- xlsx %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    # apply highlighting to corrected ISCO codes
    isco08_new = highlight_diff_digits(isco08_init, isco08_new)
  ) %>%
  dplyr::ungroup() %>%
  # drop unnecessary column, if present
  dplyr::select(-dplyr::any_of("isco08")) %>%
  dplyr::group_by(source)

xlsx_split <- dplyr::group_split(xlsx_grouped, .keep = FALSE)
names(xlsx_split) <- dplyr::group_keys(xlsx_grouped)$source

# place split groups in the global environment for downstream chunks
list2env(xlsx_split, envir = .GlobalEnv)
```

```{r}
#| label: footnote-summary
#| tbl-cap: "Summary of ISCO-08 Correction Records by Respondent Category"
#| code-fold: true
#| column: body-outset-right

# compute correction statistics per source with totals
named_list(respondent, partner, mother, father) %>%
  purrr::map_dfr(~ tibble::tibble(
    total = nrow(.x), corrected = sum(!is.na(.x$isco08_new))
  ), .id = "source") %>%
  dplyr::bind_rows(dplyr::summarise(., source = "Total", dplyr::across(where(is.integer), sum))) %>%
  dplyr::mutate(source = stringr::str_to_title(source), pct = round(corrected / total * 100, 1)) %>%
  dplyr::rename(Category = source, `Total Records` = total, Corrected = corrected, `%` = pct) %$%
  kableExtra::row_spec(kable_input = as_kable(.), nrow(.), bold = TRUE) %>%
  add_style()
```

### Respondent Corrections

```{r}
#| label: isco-corrections-respondent
#| tbl-cap: "Corrected ISCO-08 Codes with Justifications for Respondents"
#| code-fold: true
#| column: page-right

summarize_isco_corrections(respondent, source_name = "respondent")
```

### Partner Corrections

```{r}
#| label: isco-corrections-partner
#| tbl-cap: "Corrected ISCO-08 Codes with Justifications for Partners"
#| code-fold: true
#| column: page-right

summarize_isco_corrections(partner, source_name = "partner")
```

### Mother Corrections

```{r}
#| label: isco-corrections-mother
#| tbl-cap: "Corrected ISCO-08 Codes with Justifications for Mother"
#| code-fold: true
#| column: page-right

summarize_isco_corrections(mother, source_name = "mother")
```

### Father Corrections

```{r}
#| label: isco-corrections-father
#| tbl-cap: "Corrected ISCO-08 Codes with Justifications for Father"
#| code-fold: true
#| column: page-right

summarize_isco_corrections(father, source_name = "father")
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                 | Content                                        |
| -------------------- | ---------------------------------------------- |
| `main_corrected.sav` | Dataset with corrected ISCO-08, BRC14, ISEI-08 |
| `brc14_map.R`        | Function: fetch CBS BRC2014 crosswalk          |
| `isco08_to_brc14.R`  | Function: map ISCO-08 codes to BRC14 groups    |
| `isco08_to_isei08.R` | Function: map ISCO-08 codes to ISEI-08 scores  |

**Next:** [Step 3 (Data Cleaning)](/analysis/data/cleaning/index.qmd)
