---
title: "Environment Setup & Dependencies"
subtitle: "Initializing Functions, Packages, and Global Options"
author: "Siard van den Bosch"
date: 2024-10-12
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 1};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Goal**       | Define global R options, a reusable helper-function library, and a shared ggplot2 theme for Steps 2–12 |
| **Inputs**     | none |
| **Outputs**    | `*.R` |
| **Method**     | Function-library definition; serialization to sourceable `.R` scripts |
| **Requires**   | None |

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: setup-environment
#| include: false
#| column: body-outset-right

rm(list = ls())
invisible(gc(full = TRUE))

# detach non-base packages
lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE)

`%>%` <- magrittr::`%>%`

do.call(options, list(
  digits = 3,
  scipen = 100,
  knitr.kable.NA = "—",
  max.print = 100,
  width = 170
))

do.call(knitr::opts_knit$set, list(
  progress = FALSE,
  verbose = FALSE
))

knitr::opts_chunk$set(cache = FALSE)

options(knitr.table.format = "html")

as_kable <- function(
  tbl,
  cols = NULL,
  format = "html",
  row.names = FALSE,
  digits = 2,
  align = "l",
  escape = FALSE,
  ...
) {
  if (is.null(cols)) {
    cols <- get0("col_names", envir = parent.frame(), inherits = TRUE)
    if (is.null(cols)) {
      cols <- colnames(tbl)
    }
  }

  # process column headers with math expressions
  if (any(grepl("\\$|\\\\", cols))) {
    cols <- sapply(cols, function(x) {
      if (grepl("\\$|\\\\", x)) {
        x <- gsub("\\\\bar", "\\\\overline", x)
        sprintf("<span class='math inline'>%s</span>", x)
      } else {
        x
      }
    }) %>%
      md_to_html()
  }

  knitr::kable(
    x = tbl,
    format = format,
    digits = digits,
    row.names = row.names,
    align = align,
    col.names = cols,
    escape = FALSE,
    ...
  )
}

md_to_html <- function(markdown_input) {
  if (!is.character(markdown_input)) {
    stop("Input must be a character string or a character vector.")
  }

  # handle latex math expressions
  convert_to_html <- function(md_text) {
    temp_md <- tempfile(fileext = ".md")
    temp_html <- tempfile(fileext = ".html")

    writeLines(md_text, temp_md)
    on.exit(unlink(c(temp_md, temp_html)), add = TRUE)

    rmarkdown::pandoc_convert(
      input = temp_md,
      to = "html",
      from = "markdown+tex_math_dollars+tex_math_single_backslash",
      output = temp_html,
      options = c("--mathml")
    )

    paste0(readLines(temp_html, warn = FALSE), collapse = " ")
  }

  html_results <- vapply(
    markdown_input,
    convert_to_html,
    character(1),
    USE.NAMES = FALSE
  )
  html_results <- stringr::str_squish(html_results)
  html_results <- gsub("^<p>(.*)<\\/p>$", "\\1", html_results)

  if (length(html_results) == 1) {
    html_results[1]
  } else {
    html_results
  }
}

add_style <- function(tbl) {
  tbl_styled <- kableExtra::kable_styling(
    kable_input = kableExtra::kable_paper(tbl),
    bootstrap_options = c("hover", "condensed", "responsive"),
    full_width = FALSE,
    html_font = '"Garamond", serif'
  )
  kableExtra::row_spec(tbl_styled, 0, bold = TRUE)
}
```

## Leveraged Packages

Auto-generated by scanning `package::function` calls across all `analysis/` step files.

```{r}
#| label: package-info
#| tbl-cap: "List of packages used in the analysis pipeline"
#| code-fold: show
#| column: page-right

pkg_regex <- "([A-Za-z][A-Za-z0-9.]*)::([A-Za-z][A-Za-z0-9._]*)\\s*\\((?!\\s*['\"])"
purrr::map(
  list.files(
    here::here("analysis"),
    full.names = TRUE,
    pattern = "index\\.qmd$",
    recursive = TRUE
  ),
  ~ readLines(.x) %>%
    stringr::str_match_all(pkg_regex) %>%
    purrr::map(~ .x[, 2])
) %>%
  unlist() %>%
  Filter(f = nchar) %>%
  table() %>%
  sort(decreasing = TRUE) %>%
  names() %>%
  purrr::map_dfr(~ {
    x <- utils::packageDescription(.x)
    url <- if (!is.null(x$URL) && is.character(x$URL)) {
      if (startsWith(x$URL, "doi:")) {
        paste0("https://doi.org/", sub("doi:", "", x$URL))
      } else {
        strsplit(x$URL, "\n|,")[[1]][1]
      }
    } else {
      sprintf("https://cran.r-project.org/package=%s", .x)
    }

    name <- x$Package %>%
      ifelse(test = is.null(url), sprintf("[%s](%s)", ., url)) %>%
      md_to_html()

    data.frame(
      Package = name,
      Description = md_to_html(stringr::str_squish(x$Description)),
      Version = x$Version,
      stringsAsFactors = FALSE
    )
  }) %>%
  as_kable() %>%
  kableExtra::column_spec(1, extra_css = "white-space: nowrap;") %>%
  kableExtra::column_spec(1:3, extra_css = "vertical-align: top;") %>%
  add_style()
```

## Custom Function Definitions

### `set_globals`

```{r}
#| label: setup-global-options
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Set global options and theme
# ' @description Establishes R options, knitr defaults, and ggplot2 theme for
# '   consistent output across the pipeline. Uses EB Garamond font for
# '   typographic consistency in academic publications.
# ' @return Invisibly returns NULL; side effects set global options and theme
set_globals <- function() {
  do.call(options, list(
    digits = 3, scipen = 50, knitr.kable.NA = "—",
    knitr.table.format = "html",
    max.print = 250, width = 170
  ))
  knitr::opts_knit$set(progress = FALSE, verbose = FALSE)
  knitr::opts_chunk$set(cache = FALSE)
  ok <- tryCatch(
    {
      sysfonts::font_add_google("EB Garamond", "Garamond")
      TRUE
    },
    error = function(e) FALSE
  )

  base_family <- if (ok) "Garamond" else ""
  if (ok) {
    showtext::showtext_auto()
    showtext::showtext_opts(dpi = 300)
  }
  garamond_theme <- ggplot2::theme_classic(
    base_size = 12,
    base_family = "Garamond"
  ) + ggplot2::theme(
    plot.title = ggplot2::element_text(
      size = 14,
      face = "bold", hjust = 0.5, margin = ggplot2::margin(b = 10)
    ),
    plot.subtitle = ggplot2::element_text(
      size = 12, face = "italic",
      hjust = 0.5, margin = ggplot2::margin(b = 10)
    ), plot.caption = ggplot2::element_text(
      size = 10,
      face = "italic", hjust = 0, margin = ggplot2::margin(t = 10)
    ),
    axis.title = ggplot2::element_text(face = "bold"),
    panel.grid.major = ggplot2::element_line(
      color = "grey80",
      linewidth = 0.5
    ), panel.grid.minor = ggplot2::element_line(
      color = "grey90",
      linewidth = 0.25
    ), plot.background = ggplot2::element_rect(
      fill = "white",
      colour = NA
    ), plot.margin = grid::unit(c(
      1, 1, 1,
      1
    ), "cm"), legend.position = "bottom", legend.box = "horizontal",
    legend.direction = "horizontal", legend.text = ggplot2::element_text(size = 8),
    legend.title = ggplot2::element_text(size = 8, face = "bold"),
    legend.background = ggplot2::element_rect(
      fill = "white",
      colour = "black"
    ), legend.key.size = grid::unit(
      0.4,
      "cm"
    ), legend.key.width = grid::unit(0.5, "cm"),
    legend.spacing.x = grid::unit(0.3, "cm"), legend.box.just = "center",
    legend.box.margin = ggplot2::margin(5, 5, 5, 5), legend.box.spacing = grid::unit(
      0,
      "cm"
    )
  )
  assign("garamond_theme", garamond_theme, envir = .GlobalEnv)
  ggplot2::theme_set(garamond_theme)
}
```

### `suppress_output`

```{r}
#| label: suppress-output
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Suppress console output
# ' @description Captures and discards console output from an expression
# ' @param expr expression to evaluate silently
# ' @return Invisibly returns the result of the expression
suppress_output <- function(expr) {
  temp_sink <- tempfile()
  sink(temp_sink)
  on.exit(sink(), add = TRUE)
  result <- eval(expr)
  return(invisible(result))
}
```

### `import_pipes`

```{r}
#| label: import-pipes
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Import pipe operators
# ' @description Assigns magrittr pipes, stringi operator, rlang null-coalescing
# '   operator, and foreach parallel operator to the global environment
# ' @return Invisibly returns NULL; side effects create global bindings
import_pipes <- function() {
  lapply(list(
    c("%>%", "magrittr"),
    c("%<>%", "magrittr"),
    c("%$%", "magrittr"),
    c("%T>%", "magrittr"),
    c("%s+%", "stringi"),
    c("%||%", "rlang"),
    c("%dopar%", "foreach")
  ), function(x) assign(x[1], getExportedValue(x[2], x[1]), .GlobalEnv))

  assign("glue", function(...) {
    eval(substitute(glue::glue(..., .sep = " ")), parent.frame())
  }, .GlobalEnv)
}
```

### `numeric_df`

```{r}
#| label: numeric-data-frame
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Convert data frame to numeric
# ' @description Transforms labelled, factor, and character columns to numeric,
# '   converting user-defined missing values to NA first
# ' @param input_data data frame or vector to convert
# ' @return Numeric data frame or vector
numeric_df <- function(input_data) {
  input_data <- labelled::user_na_to_na(input_data)
  if (is.data.frame(input_data)) {
    output_data <- lapply(input_data, function(col) {
      if (is.factor(col)) {
        as.numeric(col)
      } else if (haven::is.labelled(col)) {
        as.numeric(col)
      } else if (is.character(col)) {
        as.numeric(col)
      } else {
        col
      }
    })
    data.frame(output_data, stringsAsFactors = FALSE)
  } else if (is.factor(input_data)) {
    as.numeric(input_data)
  } else if (haven::is.labelled(input_data)) {
    as.numeric(input_data)
  } else if (is.character(input_data)) {
    as.numeric(input_data)
  } else {
    input_data
  }
}
```

### `std_res`

```{r}
#| label: standardized-residuals
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Standardized residuals
# ' @description Computes standardized residuals using robust scale for lmrob
# '   objects or rstandard for other linear models
# ' @param model fitted model object (lmrob or lm)
# ' @param assign_to optional character name for assignment in parent frame
# ' @return Numeric vector of standardized residuals
std_res <- function(model, assign_to) {
  if (inherits(model, "lmrob")) {
    stdres <- resid(model) / model$scale
  } else {
    stdres <- rstandard(model)
  }
  if (missing(assign_to)) {
    return(stdres)
  } else if (length(assign_to) == 1 && is.character(assign_to)) {
    assign(assign_to, stdres, envir = parent.frame())
  } else {
    # no action for other cases
  }
}
```

### `as_kable`

```{r}
#| label: as-kable
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Format table with kable
# ' @description Creates styled kable tables with optional math column headers,
# '   numeric formatting, and scientific notation handling
# ' @param tbl data frame to format
# ' @param cols column names; if NULL, inherits from parent frame or uses names(tbl)
# ' @param format output format (default "html")
# ' @param row.names include row names (default FALSE)
# ' @param digits decimal places for numeric columns
# ' @param align column alignment
# ' @param escape escape HTML (default FALSE)
# ' @param omit_leading_zero remove leading zero from decimals (default TRUE)
# ' @param sci_threshold use scientific notation for small values (default FALSE)
# ' @return kable object
as_kable <- function(tbl, cols = NULL, format = "html",
                     row.names = FALSE, digits = 2,
                     align = "l", escape = FALSE,
                     omit_leading_zero = TRUE,
                     sci_threshold = FALSE,
                     ...) {
  col_nms <- get0("col_names", envir = parent.frame(), inherits = TRUE)
  cols <- if (ncol(tbl) == length(col_nms)) {
    cols %||% col_nms %||% names(tbl)
  } else {
    cols %||% names(tbl)
  }

  tbl_formatted <- tbl %>%
    dplyr::mutate(dplyr::across(where(is.numeric), ~ {
      x <- as.numeric(.x)
      is_na <- is.na(x)
      result <- character(length(x))
      result[is_na] <- NA_character_

      non_na <- !is_na

      if (sci_threshold) {
        sci_cond <- (x != 0) & (abs(x) < 10^-digits) & non_na
        result[sci_cond] <- sci_to_tex(x[sci_cond])
        non_na <- non_na & !sci_cond
      }

      int_cond <- (x %% 1 == 0) & non_na
      result[int_cond] <- as.character(x[int_cond])

      remaining <- non_na & !int_cond
      formatted <- formatC(
        x[remaining],
        digits = digits,
        format = "f",
        drop0trailing = TRUE
      )
      if (omit_leading_zero) {
        formatted <- sub("^(-?)0\\.", "\\1.", formatted)
      }
      result[remaining] <- formatted

      result
    }))

  if (any(grepl("\\$|\\\\", cols))) {
    cols <- vapply(cols, function(col) {
      if (!grepl("\\$|\\\\", col)) {
        return(col)
      }
      col <- gsub("\\\\bar", "\\\\overline", col)
      sprintf("<span class='math inline'>%s</span>", col)
    }, character(1)) %>%
      md_to_html()
  }

  knitr::kable(
    x = tbl_formatted,
    format = format,
    row.names = row.names,
    align = align,
    col.names = cols,
    escape = escape,
    ...
  )
}
```

### `add_style`

```{r}
#| label: add-style
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Apply table styling
# ' @description Adds kableExtra paper styling with hover, condensed, and
# '   responsive bootstrap options; bolds header row
# ' @param tbl kable object
# ' @return Styled kable object
add_style <- function(tbl) {
  tbl_styled <- kableExtra::kable_styling(
    kable_input = kableExtra::kable_paper(tbl),
    bootstrap_options = c("hover", "condensed", "responsive"),
    full_width = FALSE,
    html_font = '"Garamond", serif'
  )
  kableExtra::row_spec(tbl_styled, 0, bold = TRUE)
}
```

### `footnote`

```{r}
#| label: footnote
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Add table footnote
# ' @description Appends footnote text to kable tables, converting LaTeX math
# '   expressions to HTML via pandoc
# ' @param tbl kable object
# ' @param text footnote text; if missing, retrieves from parent frame
# ' @return kable object with footnote
footnote <- function(tbl, text, ...) {
  if (missing(text)) {
    text <- get0("footnote_text", envir = parent.frame())
    if (is.null(text)) {
      warning("No footnote text provided, returning the table without a footnote.")
      return(tbl)
    }
  }

  text <- paste0(text, collapse = " ")
  text <- stringr::str_squish(text)

  if (any(grepl("\\$|\\\\", text))) {
    text <- sapply(text, function(x) {
      if (grepl("\\$|\\\\", x)) {
        x <- gsub("\\\\bar", "\\\\overline", x)
        sprintf("<span class='math inline'>%s</span>", x)
      } else {
        x
      }
    })
  }

  kableExtra::footnote(
    kable_input = tbl,
    general = md_to_html(text),
    escape = FALSE,
    footnote_as_chunk = TRUE,
    ...
  )
}
```

### `md_to_html`

```{r}
#| label: markdown-to-html
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Convert markdown to HTML
# ' @description Uses Pandoc to convert markdown text with LaTeX math expressions
# '   to HTML with MathML
# ' @param markdown_input character vector of markdown text
# ' @return Character vector of HTML output
md_to_html <- function(markdown_input) {
  if (!is.character(markdown_input)) {
    warning("Input set to character vector.")
    markdown_input <- as.character(markdown_input)
  }

  convert_to_html <- function(md_text) {
    temp_md <- tempfile(fileext = ".md")
    temp_html <- tempfile(fileext = ".html")

    writeLines(md_text, temp_md)
    on.exit(unlink(c(temp_md, temp_html)), add = TRUE)

    rmarkdown::pandoc_convert(
      input = temp_md,
      to = "html",
      from = "markdown+tex_math_dollars+tex_math_single_backslash",
      output = temp_html,
      options = c("--mathml")
    )

    paste0(readLines(temp_html, warn = FALSE), collapse = " ")
  }

  html_results <- vapply(
    markdown_input,
    convert_to_html,
    character(1),
    USE.NAMES = FALSE
  )
  html_results <- stringr::str_squish(html_results)
  html_results <- gsub("^<p>(.*)<\\/p>$", "\\1", html_results)

  if (length(html_results) == 1) {
    html_results[1]
  } else {
    html_results
  }
}
```

### `std_text`

```{r}
#| label: standardize-text-input
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Standardize text
# ' @description Converts to lowercase, replaces non-ASCII characters, removes
# '   non-alphanumeric characters, and normalizes whitespace
# ' @param x character vector
# ' @return Standardized character vector
std_text <- function(x) {
  x <- tolower(x)
  x <- textclean::replace_non_ascii(x)
  x <- gsub("[^[:alnum:]]", " ", x)
  x <- gsub("\\s+", " ", x)
  x <- gsub("^\\s+|\\s+$", "", x)
  return(x)
}
```

### `describe`

```{r}
#| label: describe
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Descriptive statistics
# ' @description Computes mean, SD, min, max, skewness, kurtosis, SE, and
# '   optional normality tests (Shapiro-Wilk for n <= 5000, Lilliefors otherwise)
# ' @param df data frame
# ' @param exclude column names to exclude
# ' @param kable_format output format
# ' @param caption table caption
# ' @param normality include normality test (default TRUE)
# ' @param print_only return formatted table (TRUE) or raw data (FALSE)
# ' @param add_footnote include method footnote
# ' @param latex use LaTeX column names
# ' @param decimals decimal places
# ' @return kable object or data frame
describe <- function(
  df,
  exclude = NULL,
  kable_format = "html",
  caption = "Descriptive Statistics",
  normality = TRUE,
  print_only = TRUE,
  add_footnote = TRUE,
  latex = TRUE,
  decimals = 3,
  ...
) {
  df <- df %>%
    dplyr::mutate(dplyr::across(where(is.logical), as.numeric)) %>%
    dplyr::select(-dplyr::any_of(exclude))

  nt <- if (normality) {
    df %>%
      dplyr::select(where(\(x) is.numeric(x) && length(unique(x)) > 4)) %>%
      purrr::map_dfr(\(x) {
        tst <- if (nrow(df) > 5000) {
          DescTools::LillieTest(x)
        } else {
          shapiro.test(x)
        }
        tibble::tibble(norm = tst$statistic, p = tst$p.value)
      }, .id = "var")
  } else {
    tibble::tibble(var = character(), norm = numeric(), p = numeric())
  }

  ds <- psych::describe(df, ranges = TRUE) %>%
    tibble::as_tibble(rownames = "var") %>%
    dplyr::select(var, n, mean, min, max, sd, skew, kurtosis, se) %>%
    dplyr::left_join(nt, by = "var") %>%
    dplyr::mutate(p = format_p_value(p, TRUE, TRUE, TRUE))

  if (print_only) {
    ds %<>%
      dplyr::mutate(dplyr::across(where(is.numeric), ~
        format(round(.x, decimals), nsmall = decimals)))
  }

  cn <- if (latex) {
    c(
      "$\\text{}$", "$n$", "$\\bar{x}$", "$\\text{min}$",
      "$\\text{max}$", "$\\sigma$", "$\\text{Sk}$", "$\\text{K}$",
      "$SE$", "$\\text{Normality}^{\\dagger}$", "$p$"
    )
  } else {
    c(
      "Var", "n", "Mean", "Min", "Max", "SD",
      "Skew", "Kurtosis", "SE", "Norm", "p"
    )
  }
  if (!normality) {
    cn <- head(cn, -2)
  }

  out <- knitr::kable(
    ds,
    format = kable_format,
    digits = decimals,
    row.names = FALSE,
    escape = FALSE,
    col.names = cn,
    caption = caption,
    align = c("l", rep("c", length(cn) - 1)),
    ...
  )

  if (normality && add_footnote && kable_format == "html") {
    out <- kableExtra::footnote(
      out,
      general = paste0(
        ifelse(nrow(df) <= 5000, "Shapiro-Wilk", "Lilliefors"),
        " Test$^{\\dagger}$"
      ),
      footnote_as_chunk = TRUE,
      escape = FALSE
    )
  }

  if (print_only) {
    return(out)
  } else {
    return(invisible(ds))
  }
}
```

### `summarize_na`

```{r}
#| label: summarize-missing-values
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Summarize missing values
# ' @description Returns count and percentage of missing values per variable
# ' @param x data frame or existing miss_var_summary output
# ' @return Data frame with n_miss and pct_miss columns
summarize_na <- function(x) {
  if (all(c("n_miss", "pct_miss") %in% names(x))) {
    mvs <- x
  } else {
    mvs <- naniar::miss_var_summary(x, order = TRUE, add_cumsum = FALSE) %>%
      dplyr::mutate(
        n_miss = as.numeric(n_miss),
        pct_miss = pct_miss
      )
  }
  return(mvs)
}
```

### `format_p_value`

```{r}
#| label: format-p-value
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Format p-value
# ' @description Converts numeric p-values to formatted strings with optional
# '   significance stars and LaTeX/HTML output
# ' @param p_value numeric p-value(s)
# ' @param add_stars include significance stars
# ' @param use_concat concatenate p and stars
# ' @param omit_p_label return only stars
# ' @param verbose print messages
# ' @param use_dot_for_0.1 use dot for p <= 0.1
# ' @param as_html convert to HTML
# ' @return Formatted p-value string(s)
format_p_value <- function(
  p_value,
  add_stars = FALSE,
  use_concat = TRUE,
  omit_p_label = FALSE,
  verbose = FALSE,
  use_dot_for_0.1 = FALSE,
  as_html = TRUE
) {
  if (is.character(p_value)) {
    p_value <- utils::type.convert(p_value, as.is = TRUE)
  }

  if (!omit_p_label) {
    if (isTRUE(all(p_value > 0.05, na.rm = TRUE))) {
      if (verbose) {
        message("All p-values are greater than 0.05, no changes made.")
      }
      return(p_value)
    }
  }

  formatted_p <- ifelse(
    p_value < 0.001,
    "$\\leq 0.001$", sprintf("$%.3f$", p_value)
  )
  formatted_p[is.na(formatted_p)] <- ""
  formatted_p <- format(formatted_p, nsmall = 4, justify = "left")

  ast <- function(i) sprintf("$^{%s}$", ifelse(i == 0, "\\cdot", strrep("\\ast", i)))

  if (add_stars) {
    stars_label <- ifelse(
      is.na(p_value), " ",
      ifelse(
        p_value <= 0.001, ast(3),
        ifelse(
          p_value <= 0.01, ast(2),
          ifelse(
            p_value <= 0.05, ast(1),
            ifelse(
              (use_dot_for_0.1 && p_value <= 0.1), ast(0), ""
            )
          )
        )
      )
    )

    stars_label <- format(
      stars_label,
      justify = ifelse(omit_p_label, "right", "left")
    )

    if (omit_p_label) {
      formatted_p <- paste0("$\\ $", stars_label)
    } else {
      if (use_concat) {
        formatted_p <- paste0(formatted_p, stars_label)
      } else {
        formatted_p <- list(p = formatted_p, stars = stars_label)
      }
    }
  } else {
    if (!use_concat) {
      formatted_p <- list(p = formatted_p, stars = "")
    }
  }

  if (as_html) {
    formatted_p <- md_to_html(formatted_p)
  }
  return(formatted_p)
}
```

### `structure_models`

```{r}
#| label: structure-models
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Structure model results
# ' @description Combines unadjusted and adjusted model fits (robust and OLS)
# '   into a nested list with consistent naming and optional bootstrap components
# ' @param unadj_mods list of unadjusted model results
# ' @param adj_mods list of adjusted model results
# ' @return Nested list: adjusted/unadjusted -> robust/ols -> model_0..model_n
structure_models <- function(unadj_mods, adj_mods) {
  update_model <- function(mod) {
    if (!is.null(mod$call$formula)) {
      mod$call$formula <- formula(mod)
    }
    if (!is.null(mod$model)) {
      if (!is.null(mod$model$call$formula)) {
        mod$model$call$formula <- formula(mod$model)
      }
      mod$model$call$data <- mod$call$data
    }
    mod
  }

  update_mods <- function(mods) {
    lapply(mods, function(m) {
      if (!is.null(m$robust_model_results$model_object)) {
        m$robust_model_results$model_object <- update_model(
          m$robust_model_results$model_object
        )
      }
      if (!is.null(m$ols_model_results$model_object)) {
        m$ols_model_results$model_object <- update_model(
          m$ols_model_results$model_object
        )
      }
      m
    })
  }

  adj_mods <- update_mods(adj_mods)
  unadj_mods <- update_mods(unadj_mods)

  create_model_structure <- function(fit, include_bootstrap = TRUE) {
    process_fit <- function(fit_res) {
      if (is.null(fit_res) || is.null(fit_res$analysis_output)) {
        return(list(
          fit = NULL,
          bootstrap = list(
            std_errors = NULL, coefs = NULL,
            p_value = NULL, conf_int = NULL, ests = NULL
          )
        ))
      }
      bootstrap_data <- if (include_bootstrap) {
        list(
          std_errors = fit_res$analysis_output$standard_errors,
          coefs = if (!is.null(fit_res$analysis_output$bootstrap_estimates)) {
            colMeans(fit_res$analysis_output$bootstrap_estimates, na.rm = TRUE)
          } else {
            NULL
          },
          p_value = unlist(fit_res$analysis_output$p_values),
          conf_int = fit_res$analysis_output$confidence_intervals,
          ests = fit_res$analysis_output$bootstrap_estimates
        )
      } else {
        list(
          std_errors = NULL, coefs = NULL,
          p_value = NULL, conf_int = NULL, ests = NULL
        )
      }
      list(fit = fit_res$model_object, bootstrap = bootstrap_data)
    }
    list(
      robust = process_fit(fit$robust_model_results),
      ols = process_fit(fit$ols_model_results)
    )
  }

  model_names <- paste0("model_", seq_along(adj_mods) - 1)
  adjusted <- lapply(adj_mods, create_model_structure)
  unadjusted <- lapply(unadj_mods, create_model_structure)

  list(
    adjusted = list(
      robust = setNames(lapply(adjusted, `[[`, "robust"), model_names),
      ols = setNames(lapply(adjusted, `[[`, "ols"), model_names)
    ),
    unadjusted = list(
      robust = setNames(lapply(unadjusted, `[[`, "robust"), model_names),
      ols = setNames(lapply(unadjusted, `[[`, "ols"), model_names)
    )
  )
}
```

### `update_isco08r`

```{r}
#| label: update-isco08r
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Update ISCO-08 value labels
# ' @description Applies job title labels to ISCO-08 coded variables
# ' @param col_name column name to update
# ' @param data data frame containing the column
# ' @return Side effect: updates value labels in parent frame
update_isco08r <- function(col_name, data) {
  values <- unique(na.omit(data[[col_name]]))
  labels <- sapply(values, function(x) {
    isco08_to_jobtitle(x)
  })
  values_named <- setNames(values, labels)
  for (i in seq_along(values_named)) {
    eval.parent(substitute(
      update_value_label(
        data,
        col_name,
        as.numeric(values_named[i]),
        names(values_named)[i]
      )
    ))
  }
}
```

### `update_brc14`

```{r}
#| label: update-brc14
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Update BRC14 value labels
# ' @description Applies occupational group labels from CBS crosswalk
# ' @param col_name column name to update
# ' @param data data frame containing the column
# ' @param mapping BRC14 crosswalk data frame
# ' @return Side effect: updates value labels in parent frame
update_brc14 <- function(col_name, data, mapping) {
  values <- unique(na.omit(data[[col_name]]))
  labels <- sapply(values, function(x) {
    unique(mapping$BRC2014beroepsgroep_label[as.numeric(mapping$BRC2014beroepsgroep) == x])
  })
  values_named <- setNames(values, labels)
  for (i in seq_along(values_named)) {
    eval.parent(substitute(
      update_value_label(
        data,
        col_name,
        as.numeric(values_named[i]),
        names(values_named)[i]
      )
    ))
  }
}
```

### `isco08_to_jobtitle`

```{r}
#| label: isco08-to-jobtitle
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title ISCO-08 code to job title
# ' @description Converts ISCO-08 numeric codes to descriptive job titles using
# '   ISCO08ConveRsions package with fallback error handling
# ' @param x numeric ISCO-08 code(s)
# ' @return Character vector of job titles (NA for invalid codes)
isco08_to_jobtitle <- function(x) {
  sapply(x, function(code) {
    if (is.na(code)) {
      return(NA)
    }
    f <- ISCO08ConveRsions::isco08tojobtitle
    tryCatch(
      tryCatch(
        f(formatC(as.integer(code), width = 4, flag = "0", format = "d")),
        error = function(e) {
          f(code)
        }
      ),
      error = function(e) NA
    )
  })
}
```

### `labels_to_string`

```{r}
#| label: labels-to-string
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Labels to string
# ' @description Concatenates value labels as "label" = value pairs
# ' @param data data frame
# ' @param column column name
# ' @return Character string of label-value pairs
labels_to_string <- function(data, column) {
  labels <- attr(data[[column]], "labels")
  if (is.null(labels) || length(labels) == 0) {
    "No labels in original data."
  } else {
    paste(purrr::imap_chr(labels, ~ paste0(dQuote(.y, FALSE), " = ", .x)), collapse = ", ")
  }
}
```

### `drop_all_na_rows`

```{r}
#| label: remove-rows-with-all-na
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Drop all-NA rows
# ' @description Removes rows where all values are NA
# ' @param x data frame
# ' @return Data frame with all-NA rows removed
drop_all_na_rows <- function(x) {
  x[rowSums(!is.na(x)) > 0, , drop = FALSE]
}
```

### `empty_to_na`

```{r}
#| label: convert-empty-to-na
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Empty strings to NA
# ' @description Replaces empty strings ("") with NA_character_
# ' @param x character vector or other
# ' @return Vector with empty strings replaced by NA
empty_to_na <- function(x) {
  if (is.character(x)) {
    ifelse(x == "", NA_character_, x)
  } else {
    x
  }
}
```

### `decapitalize`

```{r}
#| label: decapitalize
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Decapitalize first letter
# ' @description Converts first character to lowercase if uppercase
# ' @param x character vector
# ' @return Character vector with decapitalized first letters
decapitalize <- function(x) {
  upper <- gsub("^([A-Z])[a-z].*$", "\\1", x)
  ifelse(
    nchar(upper) == 1,
    tolower(upper) %s+% substr(x, 2, nchar(x)), x
  )
}
```

### `remove_invariant`

```{r}
#| label: remove-invariant
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Remove invariant columns
# ' @description Drops columns with zero variance (single unique non-NA value)
# ' @param df data frame
# ' @return Data frame with invariant columns removed
remove_invariant <- function(df) {
  df[vapply(df, function(x) length(na.omit(unique(x))) > 1, logical(1))]
}
```

### `format_numeric`

```{r}
#| label: format-numeric
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Format numeric columns
# ' @description Rounds all numeric columns to specified decimal places
# ' @param df data frame
# ' @param digits decimal places (default 3)
# ' @return Data frame with formatted numeric columns
format_numeric <- function(df, digits = 3) {
  dplyr::mutate(df, dplyr::across(
    dplyr::where(is.numeric),
    ~ format(round(., digits), nsmall = 0, trim = TRUE)
  ))
}
```

### `order_array`

```{r}
#| label: order-array
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Order array elements
# ' @description Returns indices for standard table ordering: numeric values
# '   (ascending), non-numeric labels (alphabetic), "Total" last
# ' @param x character vector
# ' @return Integer vector of order indices
order_array <- function(x) {
  numeric_part <- suppressWarnings(as.numeric(x))
  numeric_order <- order(numeric_part, na.last = NA)
  non_numeric_order <- order(x[is.na(numeric_part) & x != "Total"])
  total_index <- which(x == "Total")
  order_indices <- c(
    numeric_order,
    which(is.na(numeric_part) & x != "Total")[non_numeric_order],
    total_index
  )
  return(order_indices)
}
```

### `copy_labels`

```{r}
#| label: copy-labels
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Copy labels between data frames
# ' @description Transfers labelled metadata from source to target, preserving
# '   interpretive mappings and handling type conversions
# ' @param source data frame with labels
# ' @param target data frame to receive labels
# ' @return Labelled target data frame
copy_labels <- function(source, target) {
  common_names <- intersect(names(target), names(source))
  unique_names <- setdiff(names(target), names(source))

  target_reformatted <- purrr::map2_df(
    target[common_names],
    common_names,
    ~ {
      src <- source[[.y]]
      tgt <- .x
      if (is.numeric(src) && is.factor(tgt)) {
        as.numeric(as.character(tgt))
      } else {
        fun_name <- paste0("as.", typeof(src))
        getFunction(fun_name)(tgt)
      }
    }
  )

  target_reformatted <- labelled::copy_labels(source, target_reformatted)
  target_reformatted <- cbind(target_reformatted, target[unique_names])
  target_reformatted <- target_reformatted[, names(target)]

  if (any(grepl("tbl", class(source)))) {
    tibble::as_tibble(target_reformatted)
  } else {
    target_reformatted
  }
}
```

### `pval_str`

```{r}
#| label: pval-str
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title P-value significance string
# ' @description Generates formatted string showing significance thresholds
# ' @param html convert to HTML (default TRUE)
# ' @param concatenate return as single string (default TRUE)
# ' @param use_dot_for_0.1 include p <= 0.1 threshold
# ' @param non_sig include non-significant label
# ' @return Character string of significance thresholds
pval_str <- function(html = TRUE, concatenate = TRUE, use_dot_for_0.1 = FALSE, non_sig = FALSE) {
  p_values <- c(
    "$^{\\ast\\ast\\ast}p \\leq 0.001$",
    "$^{\\ast\\ast}p \\leq 0.01$",
    "$^{\\ast}p \\leq 0.05$"
  )
  if (use_dot_for_0.1) {
    p_values <- c(p_values, "$^{\\cdot}p \\leq 0.1$")
  }
  if (non_sig) {
    p_values <- c(p_values, "$^{\\text{n.s.}}p > 0.05$")
  }
  if (html) {
    p_values <- md_to_html(p_values)
  }
  if (concatenate) {
    p_values <- toString(p_values)
  }
  return(p_values)
}
```

### `frequency_table`

```{r}
#| label: frequency-table
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Frequency table
# ' @description Generates frequency distribution with counts, valid percentages,
# '   and cumulative percentages; positions missing values distinctly
# ' @param x data frame
# ' @param var_name variable name (default: last column)
# ' @return Printed kable table with frequencies
frequency_table <- function(x, var_name = NULL) {
  if (is.null(var_name)) {
    var_name <- names(x)[ncol(x)]
  }
  col_data <- labelled::user_na_to_regular_na(x[[var_name]])

  val_labels <- attr(col_data, "labels")
  if (is.null(val_labels)) {
    val_labels <- labelled::get_value_labels(col_data)[[1]]
  }
  labels_df <- data.frame(
    value = as.character(val_labels),
    label = dQuote(names(val_labels)),
    stringsAsFactors = FALSE
  )

  freq_tbl <- as.data.frame(
    table(col_data, useNA = "ifany"),
    stringsAsFactors = FALSE
  )
  names(freq_tbl) <- c("value", "frequency")
  freq_tbl$value <- as.character(freq_tbl$value)

  missing_cond <- freq_tbl$value %in% c("NA", "<NA>") | is.na(freq_tbl$value)
  freq_tbl$value[missing_cond] <- "N/A"

  valid_rows <- freq_tbl[freq_tbl$value != "N/A", ]
  missing_rows <- freq_tbl[freq_tbl$value == "N/A", ]
  valid_rows <- valid_rows[gtools::mixedorder(valid_rows$value), ]
  freq_tbl <- dplyr::bind_rows(valid_rows, missing_rows)

  total_count <- sum(freq_tbl$frequency)
  valid_total <- sum(valid_rows$frequency)
  freq_tbl$cum_perc_total <- round(100 * cumsum(freq_tbl$frequency) / total_count, 2)

  freq_tbl$perc_valid <- ifelse(
    freq_tbl$value == "N/A", NA,
    round(100 * freq_tbl$frequency / valid_total, 2)
  )
  valid_freq <- freq_tbl$frequency[freq_tbl$value != "N/A"]
  cum_valid <- round(100 * cumsum(valid_freq) / valid_total, 2)
  freq_tbl$cum_perc_valid <- NA
  freq_tbl$cum_perc_valid[freq_tbl$value != "N/A"] <- cum_valid

  if (nrow(freq_tbl) > 0) {
    freq_tbl$cum_perc_total[nrow(freq_tbl)] <- 100
    if (any(freq_tbl$value != "N/A")) {
      last_valid <- max(which(freq_tbl$value != "N/A"))
      freq_tbl$cum_perc_valid[last_valid] <- 100
    }
  }

  tbl_names <- c(
    "Value", "Label", "Frequency", "$\\%$ Valid",
    "$\\%$ Cum. Valid", "$\\%$ Cum. Total"
  )

  tbl <- dplyr::left_join(freq_tbl, labels_df, by = "value") %>%
    dplyr::select(
      value, label, frequency, perc_valid, cum_perc_valid, cum_perc_total
    ) %>%
    as_kable(cols = tbl_names) %>%
    footnote(general_title = "Original coding: ") %>%
    add_style()

  print(tbl)
}
```

### `col_to_fct`

```{r}
#| label: convert-to-factor
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Convert columns to factors
# ' @description Transforms labelled numeric columns to factors using embedded
# '   value labels as factor levels
# ' @param df data frame
# ' @param cols columns to convert (default: all)
# ' @param exclude columns to exclude
# ' @param inplace assign result in parent frame (default FALSE)
# ' @return Data frame with converted factors
col_to_fct <- function(df, cols, exclude = NULL, inplace = FALSE) {
  if (missing(cols)) {
    warning("No columns specified. All columns will be converted to factors.")
    cols <- names(df)
  }
  to_convert <- na.omit(setdiff(names(df[cols]), exclude))
  df_copy <- df
  df_copy[to_convert] <- lapply(df_copy[to_convert], \(x) {
    lvls <- attr(x, "labels")
    if (!is.null(lvls)) factor(x, levels = unname(lvls), labels = names(lvls)) else factor(x)
  })

  if (inplace) {
    assign(deparse(substitute(df)), df_copy, envir = parent.frame())
  } else {
    df_copy
  }
}
```

### `sci_to_tex`

```{r}
#| label: format-scientific-latex
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Scientific notation to LaTeX
# ' @description Converts numeric values to scientific notation LaTeX strings
# ' @param x numeric value
# ' @return HTML-formatted scientific notation string
sci_to_tex <- function(x) {
  sci <- format(x, scientific = TRUE)
  m <- as.numeric(strsplit(sci, "e")[[1]][1])
  exp <- as.numeric(strsplit(sci, "e")[[1]][2])
  sci <- sprintf("$%.2f \\times 10^{%d}$", m, exp)
  md_to_html(sci)
}
```

### `named_list`

```{r}
#| label: named-list
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Named list from expressions
# ' @description Builds a named list using expression names as keys
# ' @param ... expressions to include
# ' @return Named list
named_list <- function(...) {
  quos <- rlang::enquos(...)
  vals <- list(...)
  nms <- names(vals)
  inf_nms <- sapply(quos, rlang::as_label)

  if (!is.null(nms) && length(nms)) {
    nms[nms == ""] <- inf_nms[nms == ""]
  } else {
    nms <- inf_nms
  }

  setNames(lapply(vals, eval), nms)
}
```

### `get_formula`

```{r}
#| label: get-formula
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Extract model formula
# ' @description Retrieves formula from model object, removes extraneous
# '   whitespace, preserves environment context
# ' @param model_object fitted model
# ' @return Cleaned formula object
get_formula <- function(model_object) {
  f <- formula(model_object)
  env <- environment(f)

  formula_str <- paste(deparse(f), collapse = "")
  clean_formula_str <- gsub("\\s+", " ", formula_str)
  clean_formula <- as.formula(clean_formula_str, env = env)
  return(clean_formula)
}
```

### `format_coef`

```{r}
#| label: format-coef
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Format coefficient
# ' @description Formats regression coefficients with significance stars,
# '   optional standard errors, and alignment spacing
# ' @param coef coefficient value
# ' @param pval p-value for significance stars
# ' @param se standard error (optional)
# ' @param digits decimal places
# ' @param linebreak add line break before SE
# ' @return HTML-formatted coefficient string
format_coef <- function(coef, pval = NULL, se = NULL, digits = 3, linebreak = TRUE) {
  if (is.na(coef)) {
    return("--")
  }
  p_fmt <- format_p_value(
    pval,
    add_stars = TRUE,
    omit_p_label = TRUE
  )
  spc <- ifelse(
    coef >= 0 & coef < 10,
    "<span style=\"display:inline-block; width: 4px\"> </span>",
    ""
  )
  fmt_num <- function(x, digits) {
    num_str <- sprintf(paste0("%.", digits, "f"), x)
    sub("^(-?)0\\.", "\\1.", num_str)
  }
  main_str <- sprintf(
    "%s%s<sup>%s</sup>",
    spc,
    fmt_num(coef, digits),
    p_fmt
  )
  if (!is.null(se)) {
    se_str <- sprintf(
      "<span style='font-size: 0.8em'>(%s)</span>",
      fmt_num(se, digits)
    )
    main_str <- paste0(main_str, ifelse(linebreak, "<br>", ""), se_str)
  }
  return(main_str)
}
```

### `rm_dup_cols`

```{r}
#| label: remove-duplicate-columns
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Remove duplicate columns
# ' @description Identifies columns with identical content and retains only one
# ' @param df data frame
# ' @param keep "first" or "last" occurrence to retain
# ' @return Data frame with duplicate columns removed
rm_dup_cols <- function(df, keep = "first") {
  if (!is.data.frame(df)) {
    stop("Input must be a data frame.")
  }

  col_cont <- apply(df, 2, function(col) paste(col, collapse = "\r"))
  uniq_cols <- unique(col_cont)
  col_inds <- lapply(uniq_cols, function(x) which(col_cont == x))

  sel_cols <- sapply(col_inds, function(inds) {
    if (keep == "last") {
      return(tail(inds, 1))
    } else {
      return(head(inds, 1))
    }
  })

  df[, sel_cols, drop = FALSE]
}
```

### `row_grouping`

```{r}
#| label: get-row-grouping
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Row grouping counts
# ' @description Counts occurrences of each unique group for table row grouping
# ' @param x vector of group identifiers
# ' @return Named integer vector of group sizes
row_grouping <- function(x) {
  x <- table(x)[unique(x)]
  names(x) <- make.unique(names(x))
  invisible(capture.output(dput(x, control = "niceNames")))
  return(x)
}
```

### `dim_ci_rows`

```{r}
#| label: dim-ci-intervals
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Dim non-significant rows
# ' @description Reduces opacity for rows where CI includes zero (sig_perc < 100)
# ' @param kbl kable object
# ' @param df data frame with sig_perc column
# ' @param opacity_level opacity for dimmed rows (default 0.5)
# ' @return Styled kable object
dim_ci_rows <- function(kbl, df, opacity_level = 0.5) {
  row_indices <- which(suppressWarnings(as.numeric(df$sig_perc)) < 100)
  opacity_str <- formatC(opacity_level, format = "f", digits = 1)
  kableExtra::row_spec(kbl, row_indices, extra_css = paste0("opacity: ", opacity_str, ";"))
}
```

### `replace_duplicates`

```{r}
#| label: replace-duplicates
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Replace duplicate values
# ' @description Replaces repeated values in a column with placeholder, keeping
# '   first occurrence
# ' @param df data frame
# ' @param col column name (unquoted)
# ' @param replace_with replacement value (default NA)
# ' @return Data frame with duplicates replaced
replace_duplicates <- function(df, col, replace_with = NA) {
  col <- rlang::ensym(col)
  dplyr::mutate(df, !!col := as.character(!!col)) %>%
    dplyr::mutate(dup_flag = duplicated(!!col)) %>%
    dplyr::mutate(!!col := ifelse(dup_flag, replace_with, !!col)) %>%
    dplyr::select(-dup_flag) %>%
    dplyr::mutate(!!col := if (is.factor(dplyr::pull(df, !!col))) factor(!!col) else !!col)
}
```

## Save Functions

```{r}
#| label: save-functions
#| message: false
#| code-fold: show
#| column: body-outset-right

# ' @title Save function to file
# ' @description Writes function definition to R script and applies styler
# ' @param fun function object
# ' @param name optional file name (default: function name)
# ' @param dir output directory
# ' @return Invisibly returns NULL
save_function <- function(fun, name = NULL, dir = here::here("assets", "functions")) {
  stopifnot(is.function(fun) && dir.exists(dir))
  fun_name <- if (is.null(name)) deparse(substitute(fun)) else name
  file_path <- file.path(dir, paste0(fun_name, ".R"))
  x <- deparse(fun)
  x[1] <- paste0(fun_name, " <- ", x[1])
  fun_string <- paste0(x, collapse = "\n")
  writeLines(fun_string, con = file_path)
  styler::style_file(file_path)
  return(invisible())
}

# identify all user-defined functions in global environment
.funs <- purrr::keep(ls(), ~ is.function(get(.x)) &&
  identical(environmentName(environment(get(.x))), "R_GlobalEnv"))

# save each function to assets/functions/
purrr::walk(.funs, ~ {
  fun <- get(.x)
  invisible(suppress_output(save_function(fun, name = .x)))
})
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                   | Content                                           |
| ---------------------- | ------------------------------------------------- |
| `assets/functions/*.R` | Sourceable helper-function library for Steps 2–12 |

**Next:** [Step 2 (ISCO-08 Standardization)](/analysis/data/preparation/index.qmd)
