---
title: "Post-Imputation Data Finalization"
subtitle: "Consolidating Variable Structure and Preparing Analysis-Ready Dataset"
author: "Siard van den Bosch"
date: 2025-01-20
date-modified: last-modified
---

```{=html}
<script>
  window.config = {offset: 6};
</script>
```

::: {style="margin-block-end: 2rem;"}
:::

|                |         |
| -------------- | ------- |
| **Purpose**    | Finalize imputed dataset for downstream regression modeling. |
| **Inputs**     | `imputed_data_combined.sav` |
| **Outputs**    | `analysis_ready.sav` |
| **Method**     | Variable verification; composite integrity checks; descriptive summarization |
| **Requires**   | [Step 5 (Missing Data Imputation)](/analysis/data/imputation/index.qmd) |

::: {style="margin-block-end: 1rem;"}
:::

```{r}
#| label: load-project-resources
#| message: false
#| warning: false
#| code-fold: show
#| column: body-outset-right

# clear global env and force full garbage collection
rm(list = ls(envir = .GlobalEnv), envir = .GlobalEnv)
invisible(gc(full = TRUE))

# detach non-base packages to avoid namespace interference
invisible(lapply(grep(":|autoloads|globalenv", search(),
  invert = TRUE, value = TRUE, ignore.case = TRUE
), detach, character.only = TRUE, unload = TRUE, force = TRUE))

# source custom functions for data handling and analysis
purrr::walk(file.path(here::here("assets", "functions"), paste0(c(
  "summarize_na", "set_globals", "describe", "import_pipes",
  "add_style", "beautify_names", "as_kable", "format_p_value",
  "footnote", "md_to_html", "numeric_df"
), ".R")), source)

# set global options and import pipe operators
set_globals()
import_pipes()

# load imputed dataset from step 5
imputed_data <- haven::read_sav(
  here::here("assets", "data", "saved", "imputed_data_combined.sav")
)
```

## Verify Dataset from Step 5

```{r}
#| label: define-expected-variables
#| code-fold: true
#| column: body-outset-right

# define expected variable structure
expected_vars <- list(
  # respondent's own SES
  respondent = c("education_self", "isei08_self"),

  # parental SES (individual indicators)
  parental_individual = c(
    "education_father", "education_mother",
    "isei08_father", "isei08_mother"
  ),

  # parental SES (composites)
  parental_composites = c("education_parental", "isei08_parental"),

  # partner SES
  partner = c("education_partner", "isei08_partner"),

  # core analysis variables
  core = c("idnummer", "partner", "dhd_index", "dhd_kcal")
)

# flatten for checking
all_expected <- unlist(expected_vars, use.names = FALSE)
```

- **in**: `imputed_data` (from Step 5)
- **out**: `imputed_data` (canonical naming, composites verified)
- **rule**: composites = row-mean of parental components, NA only when both parents missing

```{r}
#| label: verify-step5-data
#| results: asis
#| code-fold: show
#| column: body-outset-right

# rename legacy occupation_* variables to isei08_* if present
# (temporary fix until upstream Step 5 is re-run with corrected code)
legacy_renames <- c(
  "occupation_self" = "isei08_self",
  "occupation_father" = "isei08_father",
  "occupation_mother" = "isei08_mother",
  "occupation_partner" = "isei08_partner"
)

legacy_found <- names(legacy_renames)[names(legacy_renames) %in% names(imputed_data)]
if (length(legacy_found) > 0) {
  for (old_name in legacy_found) {
    new_name <- legacy_renames[[old_name]]
    if (!new_name %in% names(imputed_data)) {
      names(imputed_data)[names(imputed_data) == old_name] <- new_name
    }
  }
}

# identify present and missing expected variables
vars_present <- all_expected[all_expected %in% names(imputed_data)]
vars_missing <- setdiff(all_expected, vars_present)

# count observations by partner status
n_obs <- nrow(imputed_data)
n_partnered <- sum(imputed_data$partner == 1, na.rm = TRUE)
n_non_partnered <- sum(imputed_data$partner == 0, na.rm = TRUE)

# check for any remaining legacy variable names
legacy_names <- c(
  "parental_edu", "parental_occ", "partner_edu", "partner_occ",
  "occupation_self", "occupation_father", "occupation_mother", "occupation_partner"
)
legacy_present <- legacy_names[legacy_names %in% names(imputed_data)]

# create or recreate parental composites to ensure consistent computation
# use na.rm = FALSE for complete-case means, consistent with step 3
composites_created <- character(0)
composites_updated <- character(0)

# education_parental: mean of father and mother education
if (all(c("education_father", "education_mother") %in% names(imputed_data))) {
  edu_new <- rowMeans(
    imputed_data[, c("education_father", "education_mother")],
    na.rm = FALSE
  )
  # set to NA where both components are missing
  both_missing <- is.na(imputed_data$education_father) &
    is.na(imputed_data$education_mother)
  edu_new[both_missing] <- NA

  if (!"education_parental" %in% names(imputed_data)) {
    imputed_data$education_parental <- edu_new
    composites_created <- c(composites_created, "education_parental")
  } else if (!isTRUE(all.equal(imputed_data$education_parental, edu_new,
    tolerance = 1e-10, check.attributes = FALSE
  ))) {
    imputed_data$education_parental <- edu_new
    composites_updated <- c(composites_updated, "education_parental")
  }
}

# isei08_parental: mean of father and mother occupation
if (all(c("isei08_father", "isei08_mother") %in% names(imputed_data))) {
  occ_new <- rowMeans(
    imputed_data[, c("isei08_father", "isei08_mother")],
    na.rm = FALSE
  )
  # set to NA where both components are missing
  both_missing <- is.na(imputed_data$isei08_father) &
    is.na(imputed_data$isei08_mother)
  occ_new[both_missing] <- NA

  if (!"isei08_parental" %in% names(imputed_data)) {
    imputed_data$isei08_parental <- occ_new
    composites_created <- c(composites_created, "isei08_parental")
  } else if (!isTRUE(all.equal(imputed_data$isei08_parental, occ_new,
    tolerance = 1e-10, check.attributes = FALSE
  ))) {
    imputed_data$isei08_parental <- occ_new
    composites_updated <- c(composites_updated, "isei08_parental")
  }
}

# re-check after creating/updating composites
vars_present <- all_expected[all_expected %in% names(imputed_data)]
vars_missing <- setdiff(all_expected, vars_present)

# generate verification report
verification_status <- if (length(vars_missing) == 0 && length(legacy_present) == 0) {
  "<span style='color: green;'>&#x2714; All expected variables present</span>"
} else if (length(legacy_present) > 0) {
  glue(
    "<span style='color: red;'>&#x2718; Legacy variable names detected: ",
    "<code>{paste(legacy_present, collapse = '</code>, <code>')}</code></span>"
  )
} else {
  glue(
    "<span style='color: orange;'>&#x26A0; Missing expected variables: ",
    "<code>{paste(vars_missing, collapse = '</code>, <code>')}</code></span>"
  )
}

# note about renamed and created variables
fixes_applied <- character(0)
if (length(legacy_found) > 0) {
  fixes_applied <- c(
    fixes_applied,
    glue(
      "Renamed: <code>{paste(legacy_found, collapse = '</code>, <code>')}</code>",
      " &rarr; <code>isei08_*</code>"
    )
  )
}
if (length(composites_created) > 0) {
  fixes_applied <- c(
    fixes_applied,
    glue(
      "Created: <code>{paste(composites_created, collapse = '</code>, <code>')}</code>"
    )
  )
}
if (length(composites_updated) > 0) {
  fixes_applied <- c(
    fixes_applied,
    glue(
      "Recomputed: <code>{paste(composites_updated, collapse = '</code>, <code>')}</code>"
    )
  )
}
fixes_note <- if (length(fixes_applied) > 0) {
  glue("<br><strong>Fixes applied:</strong> {paste(fixes_applied, collapse = '; ')}")
} else {
  ""
}

htmltools::HTML(glue(
  "<div class='callout-note'>
  <strong>Step 5 Data Verification</strong><br><br>
  <strong>Observations:</strong> {n_obs}<br>
  <strong>Partnered respondents:</strong> {n_partnered}<br>
  <strong>Non-partnered respondents:</strong> {n_non_partnered}<br><br>
  <strong>Variable status:</strong> {verification_status}<br>
  <strong>Expected variables present:</strong> {length(vars_present)} / {length(all_expected)}{fixes_note}
  </div>"
))
```

## Verify Parental Composite Integrity

```{r}
#| label: verify-composites
#| results: asis
#| code-fold: show
#| column: body-outset-right

# helper function to compute parental mean with consistent na.rm = FALSE logic
compute_parental_mean <- function(data, var1, var2) {
  result <- rowMeans(data[, c(var1, var2)], na.rm = FALSE)
  both_missing <- is.na(data[[var1]]) & is.na(data[[var2]])
  result[both_missing] <- NA
  result
}

# verify education_parental computation
edu_components_present <- all(
  c("education_father", "education_mother") %in% names(imputed_data)
)

if ("education_parental" %in% names(imputed_data) && edu_components_present) {
  edu_expected <- compute_parental_mean(
    imputed_data, "education_father", "education_mother"
  )
  edu_match <- all.equal(
    imputed_data$education_parental,
    edu_expected,
    tolerance = 1e-10,
    check.attributes = FALSE
  )
  edu_status <- if (isTRUE(edu_match)) {
    if ("education_parental" %in% composites_created) {
      "<span style='color: blue;'>&#x2714; Created from components</span>"
    } else if ("education_parental" %in% composites_updated) {
      "<span style='color: blue;'>&#x2714; Recomputed from components</span>"
    } else {
      "<span style='color: green;'>&#x2714; Verified</span>"
    }
  } else {
    "<span style='color: red;'>&#x2718; Mismatch detected</span>"
  }
} else if (!"education_parental" %in% names(imputed_data)) {
  edu_status <- "<span style='color: red;'>&#x2718; Variable missing</span>"
} else {
  edu_status <- "<span style='color: orange;'>&#x26A0; Component variables missing</span>"
}

# verify isei08_parental computation
occ_components_present <- all(
  c("isei08_father", "isei08_mother") %in% names(imputed_data)
)

if ("isei08_parental" %in% names(imputed_data) && occ_components_present) {
  occ_expected <- compute_parental_mean(
    imputed_data, "isei08_father", "isei08_mother"
  )
  occ_match <- all.equal(
    imputed_data$isei08_parental,
    occ_expected,
    tolerance = 1e-10,
    check.attributes = FALSE
  )
  occ_status <- if (isTRUE(occ_match)) {
    if ("isei08_parental" %in% composites_created) {
      "<span style='color: blue;'>&#x2714; Created from components</span>"
    } else if ("isei08_parental" %in% composites_updated) {
      "<span style='color: blue;'>&#x2714; Recomputed from components</span>"
    } else {
      "<span style='color: green;'>&#x2714; Verified</span>"
    }
  } else {
    "<span style='color: red;'>&#x2718; Mismatch detected</span>"
  }
} else if (!"isei08_parental" %in% names(imputed_data)) {
  occ_status <- "<span style='color: red;'>&#x2718; Variable missing</span>"
} else {
  occ_status <- "<span style='color: orange;'>&#x26A0; Component variables missing</span>"
}

htmltools::HTML(glue(
  "<div class='callout-note'>
  <strong>Composite Integrity Verification</strong><br><br>
  <code>education_parental</code> = mean(education_father, education_mother): {edu_status}<br>
  <code>isei08_parental</code> = mean(isei08_father, isei08_mother): {occ_status}
  </div>"
))
```

## SES Variable Structure

```{r}
#| label: tbl-ses-structure
#| tbl-cap: "SES Variable Structure"
#| code-fold: show
#| column: body-outset-right

col_names <- c("Variable", "Present", "$n_{valid}$", "$\\%_{valid}$")
footnote_text <- glue("$N = {n_obs}$")

ses_structure <- data.frame(
  Variable = c(
    "education_self", "education_partner", "education_father",
    "education_mother", "education_parental",
    "isei08_self", "isei08_partner", "isei08_father",
    "isei08_mother", "isei08_parental"
  ),
  stringsAsFactors = FALSE
) %>%
  dplyr::mutate(
    Present = ifelse(Variable %in% names(imputed_data),
      "$\\checkmark$", "$\\times$"
    ),
    n_valid = sapply(Variable, function(v) {
      if (v %in% names(imputed_data)) {
        sum(!is.na(imputed_data[[v]]))
      } else {
        NA_integer_
      }
    }),
    pct_valid = sapply(Variable, function(v) {
      if (v %in% names(imputed_data)) {
        round(100 * mean(!is.na(imputed_data[[v]])), 1)
      } else {
        NA_real_
      }
    })
  )

ses_structure$Present %<>% md_to_html()

ses_structure %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

## Summarize Missing Data Post-Imputation

After imputation, only structural missingness should remain (partner variables for non-partnered respondents).

```{r}
#| label: missing-post-imputation-attrs
#| code-fold: true
#| column: body-outset-right

col_names <- c("Variable", "$n_{miss}$", "$\\%_{miss}$", "Structural$^{1}$")

footnote_text <- glue(
  "$N = {n_obs}$; ",
  "$^{{1}}$ Structural missingness = partner-specific variables for non-partnered respondents."
)
```

```{r}
#| label: tbl-missing-post-imputation
#| tbl-cap: "Missing Values After Imputation"
#| code-fold: show
#| column: body-outset-right

# summarize missingness for key analysis variables
missing_summary <- imputed_data %>%
  dplyr::select(dplyr::any_of(all_expected)) %>%
  summarize_na() %>%
  dplyr::mutate(
    variable = beautify_names(variable),
    structural = ifelse(
      grepl("partner", variable, ignore.case = TRUE) & n_miss > 0,
      "Yes", "No"
    )
  )

missing_summary %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

## Descriptive Statistics

```{r}
#| label: descriptive-stats-col-names
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "$\\ $", "$n$", "$\\bar{x}$", "$\\text{Min.}$", "$\\text{Max.}$",
  "$SD$", "$\\text{Skewness}$", "$\\text{Kurtosis}$", "$SE$",
  "$\\text{Normality}^{\\dagger}$"
)

footnote_text <- glue(
  "Shapiro-Wilk Test$^{{\\dagger}}$<br>",
  "'&#x2014;' indicates variables with $\\leq 4$ unique values ",
  "(normality test not applicable)"
)
```

```{r}
#| label: tbl-descriptive-stats-all
#| tbl-cap: "Descriptive Statistics for Analysis-Ready Dataset"
#| code-fold: show
#| column: page-right

imputed_data %>%
  dplyr::select(-idnummer) %>%
  setNames(beautify_names(names(.))) %>%
  describe(
    normality = TRUE,
    print_only = FALSE
  ) %>%
  dplyr::rowwise() %>%
  dplyr::ungroup() %>%
  dplyr::select(-p) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

## Stratified Descriptive Statistics

### Partnered Respondents

```{r}
#| label: tbl-descriptive-stats-partnered
#| tbl-cap: "Descriptive Statistics: Partnered Respondents"
#| code-fold: show
#| column: page-right

imputed_data %>%
  dplyr::filter(partner == 1) %>%
  dplyr::select(-idnummer, -partner) %>%
  setNames(beautify_names(names(.))) %>%
  describe(
    normality = TRUE,
    print_only = FALSE
  ) %>%
  dplyr::rowwise() %>%
  dplyr::ungroup() %>%
  dplyr::select(-p) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

### Non-Partnered Respondents

```{r}
#| label: tbl-descriptive-stats-non-partnered
#| tbl-cap: "Descriptive Statistics: Non-Partnered Respondents"
#| code-fold: show
#| column: page-right

# prepare non-partnered data, removing partner-specific variables
non_partnered_data <- imputed_data %>%
  dplyr::filter(partner == 0) %>%
  dplyr::select(
    -idnummer, -partner,
    -dplyr::any_of(c("education_partner", "isei08_partner"))
  )

# remove columns with all NA values (if any remain)
non_partnered_data <- non_partnered_data %>%
  dplyr::select(where(~ !all(is.na(.x))))

non_partnered_data %>%
  setNames(beautify_names(names(.))) %>%
  describe(
    normality = TRUE,
    print_only = FALSE
  ) %>%
  dplyr::rowwise() %>%
  dplyr::ungroup() %>%
  dplyr::select(-p) %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

## Export Analysis-Ready Dataset

```{r}
#| label: export-final-dataset
#| code-fold: show
#| column: body-outset-right

# save finalized dataset for downstream analysis
haven::write_sav(
  imputed_data,
  here::here("assets", "data", "saved", "analysis_ready.sav")
)
```

## Variable Summary

```{r}
#| label: variable-summary-attrs
#| code-fold: true
#| column: body-outset-right

col_names <- c(
  "Variable", "Type", "$n_{valid}$", "$n_{miss}$",
  "$\\%_{miss}$", "$n_{unique}$", "Category"
)

footnote_text <- glue("$N = {n_obs}$")
```

```{r}
#| label: tbl-variable-summary
#| tbl-cap: "Variable Summary for Analysis-Ready Dataset"
#| code-fold: show
#| column: page-right

# create comprehensive variable summary
var_summary <- data.frame(
  variable = names(imputed_data),
  stringsAsFactors = FALSE
) %>%
  dplyr::mutate(
    type = sapply(imputed_data, function(x) {
      if (is.numeric(x)) "Numeric" else if (is.factor(x)) "Factor" else class(x)[1]
    }),
    n_valid = sapply(imputed_data, function(x) sum(!is.na(x))),
    n_miss = sapply(imputed_data, function(x) sum(is.na(x))),
    pct_miss = round(100 * n_miss / n_obs, 1),
    n_unique = sapply(imputed_data, function(x) length(unique(x[!is.na(x)]))),
    category = dplyr::case_when(
      variable == "idnummer" ~ "Identifier",
      grepl("^education", variable) ~ "Education",
      grepl("^isei08", variable) ~ "Occupation (ISEI-08)",
      grepl("^isco08r", variable) ~ "Occupation (ISCO-08)",
      grepl("^brc14", variable) ~ "Occupation (BRC-14)",
      grepl("^dhd", variable) ~ "Dietary Outcome",
      grepl("income", variable) ~ "Income",
      grepl("partner", variable) ~ "Partner Status",
      TRUE ~ "Other"
    )
  ) %>%
  dplyr::arrange(category, variable) %>%
  dplyr::mutate(variable = beautify_names(variable))

var_summary %>%
  as_kable() %>%
  footnote() %>%
  add_style()
```

::: {style="margin-block-end: 2rem;"}
:::

## Output

| File                                     | Content                                             |
|------------------------------------------|-----------------------------------------------------|
| `assets/data/saved/analysis_ready.sav`   | Analysis-ready dataset with verified SES composites |

**Next:** [Step 7 (Unadjusted Models)](/analysis/models/unadjusted/index.qmd)
